---
title: "cryo_shared_ASVs"
author: "Holly Nichols"
date: "2024-02-01"
output: html_document
---

#Goals

**Motivating questions: are the ASVs in the donor pools making it to adults? What's the fidelity of this transplant at the "strain" level **

* Figure out which ASVs in water, larvae, and adults are shared with donor pools 
* for the adult ASVs - how many can be linked back to paired donor pool samples? 
* which ASVs are most often getting from donor pool all the way to adults, and does this vary by uni_tx? 
    



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(phyloseq)
library(tidyverse)
#install.packages("janitor")
library("janitor") #for cleaning up some row names 
library("cowplot") #for plotting




ps.filled = readRDS("../data/phyloseq/crPSFilled")


ps.homogenate_UW <- subset_samples(ps.filled, sample_type == "homogenate" & uni == "UW") %>% prune_taxa(taxa_sums(.)>0,.)


ps.homogenate_LSTM <- subset_samples(ps.filled, sample_type == "homogenate" & uni == "LSTM")%>% prune_taxa(taxa_sums(.)>0,.)




```

#Elizabethkingia ASVs 

Additional File 6 and 11

```{r elizabethkingia}

#make a phyloseq object with JUST elizabethkingia ASVs


#let's get relabund psobj first 
ps.relabund <- transform_sample_counts(ps.filled, function(x) { x/sum(x) })



ps.elizabethkingia <- subset_taxa(ps.filled, Genus == "Elizabethkingia")
ps.elizabethkingia.relabund <- subset_taxa(ps.relabund, Genus == "Elizabethkingia")
#60 Elizabethkingia ASVs, wow 


#plot the relative abundance by whether eaach ASV is Elizabethkignia or not  

taxa_abundance_table_ez<- psmelt(ps.elizabethkingia.relabund)
#split homogenate into UW-h and LSTM-h , just mutate a new column 
ez_hsplit <- taxa_abundance_table_ez %>%
  mutate(
    sample_type_hsplit = case_when(
      sample_type == "homogenate" & uni == "UW" ~ "homogenate_UW",
      sample_type == "homogenate" & uni == "LSTM" ~ "homogenate_LSTM",
      TRUE ~ sample_type # Keep other values unchanged
    )
  )

no_h <- filter(ez_hsplit, sample_type != "homogenate")

only_h <- filter(ez_hsplit, sample_type == "homogenate")


StackedBarPlot <- no_h %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = OTU)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance",title = "Elizabethkingia Relative Abundance")+ 
   facet_wrap(~interaction(uni_tx,sample_type_hsplit), scales = "free_x") +
 theme_cowplot() + 
      theme(legend.position="none") +
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
      scale_y_continuous(expand = c(0,0.001))+# move x-axis up so it's nearly touching bars but not obscuring 
    coord_cartesian(ylim=c(0,1))
StackedBarPlot
# Very clearly shows that all of the LSTM samples have tons of Elizabethkingia while only the UW-mismatch have elizabethkingia - also that there's one main ASV that dominates even though there are 60 ASVs of eliz in the dataset 

#ggsave("./plots/Elizabethkingia_ASV_by_sample.pdf", height=5, width=9, device="pdf") 

#just homogenate
StackedBarPlot_h <- only_h %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = OTU)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance",title = "Elizabethkingia Relative Abundance")+ 
   facet_wrap(~interaction(uni_tx,sample_type_hsplit), scales = "free_x") +
 theme_cowplot() + 
      theme(legend.position="none") +
  theme(text = element_text(size=20), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8)) +
    scale_y_continuous(expand = c(0,0.001))+ # move x-axis up so it's nearly touching bars but not obscuring 
  coord_cartesian(ylim=c(0,1))

StackedBarPlot_h
#ggsave("./plots/Elizabethkingia_ASV_donor_pools.pdf", height=2.5, width=4, device="pdf") 



head(taxa_abundance_table_ez)

temp <- taxa_abundance_table_ez %>% group_by(OTU) %>% summarise(mean_perc = 100* mean(Abundance),
                                                                max_perc = 100*max(Abundance))

#ok, this tells me that most of the Elizabethkingia ASVs are rare, and there are 7 that get >1% abundance, but it's one that really dominates




```












# Asaia - where did she go? 
Asaia is prevalent in LSTM homogenate samples, but doesn't seem to show up anywhere else. 
A tiny bit is visible in some samples, like less than 1% relabund, but clearly is not a strong contender for surviving in a cryopreserved community 
Also i am reassured that it is present anywhere - not a decontam glitch, just biology!

```{r asaia}
melt_relabund <- psmelt(ps.relabund)

head(melt_relabund)

dfAsaia <- filter(melt_relabund, Genus == "Asaia")
unique(dfAsaia$OTU)
#four different asaia ASVs. 



StackedBarPlot <- dfAsaia %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity", fill = "purple") +
  labs(x = "",y = "Relative Abundance",title = "Asaia Relative Abundance")+ 
   facet_wrap(~interaction(sample_type,uni_tx), scales = "free_x", nrow = 2) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))
StackedBarPlot


#mostly one asaia ASV, doesnt's show up in downstream samples to any significant extent. 

```


#Are the Actinos falling off in later time points? 

Faceted this by both interaction(uni_tx,sample_type) and timepoint. 

* First plot shows that there's more Actinos in larvae - though also some in LSTM-mismatch adults. 
* Second plot shows that Actinos do not actually fall off in later time points, but this maybe driven by a particularly strong showing in LSTM samples at the 6 month timepoint. 


```{r actino}

dfActino <- filter(melt_relabund, Phylum == "Actinobacteriota")
length(unique(dfActino$OTU))
#452 different Actino ASVs. 


#Actinos in different sample types? 
StackedBarPlot <- dfActino %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity", show.legend = FALSE) +
  labs(x = "",y = "Relative Abundance",title = "Actinobacteriota Relative Abundance")+ 
   facet_wrap(~interaction(sample_type,uni_tx), scales = "free_x", nrow = 2) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))
StackedBarPlot


#actinos at different timepoints? 
StackedBarPlot2 <- dfActino %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity", show.legend = FALSE) +
  labs(x = "",y = "Relative Abundance",title = "Actinobacteriota Relative Abundance")+ 
   facet_wrap(~interaction(timepoint), scales = "free_x", nrow = 2) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))
StackedBarPlot2
```



## Frequently transmitted ASVs ASVs

**Goal:** What proportion of ASVs in donor pool samples go on to be found in the water/larvae/adults that they went into? 

Pseudocode: 

Make a dataframe with just the UW-match ASVs and counts per sample (OTU table)


Make new phyloseq object with just UW homogenate ASVs in it
Loop through all of the ASVs found in UW homogenate
  for each ASV
    check if nonzero in adults 
    if so, put a TRUE in a newcol
    
Want a dataframe that's like: 


ASV                         homogenate1    ...   homogenate 40    in_adults     in_larvae       in_water
q4ou25gtlaedfkjghqvob7        0.24         ...        0.35            TRUE         FALSE          TRUE
  
This could help build that single stacked bar, and we could have one bar for each condition to see whether the swap affects the **(1)** number of ASVs that get transmitted and **(2)** whether the relative abundance of ASVs that are shared with transplants is higher 
* From LSTM to LSTM 
* From LSTM to UW 
* From UW to UW
* From UW to LSTM 



  

```{r byflask_transformations}

#first goal: for those ASVs in homogenate, which ASVs made it to the most transplants (larvae/water/adults)?
#second goal: how does that change between the matched and mismatched tx? 


#so very first - how many LSTM homogenate ASVs made it into the UW-mismatch adults? 

#start by working with the NON-relabund phyloseq objects
ps.homogenate_UW <- subset_samples(ps.filled, sample_type == "homogenate" & uni == "UW") %>% prune_taxa(taxa_sums(.)>0,.)
#99 taxa


ps.homogenate_LSTM <- subset_samples(ps.filled, sample_type == "homogenate" & uni == "LSTM") %>% prune_taxa(taxa_sums(.)>0,.)
#729 taxa


# melt phyloseq data to data frame

mdf <- psmelt(ps.filled) #this takes a minute or two  to run 



# compute taxa sum according to sample type (how many adult vs larvae vs water samples have a given ASV? )
sampletype_abund <- dplyr::group_by(mdf, OTU, sample_type) %>% 
  dplyr::summarise(abundance = sum(Abundance))




sampletype_abund_spread <- tidyr::spread(sampletype_abund, sample_type, abundance)



sampletype_unitx_abund <- dplyr::group_by(mdf, OTU, interaction(uni_tx,sample_type)) %>% 
  dplyr::summarise(abundance = sum(Abundance))

# Rename column to something that doesn't look like a function
names(sampletype_unitx_abund)[names(sampletype_unitx_abund) == "interaction(uni_tx, sample_type)"] <- "ix"


tidyr::spread(sampletype_unitx_abund, ix, abundance)
#does not contain homogenate samples 
#rows = ASVs
#columns = sample_type by uni_tx
#entry  = number of reads found in samples that match that description
#also need to count how many samples fit that description 
#AND most importantly - how many samples just have detectable amounts of that ASV? 

#if I then filter this dataframe to only have ASVs that are found in LSTM_homogenate, that tells me which LSTM homogenates 
  

####ok, need a dataframe that has OTUs and then isntad of summing Abundance, just count the number of samples with nonzero values for Abundance. 


#old version that also counts abundance and total samples
# ASVs_in_sampletype <- dplyr::group_by(mdf, OTU, sample_type) %>% 
#   dplyr::summarise(abundance = sum(Abundance), total_sams = n(), nonzero_sams = sum(Abundance != 0) ) 


#for full df - 
#389 adult samples 
#78 homogenate samples 
#56 larvae
#72 water 





grandASVdf <- dplyr::group_by(mdf, OTU, interaction(uni_tx,sample_type)) %>% 
  dplyr::summarise(nonzero_sams = sum(Abundance != 0)) #this is the operative bit of code that sums up how many samples have nonzero read counts for each ASV 



# Rename column to something that doesn't look like a function, and replace hyphens with underscores
names(grandASVdf)[names(grandASVdf) == "interaction(uni_tx, sample_type)"] <- "ix"
grandASVdf$ix <- sub("-", "_", grandASVdf$ix)
grandASVdf_wide <- tidyr::spread(grandASVdf, ix, nonzero_sams) 



  
#make a list that's all of the ASV names in ps.homogenate_LSTM
LSTM_list <- c(rownames(otu_table(ps.homogenate_LSTM)))

#then only keep rows that are in a the LSTM homogenate 
inLSTM_df <- filter(grandASVdf_wide, OTU %in% LSTM_list)
#awesome, 729x14


#Create lists of sample identifiers that recieved each homogenate 
UW_recip <- c("LSTM_mismatch.adult","LSTM_mismatch.water", "LSTM_mismatch.larvae", "UW_match.adult", "UW_match.larvae", "UW_match.water")
LSTM_recip <- c("UW_mismatch.adult", "UW_mismatch.water", "UW_mismatch.larvae", "LSTM_match.adult", "LSTM_match.water", "LSTM_match.larvae")
  

  
LSTM_oft_transmitted <- select(inLSTM_df, one_of(LSTM_recip))


####top 20 ASVs transmitted from LSTM homogenate to UW adults BY NUMBER OF ADULTS THAT GOT IT  ######
LSTM_oft_transmitted <- LSTM_oft_transmitted[order(LSTM_oft_transmitted$UW_mismatch.adult,decreasing = TRUE) , ]
top20_LSTM_to_UW <- LSTM_oft_transmitted[1:20,1]
#interesting one to look through because this can tell you what made it to LSTM but not UW recipients - perhaps those are ASVs that are getting filtered out by the host? 
#for example: 938822a7d46b490d8a8d38409ef03f17 is NOT found in any UW mismatch samples, but it's in 66 LSTM match adults and 11mismatch waters (no larvae though)
#that asv is classified as: Family_Burkholderiaceae_A_592522


#####Top 20 ASVs transmitted from LSTM homogenate to LSTM adults BY NUMBER OF ADULTS THAT GOT IT########
LSTM_oft_transmitted <- LSTM_oft_transmitted[order(LSTM_oft_transmitted$LSTM_match.adult,decreasing = TRUE) , ]
top20_LSTM_to_LSTM <- LSTM_oft_transmitted[1:20,1]


```


#Oft-transmitted ASVs - WHich ASVs are present in more than half of adults that got LSTM homogneate?

```{r oft-trans_LSTM}

#count how many adults are in each condition

count_conditions <- dplyr::group_by(mdf, OTU, interaction(uni_tx,sample_type)) %>% 
  dplyr::summarise(nonzero_sams = sum(Abundance != 0), total_sams = n()) 
names(count_conditions)[names(count_conditions) == "interaction(uni_tx, sample_type)"] <- "ix"


count_conditions_small <- count_conditions %>% as.data.frame() %>%  select(ix, total_sams)
count_conditions_small <- count_conditions_small[1:13, ] 
count_conditions_small$ix <- sub("-", "_", count_conditions_small$ix)



#### All ASVs present in at least half of LSTM-match adults

inHalfLSTM_match <- filter(LSTM_oft_transmitted, LSTM_match.adult >= 62)
#wow, only 4 ASVs present in more than half of LSTM match adults that can be traced back to LSTM homogneate

# 7e18a39dd964eb207545655ebe71a459
# dcba105f35d8ebc9e22269c7491ad3a7	
# 945184b6386c192c0066e0a98a154780
# 938822a7d46b490d8a8d38409ef03f17


#in a third of adults
in33LSTM_match <- filter(LSTM_oft_transmitted, LSTM_match.adult >= 41)



#### All ASVs present in at least half of UW-mismatch adults
inHalfUW_mismatch <- filter(LSTM_oft_transmitted, UW_mismatch.adult >= 24)


in33UW_mismatch <- filter(LSTM_oft_transmitted, UW_mismatch.adult >= 16)


```


##Merge data by sorting_var

Current question - which ASVs made it to the  most "populations" or "instances of inoculation?" 
 
* Each flask was basically a microcosm, so to show that an ASV was found in a flask means that the transfer was successful even if it wasn't in every single mosquito in that flask 
* This only makes sense for adult samples though, so get rid of homogenate, water, and larvae before merge
* then let's merge the phyloseq object by sorting_var
* then we can do the same as above and count how many flasks share a particular ASV with the LSTM homogenate. 


At the very end of this chunk, it generates half of  AF13


```{r merge_flasks}


#CURRENT GOAL - count # of flasks with ASVs rather than number of samples 


#first KEEP ONLY ADULTS from mdf 

mdf_adults <- filter(mdf, sample_type == "adult")
mdf_adults$sorting_var <- sub("-", "_", mdf_adults$sorting_var) # clean up the sorting_var names 
mdf_adults$sorting_var <- sub(" ", "_", mdf_adults$sorting_var)
mdf_adults$sorting_var <- sub("-", "_", mdf_adults$sorting_var)




# flasks_df_allinfo <- group_by(mdf_adults, OTU, sorting_var) %>% summarise(abundance = sum(Abundance), adults_from_flask = n(), nonzero_adults = sum(Abundance != 0))

flasks_df <- group_by(mdf_adults, OTU, sorting_var) %>% summarise(abundance = sum(Abundance)) 
flasks_wide <- tidyr::spread(flasks_df, OTU, abundance)
temp <- as.data.frame(t(flasks_wide))
temp2 <- janitor::row_to_names(temp,row_number = 1)
temp3 <- tibble::rownames_to_column(temp2, "OTU")

#now only keep the ASVs that are also in LSTM homogenate

#LSTM_list - defined above as just a list of the 729 ASVs in LSTM homogneate 

inLSTM_byflask <- filter(temp3, OTU %in% LSTM_list)
#remember, at this point we only care whether each ASV was detected in that flask, not the actual read count. 



####################################
#################start the better sorting variable #########

#Define a function to extract characters from the right side of a string 
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

#substrRight(sample_data(ps.adult)$sorting_var, 1) # confirmed it produces a bunch of A/B/C/D



#goal - create a new column flaskID
  
ps.temp <- ps.filled
sample_data(ps.temp)$flaskID <- paste(sample_data(ps.temp)$uni_tx,sample_data(ps.temp)$timepoint, substrRight(sample_data(ps.temp)$sorting_var, 1), sep = "_")
sample_data(ps.temp)$flaskID  <- sub("-", "_", sample_data(ps.temp)$flaskID)
ps.flaskASVs <- subset_samples(ps.temp, flaskID != "__e")
ps.flaskASVs <- prune_taxa(taxa_sums(ps.flaskASVs) > 0, ps.flaskASVs) #trim taxa not found in this subset
#517 samples, 4107 taxa 

#ok, now make this long form, group by flaskID, sum the abudnace across teh whole flask group for each ASV
#also need to count how many samples are in each flask group, and how many total flask groups there are 
length((unique(sample_data(ps.flaskASVs)$flaskID)))
#72 flask groups (one of them has no adults though - that's #UW_d30_away_A or #UW_mismatch_d30_A )

#Now count how many samples are in each flask (This includes larvae, water, and adult samples). Might come in handy later
samples_by_flask <- table((sample_data(ps.flaskASVs)$flaskID), useNA = "ifany")
#max 13 samples, min 2, median 7, mean ~7.18


#ok NOW let's make this long form 
mdf_byflask <- psmelt(ps.flaskASVs) #this takes a minute to run, very big df 
dim(mdf_byflask)
#df [2,035,946 × 42]



abundance_byflask <- dplyr::group_by(mdf_byflask, OTU, flaskID) %>% 
  dplyr::summarise(abundance = sum(Abundance))
#will give total # of reads across that whole flask. We just need to know if it's zero or not
max(abundance_byflask$abundance) #confirmed that this is summing abundances 



abundance_byflask_spread <- tidyr::spread(abundance_byflask, flaskID, abundance)
#now rows are ASVs and columns are flasks, entries are summed reads across a whole flask 
#we could go and set any nonzero value to 1, then count how many flasks each ASV was detected in 



detected_byflask <- ungroup(abundance_byflask_spread)
detected_byflask[-1] <-  sapply(detected_byflask[-1], as.logical) #turned evertying into true/flase
detected_byflask[-1] <- sapply(detected_byflask[-1], as.numeric) #and converted to 1/0
detected_byflask_rn <- detected_byflask %>% remove_rownames %>% column_to_rownames(var="OTU") #OTU becomes a rowname

#if an ASV was detected in a flask, that entry is a 1. If not detected, entry is a 0. 


##########################LSTM homogneate ASVs found in LSTM-match 

LSTM_ASV_list <- c(rownames(otu_table(ps.homogenate_LSTM)))

inLSTM_byflask <- filter(detected_byflask, OTU %in% LSTM_ASV_list)
#217x73 - there are 729 ASVs in the LSTM homogneate, which means that like 500 ASVs are not detected in the flasks at all 
#we pruned taxa after removing the homogneate samples, so that's where that's coming from. 
#wonder what those ASVs are, and what % of the homogenate community they make up 
inLSTM_byflask_rn <- inLSTM_byflask %>% remove_rownames %>% column_to_rownames(var="OTU") #OTU becomes a rowname



#make a new df with only flasks starting with LSTM-match, UW-mismatch, or both (after filtering for just LSTM-homogenate ASVs)

ASVflasks_LSTM_match <- select(inLSTM_byflask_rn, starts_with("LSTM_match")) #24 flasks - 6 timepoints, 4 flasks
ASVflasks_UW_mismatch <- select(inLSTM_byflask_rn, starts_with("UW_mismatch")) #12 flasks
ASVflasks_LSTM_donor <- select(inLSTM_byflask_rn, starts_with("LSTM_match") | starts_with("UW_mismatch")) #36 flasks, includes all that got the LSTM homogneate


#change column to rownames

detected_byflask_rn <- detected_byflask %>% remove_rownames %>% column_to_rownames(var="OTU")
number_flasks_detected <- rowSums(detected_byflask_rn)
#ok, but let's just keep entries that are found in

LSTM_match_flasks_24 <- c(rowSums(ASVflasks_LSTM_match))
UW_mismatch_flasks_12 <- c(rowSums(ASVflasks_UW_mismatch))
combined_flasks_36 <- c(rowSums(ASVflasks_LSTM_donor))

LSTM_transmitted_byflask <- data.frame(LSTM_match_flasks_24,UW_mismatch_flasks_12,combined_flasks_36)
LSTM_transmitted_byflask <- LSTM_transmitted_byflask[order(LSTM_transmitted_byflask$combined_flasks_36,decreasing=TRUE),]

#general agreement between those ASVs that were transmitted to LSTM_match and UW-mismatch 

#keep only ASVs that were transmitted to at least 12 LSTM_match OR at least 6 UW-mismatch 

LSTM_oft_transmitted_byflask <- filter(LSTM_transmitted_byflask, LSTM_match_flasks_24 >= 12 | UW_mismatch_flasks_12 >= 6)
#narrowed to 39 ASVs that frequently made it to the next flask 
#cool to note which do not make it, see if we can add some tax info 
tax_info_genus <- tax_table(ps.filled) %>% as.matrix() %>% as.data.frame() %>% select(Genus)

LSTM_oft_transmitted_byflask_g <- left_join(rownames_to_column(LSTM_oft_transmitted_byflask), rownames_to_column(tax_info_genus), by = "rowname")
#write.csv(LSTM_oft_transmitted_byflask_g,"./ASVS_lost_LSTM.csv")



asvs_lost_to_UW <- filter(LSTM_oft_transmitted_byflask_g, UW_mismatch_flasks_12 == 0)
asvs_lost_to_UW$Genus



```


# Oft-transmitted from UW donor pool 

Find ASVs that made it to at least half of UW-match flasks, then see how many of those ASVs made it to LSTM-mismatch flasks (metric of "making it" to a flask = at least 1 adult detected with that ASV in the flask. )

The other half of AF13

```{r}


UW_list <- c(rownames(otu_table(ps.homogenate_UW)))


inUW_byflask <- filter(detected_byflask, OTU %in% UW_list)
#56x73 
inUW_byflask_rn <- inUW_byflask %>% remove_rownames %>% column_to_rownames(var="OTU") #OTU becomes a rowname


#make a new df with only flasks starting with LSTM-match, UW-mismatch, or both (after filtering for just LSTM-homogenate ASVs)

ASVflasks_UW_match <- select(inUW_byflask_rn, starts_with("UW_match")) #24 flasks - 6 timepoints, 4 flasks
ASVflasks_LSTM_mismatch <- select(inUW_byflask_rn, starts_with("LSTM_mismatch")) #12 flasks
ASVflasks_UW_donor <- select(inUW_byflask_rn, starts_with("UW_match") | starts_with("LSTM_mismatch")) #36 flasks, includes all that got the LSTM homogneate


#change column to rownames

detected_byflask_rn <- detected_byflask %>% remove_rownames %>% column_to_rownames(var="OTU")
number_flasks_detected <- rowSums(detected_byflask_rn)
#ok, but let's just keep entries that are found in

UW_match_flasks_24 <- c(rowSums(ASVflasks_UW_match)) #this is, per ASV, how many UW flasks it was in 
LSTM_mismatch_flasks_12 <- c(rowSums(ASVflasks_LSTM_mismatch))
combined_flasks_36 <- c(rowSums(ASVflasks_UW_donor))

UW_transmitted_byflask <- data.frame(UW_match_flasks_24,LSTM_mismatch_flasks_12,combined_flasks_36)
UW_transmitted_byflask <- UW_transmitted_byflask[order(UW_transmitted_byflask$combined_flasks_36,decreasing=TRUE),]



#keep only ASVs that were transmitted to at least 12 LSTM_match OR at least 6 UW-mismatch 

UW_oft_transmitted_byflask <- filter(UW_transmitted_byflask, UW_match_flasks_24 >= 12 | LSTM_mismatch_flasks_12 >= 6)
#narrowed to 27 ASVs that frequently made it to the next flask 
#cool to note which do not make it, see if we can add some tax info 
tax_info_genus <- tax_table(ps.filled) %>% as.matrix() %>% as.data.frame() %>% select(Genus)

UW_oft_transmitted_byflask_g <- left_join(rownames_to_column(UW_oft_transmitted_byflask), rownames_to_column(tax_info_genus), by = "rowname")
#write.csv(UW_oft_transmitted_byflask_g,"./ASVS_lost_UW.csv")



asvs_lost_to_LSTM <- filter(UW_oft_transmitted_byflask_g, LSTM_mismatch_flasks_12 == 0)
asvs_lost_to_LSTM$Genus

# Only Arthrobacter_E


```



# Graphing just the "lost taxa" from LSTM donor pool - how common are those 11 taxa in LSTM water, adults, donor pool? 

AF14

```{r}

ps.relabund <- transform_sample_counts(ps.filled, function(x) { x/sum(x) })
#melt_relabund <- psmelt(ps.relabund) #Commented out b/c it takes forever to run, and it's also generated upstream. 



#ok now need to filter for in asvs_lost_to_UW


asvs_lost_to_UW$rowname


df_UWlost <- filter(melt_relabund, OTU %in% asvs_lost_to_UW$rowname) %>% filter(sample_type == "adult")

facet_names <- c(
  "0h" = "Unfrozen", 
  "24h" = "24 hours", 
  "d10" = "10 days", 
  "d16" = "16 days", 
  "d30" = "30 days", 
  "mo6" = "6 months")

bars <- df_UWlost %>% filter(uni_tx == "LSTM-match") %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(timepoint), scales = "free_x", ncol = 6, labeller = as_labeller(facet_names)) +
  theme_cowplot() + 
  scale_fill_manual(values = c(
  "#1b9e77",  # teal-green
  "#d95f02",  # orange
  "#7570b3",  # purple
  "#e7298a",  # pink/magenta
  "#66a61e",  # olive-green
  "#e6ab02",  # mustard
  "#a6761d",  # brown
  "#666666",  # grey
  "#1f78b4",  # blue
  "#b2df8a",  # light green
  "#fb9a99"   # light salmon
)) +
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 11))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

bars

#ggsave("./plots/shared_ASVs/LSTM_lost_bars.pdf", height=3.5, width=15, device="pdf") 


```


# Graph UW donor pool ASVs that don't make it to LSTM-mismatch adults (just Arhrobacter)
Conclusion - they're really not common in UW-match either! 
AF15

```{r}

df_LSTMlost <- filter(melt_relabund, OTU %in% asvs_lost_to_LSTM$rowname) %>% filter(sample_type == "adult")

facet_names <- c(
  "0h" = "Unfrozen", 
  "24h" = "24 hours", 
  "d10" = "10 days", 
  "d16" = "16 days", 
  "d30" = "30 days", 
  "mo6" = "6 months")

bars2 <- df_LSTMlost %>% filter(uni_tx == "UW-match") %>% 
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
  facet_wrap(~timepoint, scales = "free_x", nrow = 1, labeller = as_labeller(facet_names)) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 11))+ 
  scale_y_continuous(expand = c(0,0.000001))+#+# move x-axis up so it's nearly touching bars but not obscuring 
  coord_cartesian(ylim=c(0,0.005))
bars2

#ggsave("./plots/shared_ASVs/UW_lost_bars.pdf", height=3.5, width=15, device="pdf") 
#super rare  in UW-match adults 


```






# ASVs in Adults that are shaed with donor pool samples 

On the basis of each individual adult mosquito, what prop of its ASVs can be traced back to the LSTM donor pool

Section to just stock_origin = LSTM 

For each sample - was that ASV in (a list of ASVs in LSTM homogenate) >> add (yes/no) to the tax table 


Final plot is faceted by uni_tx and time 

```{r}

ps.adult_got_LSTM <- subset_samples(ps.filled, sample_type == "adult" & stock_origin == "LSTM")
#172 samples
ps.adult_got_LSTM <- prune_taxa(taxa_sums(ps.adult_got_LSTM) > 0, ps.adult_got_LSTM) #trim taxa not found in this subset
#1533 taxa




CheckInSamples <- function(ps) {

  
  OTU <- vector(mode = "character", length = nrow(tax_table(ps)))
  Shared <- vector(mode = "character", length = nrow(tax_table(ps)))
  
  ASV_list <- rownames(tax_table(ps.homogenate_LSTM))
  
  for (i in 1:nrow(tax_table(ps))) {
  OTU[i] <- rownames(tax_table(ps))[i]
  if (rownames(tax_table(ps))[i] %in% ASV_list) {
    Shared[i] = "yes"}
  else {
    Shared[i] <- "no"}
  }
  
  Shared_df <- data.frame(Shared, row.names = OTU)    
  ps.temp <- ps
  temp_taxtable <- as.data.frame(tax_table(ps.temp)[,1:8]) #ensure that no other "Shared" column is on here 
  merged_taxtable <- merge(temp_taxtable,Shared_df, by = 'row.names')
  merged_taxtable <- merged_taxtable %>% remove_rownames %>% column_to_rownames(var="Row.names") #OTU becomes a rowname
 
  tax_table(ps.temp) <- as.matrix(merged_taxtable) #Add this dataframe back to the tax_table() from your ps object
  ps.norm = transform_sample_counts(ps.temp, function(x) x / sum(x) ) #set to relative abundaces

  melt_df <- psmelt(ps.norm)  #create long plottable df
  return(melt_df)
  
} 
  

df<- CheckInSamples(ps.adult_got_LSTM)





bar_plot<-df%>%
  ggplot(aes(x=full_id, y=Abundance, fill=Shared))+
  geom_col() +
   xlab("Individual Adult samples")+ ylab("Proportion of Reads")+
  theme_cowplot() + theme(text = element_text(size=20), axis.text.x=element_blank()) +
  scale_y_continuous(expand = c(0,0)) + 
  facet_wrap(~interaction(uni_tx,timepoint), scales = "free_x", nrow = 2) +
  theme(strip.text.x = element_text(size = 8))+
  ggtitle("Adult ASVs \n shared with LSTM homogenate")

bar_plot

#conclusion - nearly al ASVs that are found in adults can be traced ack to LSTM-homogenate, main exceptions actually being LSTM match 0h and LSTM match 24h -  perhpas this means they're getting colonized by microbes at low abudnance in stock that die off after short periods of freezing. 

```



#What about shared with UW donor pool? 


```{r adults_shared_with_UW}



ps.adult_got_UW <- subset_samples(ps.filled, sample_type == "adult" & stock_origin == "UW")
#217 samples
ps.adult_got_UW <- prune_taxa(taxa_sums(ps.adult_got_UW) > 0, ps.adult_got_UW) #trim taxa not found in this subset
#2277 taxa




CheckInSamples_UW <- function(ps) {

  
  OTU <- vector(mode = "character", length = nrow(tax_table(ps)))
  Shared <- vector(mode = "character", length = nrow(tax_table(ps)))
  
  ASV_list <- rownames(tax_table(ps.homogenate_UW))
  
  for (i in 1:nrow(tax_table(ps))) {
  OTU[i] <- rownames(tax_table(ps))[i]
  if (rownames(tax_table(ps))[i] %in% ASV_list) {
    Shared[i] = "yes"}
  else {
    Shared[i] <- "no"}
  }
  
  Shared_df <- data.frame(Shared, row.names = OTU)    
  ps.temp <- ps
  temp_taxtable <- as.data.frame(tax_table(ps.temp)[,1:8]) #ensure that no other "Shared" column is on here 
  merged_taxtable <- merge(temp_taxtable,Shared_df, by = 'row.names')
  merged_taxtable <- merged_taxtable %>% remove_rownames %>% column_to_rownames(var="Row.names") #OTU becomes a rowname
 
  tax_table(ps.temp) <- as.matrix(merged_taxtable) #Add this dataframe back to the tax_table() from your ps object
  ps.norm = transform_sample_counts(ps.temp, function(x) x / sum(x) ) #set to relative abundaces

  melt_df <- psmelt(ps.norm)  #create long plottable df
  return(melt_df)
  
} 
  

df<- CheckInSamples_UW(ps.adult_got_UW)



bar_plot<-df%>%
  ggplot(aes(x=full_id, y=Abundance, fill=Shared))+
  geom_col() +
   xlab("Individual Adult samples")+ ylab("Proportion of Reads")+
  theme_cowplot() + theme(text = element_text(size=20), axis.text.x=element_blank()) + scale_y_continuous(expand = c(0,0)) + 
  facet_wrap(~interaction(uni_tx,timepoint), scales = "free_x", nrow = 2) +
  theme(strip.text.x = element_text(size = 8))+
  ggtitle("Adult ASVs \n shared with UW homogenate")

bar_plot

#try faceting just by uni_tx

bar_plot<-df%>%
  ggplot(aes(x=full_id, y=Abundance, fill=Shared))+
  geom_col() +
   xlab("Individual Adult samples")+ ylab("Proportion of Reads")+
  theme_cowplot() + theme(text = element_text(size=20), axis.text.x=element_blank()) + scale_y_continuous(expand = c(0,0)) + 
  facet_wrap(~uni_tx, scales = "free_x", nrow = 2) +
  theme(strip.text.x = element_text(size = 8))+
  ggtitle("Adult ASVs \n shared with UW homogenate")

bar_plot




```



# Combo figure showing ASVs detected in adults that came from a given homogenate? 

(still at individual level, preview, not in final manuscript)

```{r}
#get info from adults that got LSTM homogenate

#Adults that got XYZ homogneate: (remake ps objects in case overwritten above )

ps.adult_got_UW <- subset_samples(ps.filled, sample_type == "adult" & stock_origin == "UW")
ps.adult_got_UW <- prune_taxa(taxa_sums(ps.adult_got_UW) > 0, ps.adult_got_UW) #trim taxa not found in this subset
ps.adult_got_LSTM <- subset_samples(ps.filled, sample_type == "adult" & stock_origin == "LSTM")
ps.adult_got_LSTM <- prune_taxa(taxa_sums(ps.adult_got_LSTM) > 0, ps.adult_got_LSTM)



df_UW<- CheckInSamples_UW(ps.adult_got_UW)
df_LSTM<- CheckInSamples(ps.adult_got_LSTM)





#combine them? 
df_combined <- dplyr::bind_rows(df_UW,df_LSTM)

head(df_combined)

#this code below does function, but it's just very slow. Use the simplified version below. 

# bar_plot<-df_combined%>%
#   ggplot(aes(x=full_id, y=Abundance, fill=Shared))+
#   geom_col() +
#    xlab("Individual Adult samples")+ ylab("Proportion of Reads")+
#   theme_cowplot() + theme(text = element_text(size=20), axis.text.x=element_blank()) + scale_y_continuous(expand = c(0,0)) + 
#   facet_wrap(~uni_tx, scales = "free_x", nrow = 2) +
#   theme(strip.text.x = element_text(size = 8),axis.ticks.x=element_blank())+
#   ggtitle("Adult ASVs \n shared with respective homogenate")
# 
# bar_plot


#ok - can't open in inskcape b/c too  many spearate datapoints. Can we get merge all of the "yes" values for each samples and all of the "no" samples?
#groupby




simplified <- df_combined %>% dplyr::group_by(Sample, Shared, full_id, uni_tx, uni, tx) %>% dplyr::summarise(Abundance = sum(Abundance))
#because "Shared" was included in the group_by(), it will automatically sum Abundance for "yes" vs "no"
head(simplified)


bar_plot<-simplified%>%
  ggplot(aes(x=full_id, y=Abundance, fill=Shared))+
  geom_col() +
   xlab("Individual Adult samples")+ ylab("Proportion of Reads")+
  theme_cowplot() + theme(text = element_text(size=20), axis.text.x=element_blank()) + scale_y_continuous(expand = c(0,0)) + 
  facet_wrap(~uni_tx, scales = "free_x", nrow = 2) +
  theme(strip.text.x = element_text(size = 8),axis.ticks.x=element_blank())+
    scale_fill_manual(values = c("#74C476", "#6BAED6"))+ 
  ggtitle("Adult ASVs \n shared with respective homogenate")

bar_plot

temp <- filter(simplified, Shared == "yes")
min(temp$Abundance) #min is 0.02?? only 2% of ASVs came from homogenate in one adult sample
mean(temp$Abundance) #93% shared on average
median(temp$Abundance) #median 98%
max(temp$Abundance)



  
#ggsave("./plots/shared_ASVs_adult.pdf", height=5, width=9, device="pdf") 


```



# Collapse this down into just four bars - one per uni_tx, and add error bars

Conclusion - LSTM-mismatch is the most variable (suppoorted by Levene's test) and shares the fewest ASVs (not statisticaly singificnatclaim) . 


Fig 3C

```{r}

#First we need to sum WITHIN EACH INDIVIDUAL all of the yes/no. Which exists already in df called simplified 

head(simplified)
#this groups by uni_tx and gets the  mean and sd within each group to build the error bars
#make sure we're looking at SEM instead of SD 

simplified_fourbars <- simplified %>% 
  dplyr::group_by(uni_tx, Shared) %>%
  dplyr::summarise(mean_abund = mean(Abundance), sem_abund = sd(Abundance)/sqrt(n())) %>%
            mutate(uni_tx = factor(uni_tx))

#now filter for just the error bars - i.e. we only need one entry per bar and the sd is the same for yes adn no anyway
error_bars <- filter(simplified_fourbars, Shared == "yes")

library(ggpubr)
ggplot(simplified_fourbars, aes(x = uni_tx, y = mean_abund)) +
  geom_bar(stat = 'identity', aes(fill = Shared)) +
  geom_errorbar(data = error_bars,
                aes(x = uni_tx, ymax = mean_abund + sem_abund, ymin = mean_abund - sem_abund), 
                width = 0.2) + 
  theme_pubr()+
    theme(axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5))+
  scale_fill_manual(values = c("#74C476", "#6BAED6"))+ 
  scale_y_continuous(expand = c(0,0)) 

#ggsave("./plots/shared_ASVs_adult_fourbars.pdf", height=5, width=5, device="pdf") 



```




#Stats on fourbars  - adult relabund shared with paired donor pool

```{r}


no_by_samples <- filter(simplified, Shared == "no")
yes_by_samples <- filter(simplified, Shared == "yes")



ggqqplot(no_by_samples$Abundance)
ggqqplot(yes_by_samples$Abundance)
#skewed - do KW instead of anova

kruskal.test(Abundance ~ uni_tx, data = no_by_samples) 
# Kruskal-Wallis chi-squared = 49.09, df = 3, p-value = 1.248e-10


library(dunn.test)
dunn.test(no_by_samples$Abundance,no_by_samples$uni_tx,method = "bonferroni")
#LSTM match differenct from all 
#UW match different from UW-mismatch

#LSTM mismatch only different from LSTM-match 

#gotUW are teh same as each other 
#LSTM-mismatch is only different from LSMT-match- doesn't make sense b/c UW-mismatch def looks different - probably because of differences in sample number. 

# data: x and group
# Kruskal-Wallis chi-squared = 49.0898, df = 3, p-value = 0
# 
# 
#                            Comparison of x by group                            
#                                  (Bonferroni)                                  
# Col Mean-|
# Row Mean |   LSTM-mat   LSTM-mis   UW-match
# ---------+---------------------------------
# LSTM-mis |  -5.584433
#          |    0.0000*
#          |
# UW-match |  -5.771356   0.636288
#          |    0.0000*     1.0000
#          |
# UW-misma |  -4.756477  -0.025516  -0.563582
#          |    0.0000*     1.0000     1.0000



#different answer if we do against teh abundance of shared ASVs?? 

kruskal.test(Abundance ~ uni_tx, data = yes_by_samples) # p =2.6e-14

dunn.test(yes_by_samples$Abundance,yes_by_samples$uni_tx,method = "bonferroni")
#as expected, the exact same answer. That does increase my confidence in the method 



#levene's test  testing whether they have different variances, seems like LSTM-mismatch may be more variable 

library(car)
leveneTest(Abundance~ uni_tx, data = no_by_samples) 
#        Df F value    Pr(>F)    
# group   3  13.955 1.179e-08 ***
#       385                     


#which ones are different? Apparently levene is just ANOVA on residuals (sample variance), so let's calculate residuals and then do anova/tukey


#https://stackoverflow.com/questions/43646987/multiple-comparison-post-hoc-test-for-levenes-test

#need a new column that's a median 
no_by_samples$uni_tx <- as.factor(no_by_samples$uni_tx)

#n is working, but median is not. 
temp <- simplified %>% 
  group_by(uni_tx, Abundance) %>% 
  mutate(count = n(), med = median(Abundance), mean = mean(Abundance)) %>% 
  ungroup()

#fuck it. I can figure out the median of each manually, and add it manually 

LSTM_match_med <- no_by_samples %>% filter(uni_tx == "LSTM-match")
LSTM_mismatch_med <- no_by_samples %>% filter(uni_tx == "LSTM-mismatch")
LSTM_match_med <- no_by_samples %>% filter(uni_tx == "LSTM-match")
LSTM_match_med <- no_by_samples %>% filter(uni_tx == "LSTM-match")
median(LSTM_mismatch_med$Abundance)


#summarize some stuff
find_medians <- no_by_samples %>% 
  dplyr::group_by(uni_tx, Shared) %>%
  dplyr::mutate(mean_abund = mean(Abundance), sd_abund = sd(Abundance), median_abund = median(Abundance)) %>%
            mutate(uni_tx = factor(uni_tx)) %>% ungroup()
#then use median to calculate residuals
find_residuals <- find_medians %>% mutate(resids = abs(median_abund - Abundance))



#ok. So here's the big difference. My plot is showing MEAN, which is different from MEDIAN. There's generally good agreement between mean and median, but not in LSTM-mismatch BECAUSE there's so much variation - so that's drivign the mean but not the median. Regardless, let's push through and get this post hoc for levene and move on. 




# Then we run an ANOVA, and post-hoc if necessary:
diy_levene<-aov(resids~uni_tx,find_residuals)
summary(diy_levene) #significant
TukeyHSD(diy_levene)

#significant = LSTM-mismatch vs LSTM-mismatch 
#UW-match vs LSTM-mismatch 
#UW-mismatch vs LSTM mismatch

# Levene's test shows that variance is highest in the LSTM-mismatch treatment, and not significant in other treatments 



```



