---
title: "cryo_wing_length"
author: "Holly Nichols"
date: "2023-01-12"
output: html_document
---

```{r setup, include=FALSE}

library(phyloseq)
library(tidyverse) 
library(ggplot2)
library(MetBrewer) #for colors on ggplot
library(forcats) #this might help with ordering x-axes for graphs
library(ggpubr)
library(lattice)#for histograms

```

## Wing Length

This is a quick look at the wing lengths for UW and LSTM data. We have a few broad questions: 

* How do our variables (timepoint, uni, treatment, origin of stock, and the interactions between them) affect wing length?
* Which statistical tests are most appropriate to answer these questions? 
* How can we visualize these data to tell a true and compelling narrative?



To-do list: 

* Finish all the stat comparisons desired and add to the google sheet
* Move the stat comparisons to be next to their corresponding graph?? Maybe even add significance to stats here? 


```{r import, include = FALSE}


wing_length <-read_csv("../data/wing_length.csv")
head(wing_length)
tail(wing_length)
wing_length <- as.data.frame(wing_length)
wing_length[sapply(wing_length, is.character)] <- lapply(wing_length[sapply(wing_length, is.character)], as.factor)

```


# Stats
Order of operations: 

1. Check assumptions for parametric stats (look at histograms, residuals)
2. If assumptions met for parametric stats, run linear model lm()
3. If assumptions not met, use kruskal-wallis and Dunn's

How to choose which chunks of data use? 

* Think of which comparisons are biologically relevant. Does it make sense to compare LSTM mismatch 6month to UW match d10? No. So don't run a test with all the unique identifiers, you'll lose too much power. 
* OR - do run a test with all of the things thrown in there and just pick out the comparisons of interest, but know you're only getting the strongest trends due to loss of power. 

In my case, here's what I plan to test: 

* Globally, does timepoint affect wing length? 
* Globally, does Uni affect wing length?
* Globally, does treatment (match/mismatch)affect wing length? How does this interact with uni? 



```{r stats}

#test assumptions for parametric stats

library(car)

#run the whole chunk below together 
par(mfrow = c(1,2)) #makes a 2-paned plots
qqnorm( residuals(lm(wing_length ~ timepoint, data= wing_length)),pch=1,frame=FALSE) #build residiual plots
qqline( residuals(lm(wing_length ~ timepoint, data= wing_length)),lwd=1) #put a line on that residual plot. MUST BE RUN WITH THE LINE ABOVE
plot(fitted(lm(wing_length ~ timepoint, data= wing_length)),
  residuals(lm(wing_length ~ timepoint, data= wing_length)),xlab="fitted", 
  ylab= "raw residual",main= "test", abline(h=0, lty=2)) #Approved. No weird scattering. The data are weird on the x-axis b/c there's only one fitted value for each category - ie. there's only one mean for 0h, or 24h, and all the residuals are showing spread from there. 
plot(histogram(wing_length$wing_length)) #this is a histogram, looks normal
car::boxCox(lm(wing_length ~ timepoint, data = wing_length)) 
#looks like the max here is around 0.25, transform not required 


#conclusion - proceed with linear model 


wing_length_trim <- select(wing_length, -pic_batch)
wing_length_trim <- select(wing_length_trim, -collection_date)
class(wing_length_trim$wing_length)

#now have an object with any n/a values removed. 

head(wing_length)
class(wing_length$timepoint)


#with all fixed effects
lmTime <- lm(wing_length ~ uni*treatment + timepoint, data = wing_length)
summary(lmTime)

#First, UW wings are smaller than LSTM wings. (uni)
#Next, treatment_mismatch on its own is not significant, marginal p of 0.073
#Timepoints d10, d30, and mo6 have bigger wings than the unfrozen timepoint 
#and finally, university and mismatch treatment significantly interact!! 
#so mismatch have smaller wings in UW but not in LSTM 

#Can we get a p value for the comparison between uw-match and uw-mismtach? yes>>>  
#eemeans for contrasts 
library(emmeans)

emm <- emmeans(lmTime, ~ treatment | uni)
pairs(emm)

```


# Graph it all! 

Troubleshooting the graph order and colors on the way to plotting every single full_id on its own. 


```{r plotAll}
#reorder
#wing_length$treatment <- factor(wing_length$treatment, c("fresh", "24h", "d10", "d16 away", "d16 home", "d30 away", "d30 home", "mo6 away", "mo6 home"), ordered = TRUE)
#obsolete, made a new variable called timepoint_order
#may get weird with the different away/home tx and the UW vs LSTM 


#wing_length$timepoint <- factor(wing_length$timepoint, c("fresh", "24h", "d10", "d16 away", "d16 home", "d30 away", "d30 home", "mo6 away", "mo6 home"), ordered = TRUE)

levels(wing_length$full_id)


wing_length$full_id <- factor(wing_length$full_id, c("UW-fresh", "UW-24h", "UW-d10", "UW-d16 away", "UW-d16 home", "UW-d30 away", "UW-d30 home", "UW-mo6 away", "UW-mo6 home", "LSTM-fresh", "LSTM-24h", "LSTM-d10", "LSTM-d16 away", "LSTM-d16 home", "LSTM-d30 away", "LSTM-d30 home", "LSTM-mo6 away", "LSTM-mo6 home"), ordered = TRUE)
#ok - so i manually changed the order of factors. Turns out it still has the same problem of colorign the "fresh" samples between the d30 and mo6 samples. 
#so i could've just used the first line that has fct_reorder. Let's stick with that to keep it simple. 

#p <- ggplot(wing_length, aes(x = fct_reorder(full_id, graph_order), y = wing_length))+
  #this also reorders correctly, but the color order is wack and the x axis labels are off by one. 



p <- ggplot(wing_length, aes(x = full_id, y = wing_length))+
  geom_boxplot()+
  aes(fill=timepoint)+
  scale_fill_manual(values = met.brewer("Tam", 6))+
  theme_pubr() +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5))+
  geom_jitter(width = 0.2)+
  ggtitle("Wing length for all cryopresrvation treatments")

  
p



```






# Wing length by University (UW vs LSTM)



```{r byUni}
p <- ggplot(wing_length, aes(x = uni, y = wing_length))+
  geom_boxplot()+
  aes(fill=uni)+
  scale_fill_manual(values = c("#aadce0ff", "#b6dfaeff"))+
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5))+
  geom_jitter(width = 0.2)+
  theme(legend.position="none") +
  ggtitle("Wing length by university")
  
p

#clearly looks liek UW moz have smaller wings, should NOT make comparisons across universities
```

# Wing length by duration of cryopreservation 



```{r byTimepoint}


q <- ggplot(wing_length, aes(x = timepoint, y = wing_length))+
  geom_boxplot()+
  aes(fill=timepoint)+
  theme_pubr() +
   scale_fill_manual(values = c("#b6dfaeff", "#aadce0", "#87b6c8", "#6490b0","#406a98", "#1e466e" )) +
  theme(axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5))+
  geom_jitter(width = 0.2)+
  ggtitle("Wing length by length of cryostock freeze before use ")+
  theme(legend.position="none") 

  

q

#ggsave("./CryoWing_timepoint_boxplot.pdf", height=4, width=6, device="pdf")


#no immediate red flags in terms of development - frozen donor pools don't generate adults with tiny wings 

```

#Wing Length by treatment (matched vs mismatched microbiome transplant)


```{r byTx}
q <- ggplot(wing_length, aes(x = treatment, y = wing_length))+
  geom_boxplot()+
  aes(fill=treatment)+
  theme_pubr() +
  theme(legend.position="none") +
  scale_fill_manual(values = met.brewer("Tam", 2))+
  theme(axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5))+
  geom_jitter(width = 0.2)+
  ggtitle("Wing length by treatment")
  
  
q
#ggsave("./CryoWing_treatment_boxplot.pdf", height=4, width=6, device="pdf")

#no big dif on a global scale 
```

#Wing length by uni_tx 

Goal: Did the swap treatment affect adult body size? Did that effect differ between the UW vs LSTM mosquitoes? 

First, consider whether LSTM vs UW mosquitoes generally have different body sizes. (yes, they do!)
Then, compare the match/mismatch treatments within each university. 



```{r}


global_ttest <- t.test(wing_length ~ uni, data = wing_length)
#yes! Signficant differences between UW and LSTM wings, so we need to take into account when we make conclusions about match vs mismatch treatment. 


p <- ggplot(wing_length, aes(x = uni_tx, y = wing_length))+
  geom_boxplot()+
  aes(fill=uni)+
  theme_pubr() +
  theme(legend.position="none") +
  scale_fill_manual(values = c("#a6cee3", "#1f78b4"))+
  facet_wrap(~uni, scales = "free_x", nrow = 1) +
  theme(axis.text.x = element_text(angle = 45, size = 13, vjust = 0.5))+
  #geom_jitter(width = 0.2)+
  ggtitle("Wing length by treatment")
  
p

#ggsave("./plots/wing_length_unitx.pdf", height=5, width=4, device="pdf") 

#stats! Ttest to compare between treatments, separateed by university because we know that LSTM mosquitoes are inherently bigger. 
wings_LSTM <- filter(wing_length, uni == "LSTM")
wings_UW <- filter(wing_length, uni == "UW") 

#ok, so check for equal vairance between groups 
var.test(wing_length ~ uni_tx, wings_LSTM, 
         alternative = "two.sided")
#shit, the variance is not equal to 1 p = 0.0288
var.test(wing_length ~ uni_tx, wings_UW, 
         alternative = "two.sided")
#ok, the variance on these is equal, but I don't want to use different tests for essentially the same comparison. 
#turns out  t.test() assumes unequal variances,so that's good (welch's ttest). Go with that since it seems to be well-trusted
#http://daniellakens.blogspot.com/2015/01/always-use-welchs-t-test-instead-of.html


t.test(wing_length ~ uni_tx, data = wings_LSTM) # p = 0.05653, no difference 
t.test(wing_length ~ uni_tx, data = wings_UW) # p = 0.03943, significant! UW-match mosquitoes have bigger wings than UW-mismatch mosquitoes. 




```



# Summary stats

Basic stuff - print out tables for the mean, median, sd of each sample. Also subset by different variables of interest. Also use this as a logic check for why my x-axis is so whack 
Make smaller dataframes to plot other questions of interest - particuarly, group by:

* uni (institution where experiemnt was performed)
* timepoint (aggregated across all unis at tx, does the duration of freeze matter for development params?)
* treatment (does a mismatched microbiome have an effect?)
```{r}
wing_length %>% group_by(uni) %>% summarise(mean(wing_length),
                                              median(wing_length),
                                              sd(wing_length),
                                              n_distinct(sample_id))
              
wing_length %>% group_by(treatment) %>% summarise(mean(wing_length),
                                              median(wing_length),
                                              sd(wing_length),
                                              n_distinct(sample_id))
#slightly higher mean on mismatched samples!

wing_length %>% group_by(timepoint) %>% summarise(mean(wing_length),
                                              median(wing_length),
                                              sd(wing_length),
                                              n_distinct(sample_id))

#this one is a good summary table. Consider this a quick glance at each of the groups, separated in a logical way (i.e. each combo of timepoint/treatment/uni gets its own row)
summary_table <- wing_length %>% group_by(full_id) %>% summarise(mean(wing_length),
                                              median(wing_length),
                                              sd(wing_length),
                                              n_distinct(sample_id))




```






