---
title: "cryo_aldex2_new"
author: "Holly Nichols"
date: "2024-12-09"
output: html_document
---

# Differential Abundance Analysis with Aldex2

**NOTE: aldex2 requires and oler version of R-I used 4.1.3 **


Goals: 

**Which ASVs change over cryopreservation?**
UW-match adults - timepoint
LSTM-match adults -  timepiont 


**Which ASVs thrive in LSTM vs UW mosquitoes?**
UW-match adults vs LSTM-mismatch adults 
LSTM-match adults vs  UW-mismatch adults

**Which ASVs are more common in samples that got LSTM vs UW homogenate? (stock_origin)**
Use all adult samples, and the variable would be stock_origin



**Which ASVs correlate with host fitness?**
note: need to separate this by uni b/c LSTM generally had bigger wings and that would just confuse everything 
Also need to think about what it means to test adult samples vs wing length, because all differences in wing length are set at the larval stage. 
adult samples and wing_length
larvae samples and wing_length
water samples and wing_length




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(phyloseq)
#BiocManager::install("ALDEx2")
library(ALDEx2)
library(tibble) #for rownames to column
library(MetBrewer) #for pretty colors
library(tidyverse)
library(scales) #for graphs
library(ggrepel) #for repulsive text on volcano plots 
library(cowplot) #for theme on volcano plots 
library(ggpubr) #another theme for plots 


# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ALDEx2")


#import ps.filled 
ps.filled = readRDS("phyloseq/..data/phyloseq/crPSFilled")





```




## UW-match adults - timepoint

First, let's try aldex2 with just UW-match adults to see which taxa jump out as affected by the duration of cryopreservation (timepoint) (six levels)


```{r uw-match timepoint}


#first filter to just UW-match adults 

ps.uw_match_a <- subset_samples(ps.filled, uni_tx == "UW-match" & sample_type == "adult")
ps.uw_match_a <- prune_taxa(taxa_sums(ps.uw_match_a) > 0, ps.uw_match_a) 
#1399 taxa and 138 samples

sample_data(ps.uw_match_a)$timepoint <-as.factor(sample_data(ps.uw_match_a)$timepoint) #make uni_tx a factor

head(otu_table(ps.uw_match_a))
unique(sample_data(ps.uw_match_a)$timepoint)


#this runs quickly
aldex.clr.uwmatch_tp<-aldex.clr((as.data.frame(otu_table(ps.uw_match_a))),
                             sample_data(ps.uw_match_a)$timepoint,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test 




table(sample_data(ps.uw_match_a)$timepoint)




# this is discrete for 3+ groups, there is also an option for continuous variables

glm.kru.harvest.data.cluster<-aldex.kw(aldex.clr.uwmatch_tp) #started at 12:26 PM 7/15/24 Finished before 12:44PM 
saveRDS(glm.kru.harvest.data.cluster, "aldex2/timepoint_uwmatch_adults")

glm.kru.harvest.data.cluster = readRDS("./aldex2/timepoint_uwmatch_adults")





taxa_info <- data.frame(tax_table(ps.uw_match_a))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")
#then change the rowmans to their own column with header ASV
#need rownames_to_column



#now merge your taxa_info with the output of aldex.kw and filter for significant results
sig_aldex2 <- glm.kru.harvest.data.cluster %>%
  rownames_to_column(var = "ASV") %>%
  filter(kw.eBH < 0.05) %>%
  arrange(kw.eBH) %>%
  dplyr::select(ASV, glm.eBH, kw.eBH)
sig_aldex2 <- left_join(sig_aldex2, taxa_info)

#9 taxa! 

sig_aldex2$Genus

# [1] "Pseudomonas_E_647464"             "Achromobacter"                   
# [3] "Family_Pseudomonadaceae"          "Pandoraea"                       
# [5] "Pseudomonas_E_650326"             "Agrobacterium"                   
# [7] "Acinetobacter"                    "Family_Burkholderiaceae_A_592522"
# [9] "Pedobacter_887417"   


sig_aldex2$ASV

#need a function that would make a boxplot of relabund of each timepoint to see what'd driving differences 

#but first just do the first one 
#fdee1b26073e211295a52f694af38bbd

#needs a clear x of timepont and y of relative abundance 

#relabund teh ps object 
ps.uw_match_a_r <- transform_sample_counts(ps.uw_match_a, function(x) x/sum(x))

ps.new <- prune_taxa(rownames(tax_table(ps.uw_match_a_r)) == "fdee1b26073e211295a52f694af38bbd", ps.uw_match_a_r)
sample_sums(ps.new) 

melted<- psmelt(ps.new)


plot <- melted %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=timepoint, y=Abundance, fill=timepoint) +
  scale_fill_manual(values = met.brewer("Hokusai3", 6)) + 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  xlab("timepoint") +
  ylab("Relative Abundance")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Relabund ASV fdee1b26073e211295a52f694af38bbd by timepoint ")

plot







```




# Plot Relative Abundance - UW match adults by timepoint 

Expect 6 plots, look for taxa that are lost over time

```{r}



ps.uw_match_a_r <- transform_sample_counts(ps.uw_match_a, function(x) x/sum(x))
melted_UWM_a_r<- psmelt(ps.uw_match_a_r)

####

TargetASV = as.list(sig_aldex2$ASV)

# We use the %in% function to select only taxa that are found in the TargetGenera
selected = melted_UWM_a_r %>%
  filter(OTU %in% TargetASV)


tax.names<-unique(selected$OTU)





#start empty list for plots 
aldex_plots_uwm_adults_time<- list()

# Loop through each taxon
for(i in (1:length(tax.names))) {
  
  ASV <- (sig_aldex2$ASV)[i]
  
  Genus <- (sig_aldex2$Genus)[i]
  
  # Prune to the specific taxon 
  taxon_data <- filter(selected, OTU == ASV)
  
  # Generate the plot
  plot_obj <- taxon_data %>%
    ggplot+
    geom_boxplot(outlier.shape = NA) +
    aes(x=timepoint, y=Abundance, fill=timepoint, ) +
    scale_fill_manual(values = c("#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF", "#FFFFFF"))+ #white fill to see points better
    #coord_cartesian(ylim=c(0,1)) + # adjust y-axis

    ylab("Relative Abundance")  +
    theme_pubr() +
    theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
    geom_jitter(width = 0.2, size=1.5, 
                aes(color = timepoint)) +
   # scale_color_manual(values = c("#B22222", "#228B22", "#1E3A8A","#6A0DAD" ))+
    theme(legend.position="none", 
          axis.title.x=element_blank())+ 
    ggtitle(Genus, subtitle = ASV) + 
    theme(plot.title=element_text(hjust=0.5, face = "bold"), 
          plot.subtitle=element_text(hjust=0.5))

  # Save the plot to the list
  aldex_plots_uwm_adults_time[[i]] <- plot_obj
}



do.call(plot_grid, c(aldex_plots_uwm_adults_time, ncol = 3))

ggsave("./aldex2/uw_match_adults_time.pdf", height = 12, width=12, device="pdf")

# None really decrease over time- Pseudomonas is highest at 0h and d10, but also present at other tiems?


```





# LSTM-match adults - timepoint  


```{r lstmmatchtimepoint}

#first filter to just UW-match adults 

ps.LSTM_match_a <- subset_samples(ps.filled, uni_tx == "LSTM-match" & sample_type == "adult")
ps.LSTM_match_a <- prune_taxa(taxa_sums(ps.LSTM_match_a) > 0, ps.LSTM_match_a) 
#1328 taxa and 124 samples

sample_data(ps.LSTM_match_a)$timepoint <-as.factor(sample_data(ps.LSTM_match_a)$timepoint) #make uni_tx a factor




#this runs quickly
aldex.clr.LSTMmatch_tp<-aldex.clr((as.data.frame(otu_table(ps.LSTM_match_a))),
                             sample_data(ps.LSTM_match_a)$timepoint,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test 
table(sample_data(ps.LSTM_match_a)$timepoint)



glm.kru.lstmmatch_tp<-aldex.kw(aldex.clr.LSTMmatch_tp) #takes a while
saveRDS(glm.kru.lstmmatch_tp, "aldex2/timepoint_LSTMmatch_adults")

glm.kru.lstmmatch_tp = readRDS("./aldex2/timepoint_LSTMmatch_adults")


#grab tax info from the same phyloseq object
taxa_info <- data.frame(tax_table(ps.LSTM_match_a))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")
#then change the rowmans to their own column with header ASV
#need rownames_to_column



#now merge your taxa_info with the output of aldex.kw and filter for significant results
sig_aldex2 <- glm.kru.lstmmatch_tp %>%
  rownames_to_column(var = "ASV") %>%
  filter(kw.eBH < 0.05) %>%
  arrange(kw.eBH) 

sig_aldex2 <- left_join(sig_aldex2, taxa_info)
#14 ASVs differnt by timepoint in the LSTM tx


```


# Plots LSTM-match ASVs - timepoint

#try as a function this time!

```{r}


make_asv_boxplots <- function(
  melted_df,
  target_asvs,
  asv_metadata_df,
  x_var
) {
  
  # Ensure symbols work inside ggplot/dplyr
  x_var <- rlang::ensym(x_var)
  
  # Filter to ASVs of interest
  selected <- melted_df %>%
    dplyr::filter(OTU %in% target_asvs)
  
  # Initialize list
  plot_list <- list()
  
  # Loop through ASVs
  for (i in seq_along(target_asvs)) {
    
    ASV <- target_asvs[i]
    
    Genus <- asv_metadata_df %>%
      dplyr::filter(ASV == !!ASV) %>%
      dplyr::pull(Genus)
    
    taxon_data <- selected %>%
      dplyr::filter(OTU == ASV)
    
    p <- ggplot(taxon_data, aes(x = !!x_var, y = Abundance)) +
      geom_boxplot(outlier.shape = NA, fill = "white") +
      geom_jitter(
        aes(color = !!x_var),
        width = 0.2,
        size = 1.5
      ) +
      ylab("Relative Abundance") +
      theme_pubr() +
      theme(
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.position = "none",
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)
      ) +
      ggtitle(Genus, subtitle = ASV)
    
    plot_list[[ASV]] <- p
  }
  
  return(plot_list)
}


#prep a phyloseq object for plotting
ps.lsmt_match_a_r <- transform_sample_counts(ps.LSTM_match_a, function(x) x/sum(x))
melted_LSTMM_a_r<- psmelt(ps.lsmt_match_a_r)

####


#call the function 
plots_lstmM_adults_time <- make_asv_boxplots(
  melted_df        = melted_LSTMM_a_r,
  target_asvs      = sig_aldex2$ASV,
  asv_metadata_df  = sig_aldex2,
  x_var            = timepoint
)

do.call(plot_grid, c(plots_lstmM_adults_time, list(ncol = 5))) #plot it


ggsave("./aldex2/lstm_match_adults_time.pdf", height = 18, width=18, device="pdf")


# more common at early timepoints - 
#Vibrio (common!)****
#Pseudoatlteromonas 
#Neisseria (really low abundance )
# corynebacterium (skin microbe? contam?) 
#Leucobacter
#Sphingobacterium ****
#Comamonas - though tht one rebounds at later timepoints 



```





# Fig 5 - same stock_origin, different mosquito host  - LSTM homogenate

(Fig 5)

Test LSTM-match vs UW-mismatch, this should get at the different community members that dominate when the same homogenate goes into two different hosts

Question -- should i subset this to just d16,d30, and mo6? Yes, it will allow to compare apples to apples since they have both a a match and mismatch treatment 



Significant ASVs - Seven ASVs that all seem to be more common in UW-mismatch, nothing that's associated with LSTM-match. 
Interpretation of figures:

* Dispersion (x-axis) describes the differences within groups - larger values are more variable within a group. 
* Difference (y-axis) describes the effect size. Larger values are more likely to be truly biologically significant. 




## Fig 5 - Adults that got UW homogenate

10 ASVs jump out, 2 are more common in LSTM-mismatch (Elizabethkingia and an uncl. Burkholderiaceae). 
Note that this is subset to just d16, d30, and mo6 timepoints because we know that timepoint has some effect on microbial community. 




```{r volcano}

# #subset to d16, d30, mo6
# #subset to just adults with stock_origin = UW
#
#
ps.gotUW<- subset_samples(ps.filled, uni_tx == "UW-match" | uni_tx == "LSTM-mismatch" & sample_type == "adult")
#253 samples
ps.gotUW <- subset_samples(ps.gotUW, timepoint == "d16" | timepoint == "d30" | timepoint == "mo6")

ps.gotUW <- prune_taxa(taxa_sums(ps.gotUW) > 0, ps.gotUW)
#12/14/25 - 1612 taxa and 162 samples 

sample_data(ps.gotUW)$uni_tx <-as.factor(sample_data(ps.gotUW)$uni_tx) #make uni_tx a factor

unique(sample_data(ps.gotUW)$timepoint)




#this runs quickly
aldex.clr.gotUW<-aldex.clr((as.data.frame(otu_table(ps.gotUW))),
                             sample_data(ps.gotUW)$uni_tx,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test
table(sample_data(ps.gotUW)$uni)
#79 LSTM, 83 UW

library(ggrepel)


ttest<-aldex.ttest(aldex.clr.gotUW, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.gotUW, CI=T, verbose=FALSE)

aldex.df <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)

par(mfrow=c(1,2))
aldex.plot(aldex.df, type="MA", test="welch")
aldex.plot(aldex.df, type="MW", test="welch")


#if ttest and effect size both significant, put "both" in a new column
aldex.df$one_or_two<-"one"
aldex.df$one_or_two[which(aldex.df$we.eBH < 0.05 & aldex.df$wi.eBH < 0.05)] <-"both"


#get taxa info from original phyloseq object
taxa_info <- data.frame(tax_table(ps.gotUW))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")


aldex.df<-aldex.df %>% rownames_to_column(var = "ASV")

aldex.df <- left_join(aldex.df, taxa_info) #add data info to aldexdf

sig_aldex2 <- aldex.df %>% filter(one_or_two=="both")#(not saved, just for looking - filter for just significant ASVs
#9 come up!


#figure out which families/classes are represented, and dots color by those 

table(sig_aldex2$Class) #8 gammas and 1 bacteroidia
table(sig_aldex2$Order)
table(sig_aldex2$Family)
table(sig_aldex2$Genus) 
#use family 


aldex.df.edit_UW <- aldex.df %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Family == "Burkholderiaceae_A_580492" ~ "#B34657",
                          Family == "Burkholderiaceae_A_592522" ~ "#B34657",
                          Family == "Burkholderiaceae_A_595422" ~ "#B34657",
                          Family == "Moraxellaceae" ~ "#87C080",
                          Family == "Pseudomonadaceae" ~ "#45884E",
                          Family == "Weeksellaceae" ~ "#588FBF",
                          Family == "Xanthomonadaceae_616009" ~ "#E1E04B",
                          TRUE ~ "black"
                        ), 
                        "grey"))



volcano_gotUW <-ggplot(aldex.df.edit_UW, aes(x=diff.win,y=diff.btw)) +
  xlab("Log2 Dispersion")+ ylab("Log2 Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=0, linetype="dashed") + geom_abline(slope=-1, intercept=0, linetype="dashed")+
    scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13))
volcano_gotUW

#ggsave("./aldex2/volcano_got_UW.pdf", height = 5, width=5, device="pdf")



```




## Fig 5 - Adults that got LSTM homogenate  

Remember, this is just the d16, d30, and mo6 treatments

```{r gotLSTM_aldex2}

#subset to d16, d30, mo6
#subset to just adults with stock_origin = LSTM


ps.gotLSTM<- subset_samples(ps.filled, uni_tx == "LSTM-match" | uni_tx == "UW-mismatch" & sample_type == "adult")
#220 samples
ps.gotLSTM <- subset_samples(ps.gotLSTM, timepoint == "d16" | timepoint == "d30" | timepoint == "mo6")

ps.gotLSTM <- prune_taxa(taxa_sums(ps.gotLSTM) > 0, ps.gotLSTM) 
# 1347 taxa and 144 samples 

sample_data(ps.gotLSTM)$uni_tx <-as.factor(sample_data(ps.gotLSTM)$uni_tx) #make uni_tx a factor


#this runs quickly
aldex.clr.gotLSTM<-aldex.clr((as.data.frame(otu_table(ps.gotLSTM))),
                             sample_data(ps.gotLSTM)$uni_tx,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)





ttest<-aldex.ttest(aldex.clr.gotLSTM, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.gotLSTM, CI=T, verbose=FALSE)

aldex.df <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)

par(mfrow=c(1,2))
aldex.plot(aldex.df, type="MA", test="welch")
aldex.plot(aldex.df, type="MW", test="welch")


#if ttest adn effect size both significant, put "both" in a new column
aldex.df$one_or_two<-"one"
aldex.df$one_or_two[which(aldex.df$we.eBH < 0.05 & aldex.df$wi.eBH < 0.05)] <-"both"


#get taxa info from original phyloseq object
taxa_info <- data.frame(tax_table(ps.gotLSTM))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")


aldex.df<-aldex.df %>% rownames_to_column(var = "ASV")

aldex.df <- left_join(aldex.df, taxa_info) #add data info to aldexdf

aldex.df %>% filter(one_or_two=="both")#filter for just significant ASVs
#8 come up!


aldex.df.edit_gotLSTM <- aldex.df %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Family == "Burkholderiaceae_A_580492" ~ "#B34657",
                          Family == "Burkholderiaceae_A_592522" ~ "#B34657",
                          Family == "Burkholderiaceae_A_595422" ~ "#B34657",
                          Family == "Moraxellaceae" ~ "#87C080",
                          Family == "Pseudomonadaceae" ~ "#45884E",
                          Family == "Weeksellaceae" ~ "#588FBF",
                          Family == "Xanthomonadaceae_616009" ~ "#E1E04B",
                          TRUE ~ "black"
                        ), 
                        "grey"))



#edit


volcano_gotLSTM <-ggplot(aldex.df.edit_gotLSTM, aes(x=diff.win,y=diff.btw)) +
  xlab("Log2 Dispersion")+ ylab("Log2 Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=0, linetype="dashed") + geom_abline(slope=-1, intercept=0, linetype="dashed")+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13))




ggarrange(volcano_gotUW, volcano_gotLSTM)

#ggsave("./plots/volcano_swap_effect.pdf", height=4.5, width=8.5, device="pdf") 



```






# Are those ASVS in UW moz the same? 

```{r venn}

###########12/14
#do the 8 ASVS from got LSTM overlaps with the 9 from gotUW? 
intersect((aldex.df.edit_gotLSTM %>% filter(one_or_two=="both"))$ASV, sig_aldex2$ASV)
intersect(sig_aldex2$ASV,(aldex.df.edit_gotLSTM %>% filter(one_or_two=="both"))$ASV)
#Nope! No overlap. Even though it's the same genera, they're different ASVs. Cool!
```


# How common are those ASVs in those moz? 


```{r}


ASVs_gotLSTM <- aldex.df.edit_gotLSTM %>% filter(one_or_two=="both") %>% .$ASV #8 taxa
ASVs_got_UW <- aldex.df.edit_UW %>% filter(one_or_two=="both") %>% .$ASV #9 taxa
sig_fig5 <- c(ASVs_gotLSTM,ASVs_got_UW)

#Awsome! now make a quick relabund 

ps.relabund
ps.fig5 <- prune_taxa(sig_fig5, ps.relabund) %>% subset_samples(uni == "UW" & sample_type == "adult")

#now plot! 


melt_fig5 <- psmelt(ps.fig5)



StackedBarPlot <- melt_fig5 %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance",title = "Fig5 Sig ASVs Relative Abundance")+ 
   facet_wrap(~interaction(sample_type,uni_tx), scales = "free_x", nrow = 2) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))
StackedBarPlot

#These taxa make up a large portion of adults reads - usually > 75% in UW-mismatch, and often > 50% in UW-match. 

```





# LSTM larvae vs LSTM water 
not separating out match + mismatch

No significant ASVs


```{r}

ps.lw_LSTM<- subset_samples(ps.filled, sample_type == "larvae" | sample_type == "water" & uni == "LSTM")
ps.lw_LSTM <- prune_taxa(taxa_sums(ps.lw_LSTM) > 0, ps.lw_LSTM) 
# 751 taxa and 92 samples 

sample_data(ps.lw_LSTM)$sample_type <-as.factor(sample_data(ps.lw_LSTM)$sample_type) #make uni_tx a factor


#this runs quickly
aldex.clr.larvWater_LSTM <- aldex.clr((as.data.frame(otu_table(ps.lw_LSTM))),
                             sample_data(ps.lw_LSTM)$sample_type,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test 


#now merge your taxa_info with the output of aldex.kw and filter for significant results
# sig_aldex2 <- glm.kru.larvWater_LSTM %>%
#   rownames_to_column(var = "ASV") %>%
#   filter(kw.eBH < 0.05) %>%
#   arrange(kw.eBH) %>%
#   dplyr::select(ASV, glm.eBH, kw.eBH)
# sig_aldex2 <- left_join(sig_aldex2, taxa_info)


#try this by t-test and effect size instead since it's just 2 conditions 
ttest<-aldex.ttest(aldex.clr.larvWater_LSTM, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.larvWater_LSTM, CI=T, verbose=FALSE)


aldex.df_larvWater_LSTM <- data.frame(ttest,effect) 


#if ttest and effect size both significant, put "both" in a new column
aldex.df_larvWater_LSTM$one_or_two<-"one"
aldex.df_larvWater_LSTM$one_or_two[which(aldex.df_larvWater_LSTM$we.eBH < 0.05 & aldex.df_larvWater_LSTM$wi.eBH < 0.05)] <-"both"


#get taxa info from original phyloseq object
taxa_info <- data.frame(tax_table(ps.lw_LSTM))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")


aldex.df_larvWater_LSTM<-aldex.df_larvWater_LSTM %>% rownames_to_column(var = "ASV")

aldex.df_larvWater_LSTM <- left_join(aldex.df_larvWater_LSTM, taxa_info) #add data info to aldexdf

aldex.df_larvWater_LSTM %>% filter(one_or_two=="both")
#Zero significant ASVs. 

```




# UW larvae vs water 

Actually 12 ASVs! All more common in water 
This is bulk including both UW-match and UW-mismatch.... 

```{r}

ps.lw_UW<- subset_samples(ps.filled, sample_type == "larvae" | sample_type == "water" & uni == "UW")
ps.lw_UW <- prune_taxa(taxa_sums(ps.lw_UW) > 0, ps.lw_UW) 
# 390 taxa and 92 samples 

sample_data(ps.lw_UW)$sample_type <-as.factor(sample_data(ps.lw_UW)$sample_type) #make uni_tx a factor



#this runs quickly
aldex.clr.larvWater_UW <- aldex.clr((as.data.frame(otu_table(ps.lw_UW))),
                             sample_data(ps.lw_UW)$sample_type,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test 



#since it's just two conditions,  use ttest and effect module 
ttest<-aldex.ttest(aldex.clr.larvWater_UW, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.larvWater_UW, CI=T, verbose=FALSE)

aldex.df_UWmatch_WL <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)


#if ttest and effect size both significant, put "both" in a new column
aldex.df_UWmatch_WL$one_or_two<-"one"
aldex.df_UWmatch_WL$one_or_two[which(aldex.df_UWmatch_WL$we.eBH < 0.05 & aldex.df_UWmatch_WL$wi.eBH < 0.05)] <-"both"



#get taxa info from original phyloseq object
taxa_info <- data.frame(tax_table(ps.lw_UW))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")




aldex.df_UWmatch_WL<-aldex.df_UWmatch_WL %>% rownames_to_column(var = "ASV")

aldex.df_UWmatch_WL <- left_join(aldex.df_UWmatch_WL, taxa_info) #add data info to aldexdf

aldex.df_UWmatch_WL %>% filter(one_or_two=="both")#filter for just significant ASVs, not stored 
#11 come up!

#figure out which Classes are represented, and color by those 
#also count how many families there are 


table(sig_aldex2$Class)
table(sig_aldex2$Order)
table(sig_aldex2$Family)
#use class 


aldex.df_UWmatch_WL_edit <- aldex.df_UWmatch_WL %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Class == "Gammaproteobacteria" ~ "#238b45",
                          Class == "Bacteroidia" ~ "#4292c6",
                          Class == "Actinomycetia" ~ "#ff7f00",
                          Class == "Alphaproteobacteria" ~ "#BBDE94",
                          TRUE ~ "black"
                        ), 
                        "grey"))



volcano_uwm_water_larvae <-ggplot(aldex.df_UWmatch_WL_edit, aes(x=diff.win,y=diff.btw)) +
  xlab("Dispersion")+ ylab("Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
 geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  #legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=-4, linetype="dashed") + geom_abline(slope=-1, intercept=4, linetype="dashed")+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13)) + 
  ggtitle("Differential abundance of ASVs in \n UW water (top) vs larvae (bottom) samples")

volcano_uwm_water_larvae

#huh, all more common in water than in larvae. Nothing that's strongly larvae-associated. 

#ggsave("./plots/volcano_swap_effect_UW_water_v_larvae.pdf", height=6, width=6, device="pdf") 


```



#UW-match larvae vs water 


```{r}


ps.lw_UW_match<- subset_samples(ps.filled, sample_type == "larvae" | sample_type == "water") %>% subset_samples(uni_tx == "UW-match")
ps.lw_UW_match <- prune_taxa(taxa_sums(ps.lw_UW_match) > 0, ps.lw_UW_match) 
# 64 taxa and 36 samples 

sample_data(ps.lw_UW_match)$sample_type <-as.factor(sample_data(ps.lw_UW_match)$sample_type) #make uni_tx a factor



#this runs quickly
aldex.clr.larvWater_UW_match <- aldex.clr((as.data.frame(otu_table(ps.lw_UW_match))),
                             sample_data(ps.lw_UW_match)$sample_type,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test 


#grab tax info from the same phyloseq object
taxa_info <- data.frame(tax_table(ps.lw_UW_match))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")
#then change the rowmans to their own column with header ASV
#need rownames_to_column



#since it's just two conditions, might be more appropriate to use ttest and effect module 
ttest<-aldex.ttest(aldex.clr.larvWater_UW_match, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.larvWater_UW_match, CI=T, verbose=FALSE)

aldex.df_UWmatch_WL <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)


#if ttest and effect size both significant, put "both" in a new column
aldex.df_UWmatch_WL$one_or_two<-"one"
aldex.df_UWmatch_WL$one_or_two[which(aldex.df_UWmatch_WL$we.eBH < 0.05 & aldex.df_UWmatch_WL$wi.eBH < 0.05)] <-"both"


#now merge your taxa_info with the output of aldex.kw and filter for significant results
sig_aldex2 <- aldex.df_UWmatch_WL %>%
  rownames_to_column(var = "ASV") %>%
  filter(one_or_two == "both") %>%
  arrange(we.eBH) %>%
  dplyr::select(ASV, we.eBH, wi.eBH)
sig_aldex2 <- left_join(sig_aldex2, taxa_info)
#3 taxa 


# 
# #get taxa info from original phyloseq object
# taxa_info <- data.frame(tax_table(ps.lw_UW_match))
# taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")




aldex.df_UWmatch_WL<-aldex.df_UWmatch_WL %>% rownames_to_column(var = "ASV")

aldex.df_UWmatch_WL <- left_join(aldex.df_UWmatch_WL, taxa_info) #add data info to aldexdf

aldex.df_UWmatch_WL %>% filter(one_or_two=="both")#filter for just significant ASVs, not stored
#12 come up!



table(sig_aldex2$Class)
table(sig_aldex2$Order)
table(sig_aldex2$Family)
#use class 

aldex.df_UWmatch_WL_edit <- aldex.df_UWmatch_WL %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Class == "Gammaproteobacteria" ~ "#238b45",
                          Class == "Bacteroidia" ~ "#4292c6",
                          Class == "Actinomycetia" ~ "#ff7f00",
                          Class == "Alphaproteobacteria" ~ "#BBDE94",
                          TRUE ~ "black"
                        ), 
                        "grey"))



volcano_uwm_water_larvae <-ggplot(aldex.df_UWmatch_WL_edit, aes(x=diff.win,y=diff.btw)) +
  xlab("Dispersion")+ ylab("Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
 geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  #legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=0, linetype="dashed") + geom_abline(slope=-1, intercept=0, linetype="dashed")+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13)) + 
  ggtitle("Differential abundance of ASVs in \n UW-match water (top) vs larvae (bottom) samples")

volcano_uwm_water_larvae

#two chryseobacterium ASVs more abundant in water, Leucobacter more abundant in larvae

#ggsave("./plots/volcano_swap_effect_UWmatch_water_v_larvae.pdf", height=6, width=6, device="pdf") 



```




# Plot relative abundance for  UW-match larvae vs water (one plot, using ggplot)

Chryseobacrium in water, Leucobacter in larvae

AF22

```{r}


ps.UWM_WL_r <- transform_sample_counts(ps.lw_UW_match, function(x) x/sum(x))
melt_UWM_WL_r = psmelt(ps.UWM_WL_r)


TargetASV = as.list(sig_aldex2$ASV)

# We use the %in% function to select only taxa that are found in the TargetGenera
selected = melt_UWM_WL_r %>%
  filter(OTU %in% TargetASV)


tax.names<-unique(selected$OTU)
temp <- tax.names




#start empty list for plots 
aldex_plots_uwm_water_larvae<- list()

# Loop through each taxon
for(i in (1:length(tax.names))) {
  
  ASV <- (sig_aldex2$ASV)[i]
  
  Genus <- (sig_aldex2$Genus)[i]
  
  # Prune to the specific taxon 
  taxon_data <- filter(selected, OTU == ASV)
  
  # Generate the plot
  plot_obj <- taxon_data %>%
    ggplot+
    geom_boxplot(outlier.shape = NA) +
    aes(x=sample_type, y=Abundance, fill=sample_type, ) +
    scale_fill_manual(values = c("#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF"))+
    #coord_cartesian(ylim=c(0,1)) + # adjust y-axis

    ylab("Relative Abundance")  +
    theme_pubr() +
    theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
    geom_jitter(width = 0.2, size=1.5, 
                aes(color = sample_type)) +
   # scale_color_manual(values = c("#B22222", "#228B22", "#1E3A8A","#6A0DAD" ))+
    theme(legend.position="none", 
          axis.title.x=element_blank())+ 
    ggtitle(Genus, subtitle = ASV) + 
    theme(plot.title=element_text(hjust=0.5, face = "bold"), 
          plot.subtitle=element_text(hjust=0.5))

  # Save the plot to the list
  aldex_plots_uwm_water_larvae[[i]] <- plot_obj
}



do.call(plot_grid, c(aldex_plots_uwm_water_larvae, ncol = 3))

ggsave("./aldex2/uw_match_water_v_larvae/uwm_water_larvae_grid.pdf", height = 5, width=12, device="pdf")

```

# UW-mismatch water vs larvae



```{r}



ps.lw_UW_mismatch<- subset_samples(ps.filled, sample_type == "larvae" | sample_type == "water") %>% subset_samples(uni_tx == "UW-mismatch")
ps.lw_UW_mismatch <- prune_taxa(taxa_sums(ps.lw_UW_mismatch) > 0, ps.lw_UW_mismatch) 
# 69 taxa and 20 samples 

sample_data(ps.lw_UW_mismatch)$sample_type <-as.factor(sample_data(ps.lw_UW_mismatch)$sample_type) #make uni_tx a factor



#this runs quickly
aldex.clr.larvWater_UW_mismatch <- aldex.clr((as.data.frame(otu_table(ps.lw_UW_mismatch))),
                             sample_data(ps.lw_UW_mismatch)$sample_type,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test, just meeting it on UW-mismatch


#grab tax info from the same phyloseq object
taxa_info <- data.frame(tax_table(ps.lw_UW_mismatch))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")
#then change the rowmans to their own column with header ASV
#need rownames_to_column



#since it's just two conditions, might be more appropriate to use ttest and effect module 
ttest<-aldex.ttest(aldex.clr.larvWater_UW_mismatch, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.larvWater_UW_mismatch, CI=T, verbose=FALSE)

aldex.df_UWmismatch_WL <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)


#if ttest and effect size both significant, put "both" in a new column
aldex.df_UWmismatch_WL$one_or_two<-"one"
aldex.df_UWmismatch_WL$one_or_two[which(aldex.df_UWmismatch_WL$we.eBH < 0.05 & aldex.df_UWmismatch_WL$wi.eBH < 0.05)] <-"both"


#now merge your taxa_info with the output of aldex.kw and filter for significant results
sig_aldex2 <- aldex.df_UWmismatch_WL %>%
  rownames_to_column(var = "ASV") %>%
  filter(one_or_two == "both") %>%
  arrange(we.eBH) %>%
  dplyr::select(ASV, we.eBH, wi.eBH)
sig_aldex2 <- left_join(sig_aldex2, taxa_info)
#2 taxa 




aldex.df_UWmismatch_WL<-aldex.df_UWmismatch_WL %>% rownames_to_column(var = "ASV")
aldex.df_UWmismatch_WL <- left_join(aldex.df_UWmismatch_WL, taxa_info) #add taxa info to aldexdf
aldex.df_UWmismatch_WL %>% filter(one_or_two=="both")#filter for just significant ASVs, not stored



aldex.df_UWmismatch_WL_edit <- aldex.df_UWmismatch_WL %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Class == "Gammaproteobacteria" ~ "#238b45",
                          Class == "Bacteroidia" ~ "#4292c6",
                          Class == "Actinomycetia" ~ "#ff7f00",
                          Class == "Alphaproteobacteria" ~ "#BBDE94",
                          TRUE ~ "black"
                        ), 
                        "grey"))



volcano_uwm_water_larvae <-ggplot(aldex.df_UWmismatch_WL_edit, aes(x=diff.win,y=diff.btw)) +
  xlab("Dispersion")+ ylab("Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
 geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  #legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=-4, linetype="dashed") + geom_abline(slope=-1, intercept=4, linetype="dashed")+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13)) + 
  ggtitle("Differential abundance of ASVs in \n UW water (top) vs larvae (bottom) samples")

volcano_uwm_water_larvae

#huh, all more common in water than in larvae. Nothing that's strongly larvae-associated, makes since b/c larvae are swimming in water that they contain a subeset of water taxa 

#ggsave("./plots/volcano_swap_effect_UWmis_water_v_larvae.pdf", height=6, width=6, device="pdf") 

```

# Plot UW-mismatch water vs larvae 

AF23

```{r}


ps.UWMis_WL_r <- transform_sample_counts(ps.lw_UW_mismatch, function(x) x/sum(x))
melt_UWMis_WL_r = psmelt(ps.UWMis_WL_r)


TargetASV = as.list(sig_aldex2$ASV)

# We use the %in% function to select only taxa that are found in the TargetGenera
selected = melt_UWMis_WL_r %>%
  filter(OTU %in% TargetASV)


tax.names<-unique(selected$OTU)
temp <- tax.names




#start empty list for plots 
aldex_plots_uwmis_water_larvae<- list()

# Loop through each taxon
for(i in (1:length(tax.names))) {
  
  ASV <- (sig_aldex2$ASV)[i]
  
  Genus <- (sig_aldex2$Genus)[i]
  
  # Prune to the specific taxon 
  taxon_data <- filter(selected, OTU == ASV)
  
  # Generate the plot
  plot_obj <- taxon_data %>%
    ggplot+
    geom_boxplot(outlier.shape = NA) +
    aes(x=sample_type, y=Abundance, fill=sample_type, ) +
    scale_fill_manual(values = c("#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF"))+
    #coord_cartesian(ylim=c(0,1)) + # adjust y-axis

    ylab("Relative Abundance")  +
    theme_pubr() +
    theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
    geom_jitter(width = 0.2, size=1.5, 
                aes(color = sample_type)) +
   # scale_color_manual(values = c("#B22222", "#228B22", "#1E3A8A","#6A0DAD" ))+
    theme(legend.position="none", 
          axis.title.x=element_blank())+ 
    ggtitle(Genus, subtitle = ASV) + 
    theme(plot.title=element_text(hjust=0.5, face = "bold"), 
          plot.subtitle=element_text(hjust=0.5))

  # Save the plot to the list
  aldex_plots_uwmis_water_larvae[[i]] <- plot_obj
}



do.call(plot_grid, c(aldex_plots_uwmis_water_larvae, ncol = 2))

ggsave("./aldex2/uw_mismatch_water_v_larvae/uwmis_water_larvae_grid.pdf", height = 5, width=9, device="pdf")

```



# What ASVs are lost in the transition from water >> adult? 


Does it make sense to consider larvae + water together for this one? What's in aquatic stage versus the adult stage? Yes considering how water + larvae nealry always  clustered together in rpca + phylo-rpca 




```{r}


head(sample_data(ps.filled))

#need a new metadata variable that combines water + larvae into "aquatic" 

#make a new column where you've manually renamed based on old values. keeps both columns 
# Modify sample_data directly to add the new column
sample_data(ps.filled)$stage <- with(sample_data(ps.filled), 
  ifelse(sample_type == "water", "aquatic",
  ifelse(sample_type == "larvae", "aquatic",
  ifelse(sample_type == "adult", "adult", as.character(sample_type)))) #retain donor pool, we'll filter those out later anyway
)



ps.stage_UWmatch<- subset_samples(ps.filled, stage == "aquatic" | stage == "adult") %>%  subset_samples(uni_tx == "UW-match")
ps.stage_UWmatch <- prune_taxa(taxa_sums(ps.stage_UWmatch) > 0, ps.stage_UWmatch) 
# 1435 taxa and 174 samples 

sample_data(ps.stage_UWmatch)$stage <-as.factor(sample_data(ps.stage_UWmatch)$stage) #make uni_tx a factor


#this runs quickly
aldex.clr.stage_UWmatch<- aldex.clr((as.data.frame(otu_table(ps.stage_UWmatch))),
                             sample_data(ps.stage_UWmatch)$stage,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test 




#grab tax info from the same phyloseq object
taxa_info <- data.frame(tax_table(ps.stage_UWmatch))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")
#then change the rowmans to their own column with header ASV
#need rownames_to_column





ttest<-aldex.ttest(aldex.clr.stage_UWmatch, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.stage_UWmatch, CI=T, verbose=FALSE)

aldex.df_UWmatch_stage <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)





#if ttest adn effect size both significant, put "both" in a new column
aldex.df_UWmatch_stage$one_or_two<-"one"
aldex.df_UWmatch_stage$one_or_two[which(aldex.df_UWmatch_stage$we.eBH < 0.05 & aldex.df_UWmatch_stage$wi.eBH < 0.05)] <-"both"


#now merge your taxa_info with the output of aldex.kw and filter for significant results
sig_aldex2 <- aldex.df_UWmatch_stage %>%
  rownames_to_column(var = "ASV") %>%
  filter(one_or_two == "both") %>%
  arrange(we.eBH) %>%
  dplyr::select(ASV, we.eBH, wi.eBH)
sig_aldex2 <- left_join(sig_aldex2, taxa_info)
#12 taxa 



aldex.df_UWmatch_stage<-aldex.df_UWmatch_stage %>% rownames_to_column(var = "ASV")

aldex.df_UWmatch_stage <- left_join(aldex.df_UWmatch_stage, taxa_info) #add data info to aldexdf

aldex.df_UWmatch_stage %>% filter(one_or_two=="both")#filter for just significant ASVs, not stored 
#12 via effect size and t-test,


#add info for plotting
aldex.df_UWmatch_stage_edit <- aldex.df_UWmatch_stage %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Class == "Gammaproteobacteria" ~ "#238b45",
                          Class == "Bacteroidia" ~ "#4292c6",
                          Class == "Actinomycetia" ~ "#ff7f00",
                          Class == "Alphaproteobacteria" ~ "#BBDE94",
                          TRUE ~ "black"
                        ), 
                        "grey"))




volcano_uwm_stage <-ggplot(aldex.df_UWmatch_stage_edit, aes(x=diff.win,y=diff.btw)) +
  xlab("Dispersion")+ ylab("Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
 geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=0, linetype="dashed") + geom_abline(slope=-1, intercept=0, linetype="dashed")+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13)) + 
  ggtitle("Differential abundance of ASVs in \n UW-match aquatic(top) and adult (bottom) samples")


ggsave("./aldex2/volcano_UWmatch_aquatic_v_adult.pdf", height=6, width=6, device="pdf") 



```


# Plot UW-match stage significant taxa 

AF19

```{r}

ps.UWM_stage_r <- transform_sample_counts(ps.stage_UWmatch, function(x) x/sum(x))
melt_UWM_stage_r = psmelt(ps.UWM_stage_r)


TargetASV = as.list(sig_aldex2$ASV)

# We use the %in% function to select only taxa that are found in the TargetGenera
selected = melt_UWM_stage_r %>%
  filter(OTU %in% TargetASV)


tax.names<-unique(selected$OTU)
temp <- tax.names



#start empty list for plots 
aldex_plots_UWM_stage<- list()

# Loop through each taxon
for(i in (1:length(tax.names))) {
  
  ASV <- (sig_aldex2$ASV)[i]
  
  Genus <- (sig_aldex2$Genus)[i]
  
  # Prune to the specific taxon 
  taxon_data <- filter(selected, OTU == ASV)
  
  # Generate the plot
  plot_obj <- taxon_data %>%
    ggplot+
    geom_boxplot(outlier.shape = NA) +
    aes(x=stage, y=Abundance, fill=stage, ) +
    scale_fill_manual(values = c("#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF","#FFFFFF"))+
    #coord_cartesian(ylim=c(0,1)) + # adjust y-axis

    ylab("Relative Abundance")  +
    theme_pubr() +
    theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
    geom_jitter(width = 0.2, size=1.5, 
                aes(color = stage)) +
   # scale_color_manual(values = c("#B22222", "#228B22", "#1E3A8A","#6A0DAD" ))+
    theme(legend.position="none", 
          axis.title.x=element_blank())+ 
    ggtitle(Genus, subtitle = ASV) + 
    theme(plot.title=element_text(hjust=0.5, face = "bold"), 
          plot.subtitle=element_text(hjust=0.5))

  # Save the plot to the list
  aldex_plots_UWM_stage[[i]] <- plot_obj
}



do.call(plot_grid, c(aldex_plots_UWM_stage, ncol = 4))

ggsave("./aldex2/grid_UW_match_stage.pdf", height = 15, width=15, device="pdf")

```




# Try it as a function 


```{r}


run_aldex2_analysis <- function(ps_object, 
                                filter_value,
                                grouping_var = "stage",
                                output_prefix = "aldex2_output",
                                mc.samples = 128) {



  #────────────────────────────────────────────
  # 0. Filter phyloseq object
  #────────────────────────────────────────────
  
  # filter_var must be a character name of a sample_data column
  ps_filt <- subset_samples(ps_object, eval(parse(text = paste0("uni_tx == '", filter_value, "'"))))
  ps_filt <- prune_taxa(taxa_sums(ps_filt) > 0, ps_filt)



  sample_data(ps_filt)[[grouping_var]] <- as.factor(sample_data(ps_filt)[[grouping_var]])

  #────────────────────────────────────────────
  # 1. Run ALDEx2
  #────────────────────────────────────────────
  aldex_clr <- aldex.clr(
    as.data.frame(otu_table(ps_filt)),
    sample_data(ps_filt)[[grouping_var]],
    mc.samples = mc.samples,
    denom = "all",
    verbose = TRUE,
    useMC = FALSE
  )

  ttest <- aldex.ttest(aldex_clr)
  effect <- aldex.effect(aldex_clr)

  aldex_df <- data.frame(ttest, effect) %>%
    rownames_to_column("ASV")

  #────────────────────────────────────────────
  # 2. Add significance classification
  #────────────────────────────────────────────
  aldex_df <- aldex_df %>%
    mutate(one_or_two = ifelse(we.eBH < 0.05 & wi.eBH < 0.05, "both", "one"))

  #────────────────────────────────────────────
  # 3. Add taxonomy data
  #────────────────────────────────────────────
  taxa_info <- data.frame(tax_table(ps_filt)) %>%
    rownames_to_column("ASV")

  aldex_df <- left_join(aldex_df, taxa_info, by = "ASV")

  sig_df <- aldex_df %>% filter(one_or_two == "both")

  message("Significant ASVs: ", nrow(sig_df))

  #────────────────────────────────────────────
  # 4. Volcano plot
  #────────────────────────────────────────────
  aldex_plot_df <- aldex_df %>%
    mutate(pointlabel = if_else(one_or_two == "both", Genus, ""),
           size = if_else(one_or_two == "both", 3, 2),
           famcolor = case_when(
             one_or_two != "both" ~ "grey",
             Class == "Gammaproteobacteria" ~ "#238b45",
             Class == "Bacteroidia" ~ "#4292c6",
             Class == "Actinomycetia" ~ "#ff7f00",
             Class == "Alphaproteobacteria" ~ "#BBDE94",
             TRUE ~ "black"
           ))

  volcano_plot <- ggplot(aldex_plot_df, aes(x = diff.win, y = diff.btw)) +
    geom_point(aes(size = size, color = famcolor)) +
    geom_text_repel(aes(label = pointlabel), max.overlaps = Inf) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_abline(slope = -1, intercept = 0, linetype = "dashed") +
    scale_color_identity() + scale_size_identity() +
    theme_cowplot() +
    labs(title = paste("ALDEx2 Volcano Plot:", filter_value),
         x = "Dispersion (diff.win)", y = "Difference (diff.btw)")

  ggsave(paste0(output_prefix, "_volcano.pdf"),
         plot = volcano_plot, height = 6, width = 6)

  #────────────────────────────────────────────
  # 5. Create relative-abundance boxplots for sig ASVs
  #────────────────────────────────────────────
  ps_rel <- transform_sample_counts(ps_filt, function(x) x/sum(x))
  melted <- psmelt(ps_rel)

  selected <- melted %>% filter(OTU %in% sig_df$ASV)

  plot_list <- list()

  for (i in seq_len(nrow(sig_df))) {
    asv <- sig_df$ASV[i]
    genus <- sig_df$Genus[i]

    taxon_data <- selected %>% filter(OTU == asv)

    p <- ggplot(taxon_data, aes_string(x = grouping_var, y = "Abundance", color = grouping_var)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, size = 1.2) +
      theme_pubr() +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 45, vjust = 0.5)) +
      labs(title = genus, subtitle = asv, y = "Relative Abundance", x = NULL)+
        theme(plot.title=element_text(hjust=0.5, face = "bold"), 
          plot.subtitle=element_text(hjust=0.5))

    plot_list[[i]] <- p
  }

  grid_plot <- plot_grid(plotlist = plot_list, ncol = 4)
  ggsave(paste0(output_prefix, "_grid.pdf"), grid_plot, height = 14, width = 14)

  #────────────────────────────────────────────
  # Return outputs as a list
  #────────────────────────────────────────────
  return(list(
    sig_asvs = sig_df,
    volcano = volcano_plot,
    grid = grid_plot
  ))
}

#test use 
filter_value = "UW-match"

results <- run_aldex2_analysis(
  ps_object = ps.filled,
  filter_value = "UW-match",
  grouping_var = "stage",
  output_prefix = "UWmatch_stage"
)



results$grid

#ggsave("./aldex2/grid_UW_match_stage_TEST.pdf", height = 15, width=15, device="pdf")




```


# UW mismatch aquatic vs adults 

AF20

```{r}

filter_value = "UW-mismatch"

results_uwmis_stage <- run_aldex2_analysis(
  ps_object = ps.filled,
  filter_value = "UW-mismatch",
  grouping_var = "stage",
  output_prefix = "UWmismatch_stage"
)


results_uwmis_stage$grid
ggsave("./aldex2/grid_UW_mismatch_stage.pdf", height = 15, width=15, device="pdf")


```


# LSTM-match by stage 

Not significant, no corresponding plot in manuscript

```{r}
filter_value = "LSTM-match"

results_LSTMmatch_stage <- run_aldex2_analysis(
  ps_object = ps.filled,
  filter_value = "LSTM-match",
  grouping_var = "stage",
  output_prefix = "LSTM-match_stage"
)

# 0 significant ASVs


```



# LSTM-mismatch by stage 

none significant no corresponding plot in manuscript

```{r}
filter_value = "LSTM-mismatch"

results_LSTMmatch_stage <- run_aldex2_analysis(
  ps_object = ps.filled,
  filter_value = "LSTM-mismatch",
  grouping_var = "stage",
  output_prefix = "LSTM-mismatch_stage"
)

# also 0 significant ASVs


```


# LSTM-match - for water vs larvae? 

Need to feed this one a pre-filtered phyloseq object with just 2 levels of desired grouping var

```{r}




run_aldex2_analysis_WL <- function(ps_object, 
                                grouping_var = "sample_type",
                                output_prefix = "aldex2_output",
                                mc.samples = 128) {

 

  #────────────────────────────────────────────
  # 1. Run ALDEx2
  #────────────────────────────────────────────
  aldex_clr <- aldex.clr(
    as.data.frame(otu_table(ps_object)),
    sample_data(ps_object)[[grouping_var]],
    mc.samples = mc.samples,
    denom = "all",
    verbose = TRUE,
    useMC = FALSE
  )

  ttest <- aldex.ttest(aldex_clr)
  effect <- aldex.effect(aldex_clr)

  aldex_df <- data.frame(ttest, effect) %>%
    rownames_to_column("ASV")

  #────────────────────────────────────────────
  # 2. Add significance classification
  #────────────────────────────────────────────
  aldex_df <- aldex_df %>%
    mutate(one_or_two = ifelse(we.eBH < 0.05 & wi.eBH < 0.05, "both", "one"))

  #────────────────────────────────────────────
  # 3. Add taxonomy data
  #────────────────────────────────────────────
  taxa_info <- data.frame(tax_table(ps_object)) %>%
    rownames_to_column("ASV")

  aldex_df <- left_join(aldex_df, taxa_info, by = "ASV")

  sig_df <- aldex_df %>% filter(one_or_two == "both")

  message("Significant ASVs: ", nrow(sig_df))
  
  # If no significant ASVs, return basic results and skip plotting
if (nrow(sig_df) == 0) {
  message("No significant ASVs detected — exiting early.")
  return(list(
    sig_asvs = sig_df,
    volcano = NULL,
    grid = NULL
  ))
}


  #────────────────────────────────────────────
  # 4. Volcano plot
  #────────────────────────────────────────────
  aldex_plot_df <- aldex_df %>%
    mutate(pointlabel = if_else(one_or_two == "both", Genus, ""),
           size = if_else(one_or_two == "both", 3, 2),
           famcolor = case_when(
             one_or_two != "both" ~ "grey",
             Class == "Gammaproteobacteria" ~ "#238b45",
             Class == "Bacteroidia" ~ "#4292c6",
             Class == "Actinomycetia" ~ "#ff7f00",
             Class == "Alphaproteobacteria" ~ "#BBDE94",
             TRUE ~ "black"
           ))

  volcano_plot <- ggplot(aldex_plot_df, aes(x = diff.win, y = diff.btw)) +
    geom_point(aes(size = size, color = famcolor)) +
    geom_text_repel(aes(label = pointlabel), max.overlaps = Inf) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_abline(slope = -1, intercept = 0, linetype = "dashed") +
    scale_color_identity() + scale_size_identity() +
    theme_cowplot() +
    labs(title = paste("ALDEx2 Volcano Plot"),
         x = "Dispersion (diff.win)", y = "Difference (diff.btw)")

  ggsave(paste0(output_prefix, "_volcano.pdf"),
         plot = volcano_plot, height = 6, width = 6)

  #────────────────────────────────────────────
  # 5. Create relative-abundance boxplots for sig ASVs
  #────────────────────────────────────────────
  ps_rel <- transform_sample_counts(ps_object, function(x) x/sum(x))
  melted <- psmelt(ps_rel)

  selected <- melted %>% filter(OTU %in% sig_df$ASV)

  plot_list <- list()

  for (i in seq_len(nrow(sig_df))) {
    asv <- sig_df$ASV[i]
    genus <- sig_df$Genus[i]

    taxon_data <- selected %>% filter(OTU == asv)

    p <- ggplot(taxon_data, aes_string(x = grouping_var, y = "Abundance", color = grouping_var)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, size = 1.2) +
      theme_pubr() +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 45, vjust = 0.5)) +
      labs(title = genus, subtitle = asv, y = "Relative Abundance", x = NULL)+
        theme(plot.title=element_text(hjust=0.5, face = "bold"), 
          plot.subtitle=element_text(hjust=0.5))

    plot_list[[i]] <- p
  }

  grid_plot <- plot_grid(plotlist = plot_list, ncol = 3)
  ggsave(paste0(output_prefix, "_grid.pdf"), grid_plot, height = 5, width = 14)

  #────────────────────────────────────────────
  # Return outputs as a list
  #────────────────────────────────────────────
  return(list(
    sig_asvs = sig_df,
    volcano = volcano_plot,
    grid = grid_plot
  ))
}




```


# LSTM-match - water vs larvae 

None signficiant 

```{r}

lstm_match_WL <- ps.filled %>% subset_samples(sample_type == "water" | sample_type == "larvae") %>% subset_samples(uni_tx == "LSTM-match")

sample_data(lstm_match_WL)$sample_type <- as.factor(sample_data(lstm_match_WL)$sample_type)
lstm_match_WL <- prune_taxa(taxa_sums(lstm_match_WL) > 0, lstm_match_WL)
sample_data(lstm_match_WL)$sample_type <- droplevels(sample_data(lstm_match_WL)$sample_type)


results_LSTMmatch_WL <- run_aldex2_analysis_WL(
  ps_object = lstm_match_WL,
  grouping_var = "sample_type",
  output_prefix = "LSTM-match_WL"
)

#0 

```


# LSTM-mismatch - water vs larvae

```{r}
lstm_mismatch_WL <- ps.filled %>% subset_samples(sample_type == "water" | sample_type == "larvae") %>% subset_samples(uni_tx == "LSTM-mismatch")

sample_data(lstm_mismatch_WL)$sample_type <- as.factor(sample_data(lstm_mismatch_WL)$sample_type)
lstm_mismatch_WL <- prune_taxa(taxa_sums(lstm_mismatch_WL) > 0, lstm_mismatch_WL)


results_LSTMmismatch_WL <- run_aldex2_analysis_WL(
  ps_object = lstm_mismatch_WL,
  grouping_var = "sample_type",
  output_prefix = "LSTM-mismatch_WL"
)

#no significant ASVs


```

# UW-match - water vs larvae 

```{r}
uw_match_WL <- ps.filled %>% subset_samples(sample_type == "water" | sample_type == "larvae") %>% subset_samples(uni_tx == "UW-match")

sample_data(uw_match_WL)$sample_type <- as.factor(sample_data(uw_match_WL)$sample_type)
uw_match_WL <- prune_taxa(taxa_sums(uw_match_WL) > 0, uw_match_WL)


results_UWmatch_WL <- run_aldex2_analysis_WL(
  ps_object = uw_match_WL,
  grouping_var = "sample_type",
  output_prefix = "UW-match_WL"
)

#3 signficant SVs - two chryseobacterium more common in water, one Leucobacter more common in water


```

#UW -mismatch - water vs larvae

```{r}

uw_mismatch_WL <- ps.filled %>% subset_samples(sample_type == "water" | sample_type == "larvae") %>% subset_samples(uni_tx == "UW-mismatch")

sample_data(uw_mismatch_WL)$sample_type <- as.factor(sample_data(uw_mismatch_WL)$sample_type)
uw_mismatch_WL <- prune_taxa(taxa_sums(uw_mismatch_WL) > 0, uw_mismatch_WL)


results_UWmismatch_WL <- run_aldex2_analysis_WL(
  ps_object = uw_mismatch_WL,
  grouping_var = "sample_type",
  output_prefix = "UW-mismatch_WL"
)

#2 sig asvs - pedobacter and microbacterceae


```






# Dev data ( no significant results )


Adding in continuous variables! A few key things to keep in mind: 

* Don't combine LSTM moz with UW moz, we know that the LSTM moz are generally bigger (mean of 2.73 vs 2.64, p = 3.4e-10 on Wlch's two-sample t-test)
* make sure the dev  question you're asking isn't already answered by another metadata value 
* Think about whether the microbial community could possibly have affected (or been affected by) the value you're looking at. Like... what does it mean to test the adult microbiome against wing length when the wing length is set at the larval stage?  

From previous: 
These ones do have "ecorr" coeffecient - if positive, bigger wings mean more Asaia or whatever development metrics on it..


```{r}

#test whether ASVs in the water can predict average wing length for that flask - UW samples



# Filter and prune water samples
ps.water_uw <- ps.filled %>%
  subset_samples(sample_type == "water" & uni == "UW") %>%
  prune_taxa(taxa_sums(.) > 0, .)


# Extract and process metadata for adult samples
md_adult_uw <- ps.filled %>%
  subset_samples(sample_type == "adult" & uni == "UW") %>%
  sample_data() %>% 
  as.matrix() %>%
  as.data.frame() %>% # Convert to data.frame before using dplyr
  filter(!is.na(wing_length)) %>% # Filter missing wing_length values
  mutate(
    sorting_var = as.factor(sorting_var),
    wing_length = as.numeric(wing_length)
  )


# Group by sorting_var and calculate summary statistics
md_grouped <- md_adult_uw %>%
  group_by(sorting_var) %>%
  summarise(
    avg_wing_length = mean(wing_length, na.rm = TRUE),
    med_wing_length = median(wing_length, na.rm = TRUE),
    sd_wing_length = sd(wing_length, na.rm = TRUE),
    n_wings = n_distinct(sample_id)
  ) %>%
  ungroup() %>% # Ungroup after summarizing
  as.data.frame()

# Extract water sample metadata and merge with grouped wing data
water_md <- sample_data(ps.water_uw) %>%
  as.matrix() %>%
  as.data.frame() %>%
  left_join(md_grouped, by = "sorting_var") %>%
  mutate(SampleID = sample_id) %>%
  column_to_rownames(var = "sample_id")

# Update sample data in the phyloseq object
sample_data(ps.water_uw) <- sample_data(water_md)







aldex2_clr_water_wing <- aldex.clr((as.data.frame(otu_table(ps.water_uw))),
                             sample_data(ps.water_uw)$avg_wing_length,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)



aldex2_water_wing_corr <- aldex.corr(aldex2_clr_water_wing, sample_data(ps.water_uw)$avg_wing_length) #runs in less than 1 minute? 



aldex2_water_wing_corr <- aldex2_water_wing_corr %>% rownames_to_column(var = "ASV")
aldex2_water_wing_corr_significant<- aldex2_water_wing_corr%>% filter(pearson.eBH<0.05)
# aldex2_pitcher_corr_significant<-aldex2_pitcher_corr_significant[,c(1:4)]


taxa_info <- data.frame(tax_table(ps.water_uw))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")

#then join to taxonomy by ASV name and generate final data frame of correlated asvs with taxonomy
aldex2_corr_asv <- merge(aldex2_water_wing_corr_significant, taxa_info, by="ASV")
#nope, nothing significant there. No taxa found in water correlated with bigger wings from that flask. 
#4cbfff144d4e7a4e0f4619ed505be070 
#This ASV had pearson.eBH = 0.063, all other were not even close 




```






# Can UW larval community composition predict adult wing length? 

```{r}


# Filter and prune water samples
ps.larvae_uw <- ps.filled %>%
  subset_samples(sample_type == "larvae" & uni == "UW") %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Extract water sample metadata and merge with grouped wing data
larv_md <- sample_data(ps.larvae_uw) %>%
  as.matrix() %>%
  as.data.frame() %>%
  left_join(md_grouped, by = "sorting_var") %>%
  mutate(SampleID = sample_id) %>%
  column_to_rownames(var = "sample_id")

# Update sample data in the phyloseq object
sample_data(ps.larvae_uw) <- sample_data(larv_md)





aldex2_clr_larv_wing <- aldex.clr((as.data.frame(otu_table(ps.larvae_uw))),
                             sample_data(ps.larvae_uw)$avg_wing_length,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)



aldex2_larv_wing_corr <- aldex.corr(aldex2_clr_larv_wing, sample_data(ps.larvae_uw)$avg_wing_length) 

aldex2_larv_wing_corr <- aldex2_larv_wing_corr %>% rownames_to_column(var = "ASV")
aldex2_larv_wing_corr_significant<- aldex2_larv_wing_corr%>% filter(pearson.eBH<0.05)
# aldex2_pitcher_corr_significant<-aldex2_pitcher_corr_significant[,c(1:4)]


taxa_info <- data.frame(tax_table(ps.larvae_uw))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")

#then join to taxonomy by ASV name and generate final data frame of correlated asvs with taxonomy
aldex2_corr_asv <- merge(aldex2_water_wing_corr_significant, taxa_info, by="ASV")
#nope, nothing significant there either. 




```

# Can LSTM water predict wing length? 


```{r}


# Filter and prune water samples
ps.water_lstm <- ps.filled %>%
  subset_samples(sample_type == "water" & uni == "LSTM") %>%
  prune_taxa(taxa_sums(.) > 0, .)


# Extract and process metadata for adult samples
md_adult_lstm <- ps.filled %>%
  subset_samples(sample_type == "adult" & uni == "LSTM") %>%
  sample_data() %>% 
  as.matrix() %>%
  as.data.frame() %>% # Convert to data.frame before using dplyr
  filter(!is.na(wing_length)) %>% # Filter missing wing_length values
  mutate(
    sorting_var = as.factor(sorting_var),
    wing_length = as.numeric(wing_length)
  )


# Group by sorting_var and calculate summary statistics
md_grouped <- md_adult_lstm %>%
  group_by(sorting_var) %>%
  summarise(
    avg_wing_length = mean(wing_length, na.rm = TRUE),
    med_wing_length = median(wing_length, na.rm = TRUE),
    sd_wing_length = sd(wing_length, na.rm = TRUE),
    n_wings = n_distinct(sample_id)
  ) %>%
  ungroup() %>% # Ungroup after summarizing
  as.data.frame()

# Extract water sample metadata and merge with grouped wing data
water_md <- sample_data(ps.water_lstm) %>%
  as.matrix() %>%
  as.data.frame() %>%
  left_join(md_grouped, by = "sorting_var") %>%
  mutate(SampleID = sample_id) %>%
  column_to_rownames(var = "sample_id")

# Update sample data in the phyloseq object
sample_data(ps.water_lstm) <- sample_data(water_md)



######################



aldex2_clr_water_wing_lstm <- aldex.clr((as.data.frame(otu_table(ps.water_lstm))),
                             sample_data(ps.water_lstm)$avg_wing_length,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)



aldex2_water_wing_corr_lstm <- aldex.corr(aldex2_clr_water_wing_lstm, sample_data(ps.water_lstm)$avg_wing_length) 
#started at 10:30



aldex2_water_wing_corr_lstm <- aldex2_water_wing_corr_lstm %>% rownames_to_column(var = "ASV")
aldex2_water_wing_corr_lstm_significant <- aldex2_water_wing_corr_lstm%>% filter(pearson.eBH<0.05)

#nope, nothing significant there either. Nothing even close
min(aldex2_water_wing_corr_lstm$pearson.eBH) #0.5642963 as lowest P value 




```

# now try for LSTM larvae ASVs vs avg adult wing length


```{r}

# Filter and prune water samples
ps.larvae_lstm <- ps.filled %>%
  subset_samples(sample_type == "larvae" & uni == "LSTM") %>%
  prune_taxa(taxa_sums(.) > 0, .)

# 
# # Extract and process metadata for adult samples -
# md_adult_lstm <- ps.filled %>%
#   subset_samples(sample_type == "adult" & uni == "lstm") %>%
#   sample_data() %>% 
#   as.matrix() %>%
#   as.data.frame() %>% # Convert to data.frame before using dplyr
#   filter(!is.na(wing_length)) %>% # Filter missing wing_length values
#   mutate(
#     sorting_var = as.factor(sorting_var),
#     wing_length = as.numeric(wing_length)
#   )
# 
# 
# # Group by sorting_var and calculate summary statistics
# md_grouped <- md_adult_lstm %>%
#   group_by(sorting_var) %>%
#   summarise(
#     avg_wing_length = mean(wing_length, na.rm = TRUE),
#     med_wing_length = median(wing_length, na.rm = TRUE),
#     sd_wing_length = sd(wing_length, na.rm = TRUE),
#     n_wings = n_distinct(sample_id)
#   ) %>%
#   ungroup() %>% # Ungroup after summarizing
#   as.data.frame()

# Extract water sample metadata and merge with grouped wing data
larv_md <- sample_data(ps.larvae_lstm) %>%
  as.matrix() %>%
  as.data.frame() %>%
  left_join(md_grouped, by = "sorting_var") %>%
  mutate(SampleID = sample_id) %>%
  column_to_rownames(var = "sample_id")

# Update sample data in the phyloseq object
sample_data(ps.larvae_lstm) <- sample_data(larv_md)




#now just merge by sorting_var because it's the exact sample for water/adults/larvae from the same flask!
#yeehaw! We did it. 
                

#need to merge by something - make anew sorting_var column in the wing_length df? 

#mismatch between number of samples and condition vector




aldex2_clr_larv_wing <- aldex.clr((as.data.frame(otu_table(ps.larvae_lstm))),
                             sample_data(ps.larvae_lstm)$avg_wing_length,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)



aldex2_larv_wing_corr <- aldex.corr(aldex2_clr_larv_wing, sample_data(ps.larvae_lstm)$avg_wing_length) 

aldex2_larv_wing_corr <- aldex2_larv_wing_corr %>% rownames_to_column(var = "ASV")
aldex2_larv_wing_corr_significant<- aldex2_larv_wing_corr%>% filter(pearson.eBH<0.05)
#none
min(aldex2_larv_wing_corr$pearson.eBH) #0.2907014 as lowest P value 

```

