---
title: "cryo_relabund"
author: "Holly Nichols"
date: "2024-01-15"
output: html_document
---

#Cryo_relabund

The goal of this analysis is to take a first look at the taxonomic abundance in the cryopreservation samples, especially across the follwoing axes: 

* timepoint (duration of cryopreservation before thaw and use)
* uni (the university where the experiment was conducted, and the source of the mozzies)
* stock_origin (the university whence the stock came)
* new_run (to see if the sequencing runs had any particular effect on the data)


This code relies heavily on the microshades package, 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq)
library(RColorBrewer)
library(ggplot2)
#remotes::install_github("KarstensLab/microshades", auth_token = "AUTH_GOES_HERE")
library(microshades) #for a color-grouped relative abunance stacked bar plot 

##this is what microshades wants me to load 
library(cowplot)
library(patchwork)
library(forcats)
library(tidyverse)
######


#import the decontam'd phyloseq object
ps.decontam_cutoff = readRDS("../data/phyloseq/crPSdecontam_cutoff_12_11_2025")
#4493 taxa and 595 samples






```

## Taxa barplots (relative abundance plots) 





### Prepare data for microshades

General outline on how to use microshades: 


* Before starting, sub-sample data as approprate and glom to your taxonomic level of interest. 
* make a "mdf_prep" object 
* Create a color object 
* extract plottable info from the color object 
* if you want, change around the color of the Phyla using reassign_colors()
* You can also add 5 more colors to any phyla - but be warned, this does not play nicely if you've also used reassign_colors()
* Create a custom legend- it may get confused if you don't specify what subgroup you want 
* Create the plot 
* plot_grid to combine your custom legend and your beautiful barplot 



# Creation of ps.filled

any "empty" genus names called unclassified_Family


```{r fill_tax_table}


ps.filled <- ps.decontam_cutoff




tax.clean <- data.frame(tax_table(ps.decontam_cutoff))



#change empty spaces in tax_names to higher name levels
#https://github.com/joey711/phyloseq/issues/990

for (i in 1:nrow(tax.clean)){
  if (tax.clean[i,2] == ""){
  kingdom <- paste("Kingdom_", tax.clean[i,1], sep = "")
  tax.clean[i, 2:7] <- kingdom
  } else if (tax.clean[i,3] == ""){
  phylum <- paste("Phylum_", tax.clean[i,2], sep = "")
  tax.clean[i, 3:7] <- phylum
  } else if (tax.clean[i,4] == ""){
  class <- paste("Class_", tax.clean[i,3], sep = "")
  tax.clean[i, 4:7] <- class
  } else if (tax.clean[i,5] == ""){
  order <- paste("Order_", tax.clean[i,4], sep = "")
  tax.clean[i, 5:7] <- order
  } else if (tax.clean[i,6] == ""){
  family <- paste("Family_", tax.clean[i,5], sep = "")
  tax.clean[i, 6:7] <- family
  } else if (tax.clean[i,7] == ""){
  tax.clean$Species[i] <- paste("Genus",tax.clean$Genus[i], sep = "_")
  }
}


tax_table(ps.filled) <- as.matrix(tax.clean) #update teh tax table in your phyloseq object 

saveRDS(ps.filled, "../data/phyloseq/crPSFilled")

ps.filled = readRDS("../data/phyloseq/crPSFilled")





#now glom 

ps.genus_filled = tax_glom(ps.filled, taxrank = 'Genus') #glom to Genus level
saveRDS(ps.genus_filled, "../data/phyloseq/crPSGenus_filled")
# 956 genera, 595 samples
# AFTER FILL - 1177 taxa!
# after 3/12/24 - 1191 taxa
#after 7/15/24 - 1177 taxa and 595 samples
#after 12/12/2025 - 1177 taxa and 595 samples 


#to load this in the future: 
ps.genus_filled = readRDS("../data/phyloseq/crPSGenus_filled")



ps.family_filled = tax_glom(ps.filled, taxrank = 'Family') #glom to Family level

saveRDS(ps.family_filled, "../data/phyloseq/crPSFamily")
ps.family_filled = readRDS("../data/phyloseq/crPSFamily")




```


# Donor pool relative abundance plots

(preview, not final fig)


```{r filled_homogenate}

ps.genus_h = subset_samples(ps.genus_filled, sample_type == "homogenate")
#78 samples - so 2 of the samples had <100 reads
ps.genus_h <- prune_taxa(taxa_sums(ps.genus_h) > 0, ps.genus_h) #trim taxa not found in this subset
#350 taxa


mdf_prep_ggh <- ps.genus_h %>%
        phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
        psmelt() %>%
        filter(Abundance > 0)



#create a color object to be used in plotting
#ok wild - just adding the _A to just the firmicutes seems to have solved the problem?? 
color_obj_ggh <- microshades::create_color_dfs(mdf_prep_ggh, group_level= "Phylum", subgroup_level= "Genus", selected_groups = c("Proteobacteria", "Actinobacteriota", "Bacteroidota", "Firmicutes_D"))




#extract
mdf_ggh <- color_obj_ggh$mdf
cdf_ggh<- color_obj_ggh$cdf

base_plot_ggh <- plot_microshades(mdf_ggh, cdf_ggh)



legend_ggh <-custom_legend(mdf_ggh, cdf_ggh)

plot_ggh<- base_plot_ggh + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none")  +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~uni, scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_blank(), axis.title.x=element_blank()) + 
  theme(plot.margin = margin(6,6,6,6)) + ggtitle("Homogenate - Greengenes - Genus - FILLED")

plot_grid(plot_ggh, legend_ggh,rel_widths = c(1, .3))

#ggsave("./plots/homogenate_barplot_gg.pdf", height=5, width=9, device="pdf") 

```






## Generate a function to plot w/ microshades


```{r}

#define a function to plot various phyloseq objects

make_microshades_plot <- function(ps,
                                  facet_var,
                                  nrow = 2,
                                  title = "Microshades Plot") {
  # prepare melted data
  mdf_prep <- ps %>%
    phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
    psmelt() %>%
    filter(Abundance > 0)
  
  # create color object
  color_obj <- microshades::create_color_dfs(
    mdf_prep,
    group_level = "Phylum",
    subgroup_level = "Genus",
    selected_groups = c("Proteobacteria", 
                        "Actinobacteriota",
                        "Bacteroidota",
                        "Firmicutes_D")
  )
  
  # extract dfs
  mdf <- color_obj$mdf
  cdf <- color_obj$cdf
  
 cdf <- color_reassign(cdf,
                           group_assignment = c("Proteobacteria", "Actinobacteriota","Bacteroidota","Firmicutes_D"),
                           color_assignment = c("micro_green", "micro_orange","micro_blue","micro_brown"))


  new_groups <- extend_group(mdf, cdf, "Phylum", "Genus", "Proteobacteria", existing_palette = "micro_green", new_palette = "micro_purple", n_add = 5)
  #add 5 more slots to proteo
  
  
  # base plot and legend
  base_plot <- plot_microshades(new_groups$mdf, new_groups$cdf)
  legend <- custom_legend(new_groups$mdf, new_groups$cdf)
  
  # build final plot
  plot <- base_plot +
    scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
    theme(legend.position = "none",
          axis.text.x = element_text(size = 6),
          plot.margin = margin(6,6,6,6),
    # REMOVE GREY BACKGROUND!!
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank()
    ) +
    facet_wrap(as.formula(paste("~", facet_var)),
               scales = "free_x",
               nrow = nrow) +
    ggtitle(title)
  
  # combine plot + legend
  plot_grid(plot, legend, rel_widths = c(1, .7))
}



```







# All individual mozzies, include donor pools (Fig 2)

Raw material for Fig 2, rearranged for clarity in Inkscape


extended proteos - 10 colors instead of 5
Also may subset to just unfrozen + 6 months 
```{r}

ps.genus_filled

########################### WORKING HERE 

ps.genus_a_h <- subset_samples(ps.genus_filled, sample_type != "water" & sample_type != "larvae")

plot_final <- make_microshades_plot(
  ps = ps.genus_a_h, 
  facet_var = "~interaction(timepoint, uni, tx)",
  nrow = 4,
  title = "All adults "
)

plot_final 
#ggsave("./plots/SOVE/barplot_individuals_extendedproteo_donor_adults_xl.pdf", height=25, width=45, device="pdf") 




```





# Sample Type Story (suppemental figs 21 and 24)

## Water samples (preview, not in manuscript)


I'm interested in differences by timepoint and by uni_tx. 

Particularly - do the actinos (esp. leucobacter) just not wake up after freeze (LSTM?) How do match vs mismatch water samples look? 



```{r}

ps.genus_water <- subset_samples(ps.genus_filled, sample_type == "water")


mdf_prep_ggh <- ps.genus_water %>%
        phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
        psmelt() %>%
        filter(Abundance > 0)

color_obj_ggh <- microshades::create_color_dfs(mdf_prep_ggh, group_level= "Phylum", subgroup_level= "Genus", selected_groups = c("Proteobacteria", "Actinobacteriota", "Bacteroidota", "Firmicutes_D"))




#extract
mdf_ggh <- color_obj_ggh$mdf
cdf_ggh<- color_obj_ggh$cdf

base_plot_ggh <- plot_microshades(mdf_ggh, cdf_ggh)



legend_ggh <-custom_legend(mdf_ggh, cdf_ggh)

plot_ggh<- base_plot_ggh + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none")  +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~interaction(timepoint, uni_tx), scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_blank(), axis.title.x=element_blank()) + 
  theme(plot.margin = margin(6,6,6,6)) + ggtitle(" Water samples - Greengenes - Genus - FILLED")

plot_grid(plot_ggh, legend_ggh,rel_widths = c(1, .3))
#ggsave("./plots/water_unitx_tp_barplot_gg.pdf", height=5, width=9, device="pdf") 


```



## Larvae Samples (preview, not in manuscript)

These mirror their water samples pretty closely. Note that we're missing larval samples from the UW side, these were lost somehow during the processing/library prep phase. 

```{r}


ps.genus_larvae <- subset_samples(ps.genus_filled, sample_type == "larvae")

mdf_prep_ggh <- ps.genus_larvae %>%
        phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
        psmelt() %>%
        filter(Abundance > 0)

color_obj_ggh <- microshades::create_color_dfs(mdf_prep_ggh, group_level= "Phylum", subgroup_level= "Genus", selected_groups = c("Proteobacteria", "Actinobacteriota", "Bacteroidota", "Firmicutes_D"))




#extract
mdf_ggh <- color_obj_ggh$mdf
cdf_ggh<- color_obj_ggh$cdf

base_plot_ggh <- plot_microshades(mdf_ggh, cdf_ggh)



legend_ggh <-custom_legend(mdf_ggh, cdf_ggh)

plot_ggh<- base_plot_ggh + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none")  +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~interaction(timepoint, uni_tx), scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
        # REMOVE GREY BACKGROUND!!
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank()
    )  + 
  
  theme(plot.margin = margin(6,6,6,6)) + ggtitle(" Larvae samples - Greengenes - Genus - FILLED")

plot_grid(plot_ggh, legend_ggh,rel_widths = c(1, .3))
#ggsave("./plots/larvae_unitx_tp_barplot_gg.pdf", height=5, width=9, device="pdf") 

```


# Within UW, compare larvae to water

Additional File 21

```{r}
#subset to just UW larvae and water 


ps.genus_uw_compare <- ps.genus_filled %>% subset_samples(uni == "UW") %>%
                                           subset_samples(sample_type == "water" | sample_type == "larvae") 

mdf_prep_ggh <- ps.genus_uw_compare %>%
        phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
        psmelt() %>%
        filter(Abundance > 0)


mdf_prep_ggh$sample_type <- factor(mdf_prep_ggh$sample_type, levels = c("water", "larvae"))

color_obj_ggh <- microshades::create_color_dfs(mdf_prep_ggh, group_level= "Phylum", subgroup_level= "Genus", selected_groups = c("Proteobacteria", "Actinobacteriota", "Bacteroidota", "Firmicutes_D"))




#extract
mdf_ggh <- color_obj_ggh$mdf
cdf_ggh<- color_obj_ggh$cdf

base_plot_ggh <- plot_microshades(mdf_ggh, cdf_ggh)



legend_ggh <-custom_legend(mdf_ggh, cdf_ggh)


#need to reorder sample_type to be water then larvae 


plot_ggh<- base_plot_ggh + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none")  +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~interaction(timepoint, uni_tx, sample_type), scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
          # REMOVE GREY BACKGROUND!!
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank()) + 
  theme(plot.margin = margin(6,6,6,6)) + ggtitle(" UW Larvae and water samples - Greengenes - Genus - FILLED")


#will need to rearrange plot in inkscape because it wants to put larve first 

plot_grid(plot_ggh, legend_ggh,rel_widths = c(1, .3))
#ggsave("./plots/barplot_uw_larvae_v_water.pdf", height=5, width=9, device="pdf") 


```

#Within LSTM, compare larvae to water

Additional File 24

```{r}

#subset to just LSTM larvae and water 


ps.genus_lstm_compare <- ps.genus_filled %>% subset_samples(uni == "LSTM") %>%
                                           subset_samples(sample_type == "water" | sample_type == "larvae") 

mdf_prep_ggh <- ps.genus_lstm_compare %>%
        phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
        psmelt() %>%
        filter(Abundance > 0)


mdf_prep_ggh$sample_type <- factor(mdf_prep_ggh$sample_type, levels = c("water", "larvae"))

color_obj_ggh <- microshades::create_color_dfs(mdf_prep_ggh, group_level= "Phylum", subgroup_level= "Genus", selected_groups = c("Proteobacteria", "Actinobacteriota", "Bacteroidota", "Firmicutes_D"))




#extract
mdf_ggh <- color_obj_ggh$mdf
cdf_ggh<- color_obj_ggh$cdf

base_plot_ggh <- plot_microshades(mdf_ggh, cdf_ggh)



legend_ggh <-custom_legend(mdf_ggh, cdf_ggh)


#need to reorder sample_type to be water then larvae 


plot_ggh<- base_plot_ggh + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none")  +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~interaction(timepoint, uni_tx, sample_type), scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_blank(), axis.title.x=element_blank(),
          # REMOVE GREY BACKGROUND!!
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank()) + 
  theme(plot.margin = margin(6,6,6,6)) + ggtitle(" LSTM Larvae and water samples - Greengenes - Genus - FILLED")

#get it to lavel with flask ID aka sorting_var

head(base_plot_ggh)

#will need to rearrange plot in inkscape because it wants to put larve first 

plot_grid(plot_ggh, legend_ggh,rel_widths = c(1, .3))
#ggsave("./plots/barplot_lstm_larvae_v_water.pdf", height=5, width=9, device="pdf") 

```







