---
title: "cryo_corrplots"
author: "Holly Nichols"
date: "2025-02-25"
output: html_document
---

**Goal:** Determine whether there are any interactions that hold up within UW samples but not LSTM samples and vice versa 

Elizabethkingia goals: 

* Are there any ASVs correlated with Elizabethkingia relative abundance? Why might it dominate in some samples/timepoints but not others? 
* Can Elizabethkingia abundance in adults be predicted by corresponding water/larvae samples? 


```{r setup, include=FALSE}


library(phyloseq)
library(tidyverse)
library(corrplot)
library(rstatix) #for stats on corrplot 

ps.filled = readRDS("../data/phyloseq/crPSFilled")
#4493 taxa and 595 samples 



```

#Spearman correlation



Def include across multiple timepoints, we want to see trends robust to multiple thaw events

UW-match adults (most interested in Elizabethkingia in this one)
UW-mismatch adults 
LSTM-match adults 
LSTM-mismatch adults 



Try it first with just 25 of the most abundant ASVs


#UW-match 
AF8

```{r}

#generate phyloseq object of interest
ps.uw_match_adults <- subset_samples(ps.filled, sample_type == "adult" & uni_tx == "UW-match")





ps.uw_match_adults_norm  = transform_sample_counts(ps.uw_match_adults, function(x) x / sum(x) )

#now separate out the top 25 ASVs in this dataset 
read_counts <- ps.uw_match_adults %>% 
    taxa_sums() %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    set_names(c("ASV", "reads"))

read_counts <- read_counts[order(read_counts$reads, decreasing = TRUE),]


top_25 <- slice(read_counts, 1:25)
top_25_asvs <- top_25$ASV   

#ok, so let's subset our physeq obj first, then unlist

ps.top_25_uw_match_adults<- prune_taxa(top_25_asvs, ps.uw_match_adults_norm) #keep only the ASVs in top_25_mother object 


#corr wants just the OTU table - which may need to be transposed

com = ps.top_25_uw_match_adults %>% otu_table() %>% t() 


cc_uwma = cor(com, method = "spearman")  



#have it redo the colname and rownames (default is just ASV labels)
taxtab <- ps.uw_match_adults %>% tax_table() %>% as.data.frame %>% rownames_to_column(var = "ASV") #harvest tax info
merged <- left_join(top_25,taxtab, by = "ASV") #now join tax info to your df of ASVs kept 



#now need to rearrange genus by colnames(cc)
cc_names <- colnames(cc_uwma)

merged <- merged %>%
  mutate(ASV = factor(ASV, levels = cc_names)) %>%  # Set ASV order based on cc_names
  arrange(ASV)  # Sort the dataframe

#plot with ASV labels
corrplot(cc_uwma, tl.col = "red", order = "hclust", hclust.method = "average", 
         addrect = 4, 
         tl.cex = 0.6, #ASV text size 
         title = "UW - match - adults")



#plot with Genus labels, highlight comamonas and include part of ASV label 

merged_temp  <- merged %>% 
  mutate(genus_com = case_when(
    Genus == "Comamonas_F_589250" ~ paste0(Genus, "____*", substr(ASV, start = 1, stop = 5), "*"),
    TRUE ~ Genus
  ))
   



colnames(cc_uwma) <- merged_temp$genus_com
rownames(cc_uwma) <- merged_temp$genus_com

#plot with Genus labels


# pdf("./plots/corrplots/corr_uw_match_adults.pdf")
# corrplot(cc_uwma, tl.col = "black", order = "hclust", hclust.method = "average", 
#          addrect = 4, 
#          tl.cex = 0.8, #ASV text size 
#          title = "UW - match adults")
# dev.off()



```


#UW - Mismatch - Adults 
AF8

```{r}

#generate phyloseq object of interest
ps.uw_mismatch_adults <- subset_samples(ps.filled, sample_type == "adult" & uni_tx == "UW-mismatch")



ps.uw_mismatch_adults_norm  = transform_sample_counts(ps.uw_mismatch_adults, function(x) x / sum(x) )

#now separate out the top 25 ASVs in this dataset 
read_counts <- ps.uw_mismatch_adults %>% 
    taxa_sums() %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    set_names(c("ASV", "reads"))

read_counts <- read_counts[order(read_counts$reads, decreasing = TRUE),]


top_25 <- slice(read_counts, 1:25)
top_25_asvs <- top_25$ASV   

#ok, so let's subset our physeq obj first, then unlist

ps.top_25_uw_mismatch_adults<- prune_taxa(top_25_asvs, ps.uw_mismatch_adults_norm) #keep only the ASVs in top_25_mother object 


#corr wants just the OTU table - which may need to be transposed

com = ps.top_25_uw_mismatch_adults %>% otu_table() %>% t() 


cc_uw_mismatch = cor(com, method = "spearman")  

#have it redo the colname and rownames (default is just ASV labels)
taxtab <- ps.uw_mismatch_adults %>% tax_table() %>% as.data.frame %>% rownames_to_column(var = "ASV") #harvest tax info
merged <- left_join(top_25,taxtab, by = "ASV") #now join tax info to your df of ASVs kept 



#now need to rearrange genus by colnames(cc)
cc_names <- colnames(cc_uw_mismatch)

merged <- merged %>%
  mutate(ASV = factor(ASV, levels = cc_names)) %>%  # Set ASV order based on cc_names
  arrange(ASV)  # Sort the dataframe

#plot with ASV labels
corrplot(cc_uw_mismatch, tl.col = "red", order = "hclust", hclust.method = "average", 
         addrect = 4, 
         tl.cex = 0.8, #ASV text size 
         title = "uw_mismatch - adults")





#plot with Genus labels, highlight comamonas and include part of ASV label 

merged_temp  <- merged %>% 
  mutate(genus_com = case_when(
    Genus == "Comamonas_F_589250" ~ paste0(Genus, "____*", substr(ASV, start = 1, stop = 5), "*"),
    TRUE ~ Genus
  ))
   



colnames(cc_uw_mismatch) <- merged_temp$genus_com
rownames(cc_uw_mismatch) <- merged_temp$genus_com



# 
# pdf("./plots/corrplots/corr_uw_mismatch_adults.pdf")
# corrplot(cc_uw_mismatch, tl.col = "black", order = "hclust", hclust.method = "average", 
#          addrect = 4, 
#          tl.cex = 0.8, #ASV text size 
#          title = "UW-mismatch adults")
# 
# dev.off()




```




# LSTM-match adults 
AF7

```{r}


#generate phyloseq object of interest
ps.lstm_match_adults <- subset_samples(ps.filled, sample_type == "adult" & uni_tx == "LSTM-match")



ps.lstm_match_adults_norm  = transform_sample_counts(ps.lstm_match_adults, function(x) x / sum(x) )

#now separate out the top 25 ASVs in this dataset 
read_counts <- ps.lstm_match_adults %>% 
    taxa_sums() %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    set_names(c("ASV", "reads"))

read_counts <- read_counts[order(read_counts$reads, decreasing = TRUE),]


top_25 <- slice(read_counts, 1:25)
top_25_asvs <- top_25$ASV   

#ok, so let's subset our physeq obj first, then unlist

ps.top_25_lstm_match_adults<- prune_taxa(top_25_asvs, ps.lstm_match_adults_norm) #keep only the ASVs in top_25_mother object 


#corr wants just the OTU table - which may need to be transposed

com = ps.top_25_lstm_match_adults %>% otu_table() %>% t() 


cc_lstm_match = cor(com, method = "spearman")  

#have it redo the colname and rownames (default is just ASV labels)
taxtab <- ps.lstm_match_adults %>% tax_table() %>% as.data.frame %>% rownames_to_column(var = "ASV") #harvest tax info
merged <- left_join(top_25,taxtab, by = "ASV") #now join tax info to your df of ASVs kept 



#now need to rearrange genus by colnames(cc)
cc_names <- colnames(cc_lstm_match)

merged <- merged %>%
  mutate(ASV = factor(ASV, levels = cc_names)) %>%  # Set ASV order based on cc_names
  arrange(ASV)  # Sort the dataframe

#plot with ASV labels
corrplot(cc_lstm_match, tl.col = "red", order = "hclust", hclust.method = "average", 
         addrect = 4, 
         tl.cex = 0.6, #ASV text size 
         title = "LSTM_match - adults")





#plot with Genus labels, highlight comamonas and include part of ASV label 

merged_temp  <- merged %>% 
  mutate(genus_com = case_when(
    Genus == "Comamonas_F_589250" ~ paste0(Genus, "____*", substr(ASV, start = 1, stop = 5), "*"),
    TRUE ~ Genus
  ))
   



colnames(cc_lstm_match) <- merged_temp$genus_com
rownames(cc_lstm_match) <- merged_temp$genus_com



# pdf("./plots/corrplots/corr_lstm_match_adults.pdf")
# corrplot(cc_lstm_match, tl.col = "black", order = "hclust", hclust.method = "average", 
#          addrect = 4, 
#          tl.cex = 0.8, #ASV text size 
#          title = "LSTM - match adults")
# 
# dev.off()


```

#LSTM - mismatch - adults 

AF7
Main question - is this different in some way from LSTM-match adults? 
Super cool! Really similar resuts in terms of things that are antagonistsic with Elizabethkingia 

```{r}


#generate phyloseq object of interest
ps.lstm_mismatch_adults <- subset_samples(ps.filled, sample_type == "adult" & uni_tx == "LSTM-mismatch")



ps.lstm_mismatch_adults_norm  = transform_sample_counts(ps.lstm_mismatch_adults, function(x) x / sum(x) )

#now separate out the top 25 ASVs in this dataset 
read_counts <- ps.lstm_mismatch_adults %>% 
    taxa_sums() %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    set_names(c("ASV", "reads"))

read_counts <- read_counts[order(read_counts$reads, decreasing = TRUE),]


top_25 <- slice(read_counts, 1:25)
top_25_asvs <- top_25$ASV   

#ok, so let's subset our physeq obj first, then unlist

ps.top_25_lstm_mismatch_adults<- prune_taxa(top_25_asvs, ps.lstm_mismatch_adults_norm) #keep only the ASVs in top_25_mother object 


#corr wants just the OTU table - which may need to be transposed

com = ps.top_25_lstm_mismatch_adults %>% otu_table() %>% t() 


cc_lstm_mismatch = cor(com, method = "spearman")  


#have it redo the colname and rownames (default is just ASV labels)
taxtab <- ps.lstm_mismatch_adults %>% tax_table() %>% as.data.frame %>% rownames_to_column(var = "ASV") #harvest tax info
merged <- left_join(top_25,taxtab, by = "ASV") #now join tax info to your df of ASVs kept 



#now need to rearrange genus by colnames(cc)
cc_names <- colnames(cc_lstm_mismatch)

merged <- merged %>%
  mutate(ASV = factor(ASV, levels = cc_names)) %>%  # Set ASV order based on cc_names
  arrange(ASV)  # Sort the dataframe

#plot with ASV labels
corrplot(cc_lstm_mismatch, tl.col = "red", order = "hclust", hclust.method = "average", 
         addrect = 4, 
         tl.cex = 0.6, #ASV text size 
         title = "lstm_mismatch - adults")

corrplot(cc_lstm_mismatch, tl.col = "black", order = "hclust", hclust.method = "average", 
         addrect = 4, 
         tl.cex = 0.6, #ASV text size 
         title = "LSTM-mismatch adults")


#ok let's actually break in and change just teh Comomonas labels - add __ (the first 5 characters of the ASV label )

merged_temp  <- merged %>% 
  mutate(genus_com = case_when(
    Genus == "Comamonas_F_589250" ~ paste0(Genus, "____*", substr(ASV, start = 1, stop = 5), "*"),
    TRUE ~ Genus
  ))
   



colnames(cc_lstm_mismatch) <- merged_temp$genus_com
rownames(cc_lstm_mismatch) <- merged_temp$genus_com

#plot with Genus labels


# 
# pdf("./plots/corrplots/corr_lstm_mismatch_adults.pdf")
# corrplot(cc_lstm_mismatch, tl.col = "black", order = "hclust", hclust.method = "average", 
#          addrect = 4, 
#          tl.cex = 0.8, #ASV text size 
#          title = "LSTM-mismatch adults")
# 
# dev.off()







```


#Hypothesis testing 


AF9

Simplify - Elizabethkingia vs Comamonas in LSTM-match, LSTM-mismatch, UW-match, UW-mismatch 

Then use BH correction for multiple comparisons 


```{r}

#get vectors for relabund of dif ASVs across each sample 


relabund_matrix <- ps.top_25_lstm_mismatch_adults %>% otu_table() %>% t() %>%data.frame()
head(relabund_matrix)

colnames(relabund_matrix)

#Elizabethkingia ASV name 
#7e18a39dd964eb207545655ebe71a459

# Antagonistic Comamonas ASV name 
# d32e579b3ae7b2aae8d5bf9f027c29af

#oo, also just plot them as scatterplots with lines and see how it looks for each uni_tx 


eliz <- relabund_matrix$X7e18a39dd964eb207545655ebe71a459
comamonas <- relabund_matrix$d32e579b3ae7b2aae8d5bf9f027c29af



cor.test(eliz,comamonas, method = "spearm")
#p value=  1.412e-09
#rho = 0.616915

```


# ELizabethkingia and comamonas_d32e5 (synergystic) in uni_tx adults

```{r}

#define a function to correlate elizabethkingnia and comamonas ASVs within each uni_tx 
correlate_eliz_com <- function(ps) {
  #pull out relevant 
  eliz <- ps %>% otu_table() %>% t() %>% data.frame() %>% .$X7e18a39dd964eb207545655ebe71a459
  comamonas <- ps %>% otu_table() %>% t() %>% data.frame() %>% .$d32e579b3ae7b2aae8d5bf9f027c29af

  #test correlation
  corrtest <- cor.test(eliz,comamonas, method = "spearm", exact = FALSE)
  
  p_value <- corrtest$p.value
  rho_estimate <- corrtest$estimate
  
  #pull out the p value and spearman rho 
  df <- data.frame(rho_estimate, p_value, row.names  =  unique(sample_data(ps)$uni_tx))    
  return(df)
  #return a df 
}
  

df_lstm_match <- correlate_eliz_com(ps.top_25_lstm_match_adults)
df_lstm_mismatch <- correlate_eliz_com(ps.top_25_lstm_mismatch_adults)
#df_uw_match <- correlate_eliz_com(ps.top_25_uw_match_adults) #insufficient elizabethkingia reads to correlate
df_uw_mismatch <- correlate_eliz_com(ps.top_25_uw_mismatch_adults)

big_df <- dplyr::bind_rows(df_lstm_match, df_lstm_mismatch, df_uw_mismatch)
#cool! Currenlty unadjusted 

big_df$p_adj_BH <- p.adjust(big_df$p_value, method ="BH")
big_df$p_adj_bonferroni <- p.adjust(big_df$p_value, method ="bonferroni")

#conclusion - LSTM both VERY significant, , US-mismatch is noT significant 




```


## And how abundant are those Comamonas in each uni_tx? 



##  Manuscript verison of relative abundance plots, Additional File 11 
eliz - 7e18a39dd964eb207545655ebe71a459
positive corr - d32e579b3ae7b2aae8d5bf9f027c29af
negative corr - 16da9716b4d7c3c9c904f3f799d67dfe

```{r}
ps.relabund <- transform_sample_counts(ps.filled, function(x) { x/sum(x) })
melt_relabund <- psmelt(ps.relabund) #Commented out b/c it takes forever to run 



dfEliz <- filter(melt_relabund, OTU == "7e18a39dd964eb207545655ebe71a459") %>% filter(sample_type == "adult")



StackedBarPlot_eliz<- dfEliz %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity", fill = "blue") +
  labs(x = "",y = "Relative Abundance",title = "Elizabethkingia Relative Abundance")+ 
   facet_wrap(~interaction(uni_tx,sample_type), scales = "free_x", ncol = 4) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))+# move x-axis up so it's nearly touching bars but not obscuring 
    coord_cartesian(ylim=c(0,1))
StackedBarPlot_eliz

#ggsave("./plots/unitxCORR_ELIZ.pdf", height=5, width=6, device="pdf") 


dfComAnt <- filter(melt_relabund, OTU == "16da9716b4d7c3c9c904f3f799d67dfe") %>% filter(sample_type == "adult")



StackedBarPlot_ant <- dfComAnt %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity", fill = "blue") +
  labs(x = "",y = "Relative Abundance",title = "Comamonas_16da Relative Abundance")+ 
   facet_wrap(~interaction(uni_tx,sample_type), scales = "free_x", ncol = 4) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))+# move x-axis up so it's nearly touching bars but not obscuring 
    coord_cartesian(ylim=c(0,1))
StackedBarPlot_ant

#ggsave("./plots/unitxCORR_Comamonas_ant_16da9.pdf", height=5, width=6, device="pdf") 

#interesting, the antagonist Comamonas is present in kinda all the adult types, but is only big in the larvae from UW-match, whcih is where it never showed up in adults. 


dfComSyn <- filter(melt_relabund, OTU == "d32e579b3ae7b2aae8d5bf9f027c29af") %>% filter(sample_type == "adult")


StackedBarPlot_syn <- dfComSyn %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity", fill = "blue") +
  labs(x = "",y = "Relative Abundance",title = "Comamonas_d32e Relative Abundance")+ 
   facet_wrap(~interaction(uni_tx,sample_type), scales = "free_x", ncol = 4) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
    scale_y_continuous(expand = c(0,0.001))+# move x-axis up so it's nearly touching bars but not obscuring 
    coord_cartesian(ylim=c(0,1))

StackedBarPlot_syn
#ggsave("./plots/unitxCORR_Comamonas_syn_d32d5.pdf", height=5, width=6, device="pdf") 


ggarrange(StackedBarPlot_eliz,StackedBarPlot_syn,StackedBarPlot_ant, ncol = 1)
#ggsave("./plots/unitxCORR_relabunds.pdf", height=7, width=10, device="pdf") 



```


#test Elizabethkingia/Comamonas_d32e5 corr in Water

AF 9 


```{r}

#get ps objects 
ps.lstm_match_water <- subset_samples(ps.filled, sample_type == "water" & uni_tx == "LSTM-match")
ps.lstm_mismatch_water <- subset_samples(ps.filled, sample_type == "water" & uni_tx == "LSTM-mismatch")
ps.uw_match_water <- subset_samples(ps.filled, sample_type == "water" & uni_tx == "UW-match")
ps.uw_mismatch_water <- subset_samples(ps.filled, sample_type == "water" & uni_tx == "UW-mismatch")





df_lstm_match <- correlate_eliz_com(ps.lstm_match_water)
df_lstm_mismatch <- correlate_eliz_com(ps.lstm_mismatch_water)
#df_uw_match <- correlate_eliz_com(ps.uw_match_water) #insufficient reads to correlate
df_uw_mismatch <- correlate_eliz_com(ps.uw_mismatch_water)

water_df <- dplyr::bind_rows(df_lstm_match, df_lstm_mismatch, df_uw_mismatch)

water_df$p_adj_BH <- p.adjust(water_df$p_value, method ="BH")
water_df$p_adj_bonferroni <- p.adjust(water_df$p_value, method ="bonferroni")

#nothing significant here 

```


#larvae Elizabethkingia vs comamonas_d32e5 

AF 9 

```{r}
ps.lstm_match_larvae <- subset_samples(ps.filled, sample_type == "larvae" & uni_tx == "LSTM-match")
ps.lstm_mismatch_larvae <- subset_samples(ps.filled, sample_type == "larvae" & uni_tx == "LSTM-mismatch")
ps.uw_match_larvae <- subset_samples(ps.filled, sample_type == "larvae" & uni_tx == "UW-match")
ps.uw_mismatch_larvae <- subset_samples(ps.filled, sample_type == "larvae" & uni_tx == "UW-mismatch")





df_lstm_match <- correlate_eliz_com(ps.lstm_match_larvae)
df_lstm_mismatch <- correlate_eliz_com(ps.lstm_mismatch_larvae)
#df_uw_match <- correlate_eliz_com(ps.uw_match_larvae) #insuficcent reads to correlate 
df_uw_mismatch <- correlate_eliz_com(ps.uw_mismatch_larvae)

larvae_df <- dplyr::bind_rows(df_lstm_match, df_lstm_mismatch, df_uw_mismatch)
#cool! Currenlty unadjusted 

larvae_df$p_adj_BH <- p.adjust(larvae_df$p_value, method ="BH")
larvae_df$p_adj_bonferroni <- p.adjust(larvae_df$p_value, method ="bonferroni")
#only UW-mismatch is significant here - and it's a positive correlation! 

```



#Correlate the putative antagonist! comamonas_16da9

AF 10 

```{r}


#elizabethkingia 
#X7e18a39dd964eb207545655ebe71a459
#Comamonas, antagonist 
#16da9716b4d7c3c9c904f3f799d67dfe


correlate_eliz_com_2 <- function(ps) {
  #pull out relevant 
  eliz <- ps %>% otu_table() %>% t() %>% data.frame() %>% .$X7e18a39dd964eb207545655ebe71a459
  comamonas <- ps %>% otu_table() %>% t() %>% data.frame() %>% .$X16da9716b4d7c3c9c904f3f799d67dfe

  #test correlation
  corrtest <- cor.test(eliz,comamonas, method = "spearm", exact = FALSE)
  
  p_value <- corrtest$p.value
  rho_estimate <- corrtest$estimate
  
  #pull out the p value and spearman rho 
  df <- data.frame(rho_estimate, p_value, row.names  =  unique(sample_data(ps)$uni_tx))    
  return(df)
  #return a df 
}
  



df_lstm_match <- correlate_eliz_com_2(ps.top_25_lstm_match_adults)
df_lstm_mismatch <- correlate_eliz_com_2(ps.top_25_lstm_mismatch_adults)
#df_uw_match <- correlate_eliz_com_2(ps.top_25_uw_match_adults)
df_uw_mismatch <- correlate_eliz_com_2(ps.top_25_uw_mismatch_adults)

big_df <- dplyr::bind_rows(df_lstm_match, df_lstm_mismatch, df_uw_mismatch)
#cool! Currenlty unadjusted 

big_df$p_adj_BH <- p.adjust(big_df$p_value, method ="BH")
big_df$p_adj_bonferroni <- p.adjust(big_df$p_value, method ="bonferroni")

#neg corr in LSTM-match adn LSTM-mismatch, UW-mismatch p = 1




```


# Larvae - Comamonas antagonist 16da97

AF 10 

```{r}

df_lstm_match <- correlate_eliz_com_2(ps.lstm_match_larvae)
df_lstm_mismatch <- correlate_eliz_com_2(ps.lstm_mismatch_larvae)
#df_uw_match <- correlate_eliz_com_2(ps.uw_match_larvae) #not enough eliz
#df_uw_mismatch <- correlate_eliz_com_2(ps.uw_mismatch_larvae) #not enough comamonas

larvae_df <- dplyr::bind_rows(df_lstm_match, df_lstm_mismatch)
#cool! Currenlty unadjusted 

larvae_df$p_adj_BH <- p.adjust(larvae_df$p_value, method ="BH")
larvae_df$p_adj_bonferroni <- p.adjust(larvae_df$p_value, method ="bonferroni")

#NOT significant 
```

# Water - Comamonas antagonist 16da97

AF 10 

```{r}


df_lstm_match <- correlate_eliz_com_2(ps.lstm_match_water)#stdev 0 
df_lstm_mismatch <- correlate_eliz_com_2(ps.lstm_mismatch_water)#stdev is - 
#df_uw_match <- correlate_eliz_com_2(ps.uw_match_water) #
df_uw_mismatch <- correlate_eliz_com_2(ps.uw_mismatch_water) #stdev is 0 on this one 

water_df <- dplyr::bind_rows(df_lstm_match, df_lstm_mismatch, df_uw_mismatch)


water_df$p_adj_BH <- p.adjust(water_df$p_value, method ="BH")
water_df$p_adj_bonferroni <- p.adjust(water_df$p_value, method ="bonferroni")

#huh- NA for all, not enough comaonas in water


```


