---
title: "cryo_biotyping"
author: "Holly Nichols"
date: "2025-12-15"
output: html_document
---

# Biotyping

**Goal**: Elizabethkingia seems to really dominate samples where it's found. But at times it feels all-or-nothing, especially in LSTM adult samples. Do these Elizabethkingia-heavy samples cluster separately from other? If so,which other taxa are found in the Elizabethkignia-heavy group?  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(devtools)
#install_github("tapj/BiotypeR")
library(BiotypeR)


#install.packages("clusterSim")
library("clusterSim")
# library(clusterSim)
library(cluster)
library(ade4)
library(BiotypeR)
library(phyloseq)
library(tibble)

library(qiime2R)
library(microshades)
library(ALDEx2)
library(ggrepel) #for alsdex2 "volcano" plots 

ps.filled = readRDS("../data/phyloseq/crPSFilled")

```



# First, plot an oridination, but color each sample by whether it has lots of Eliz or nah 

```{r}


#read in RPCA ordination
rpca_ah <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/rpca_ordination.qza")
rpca_ah_axes <- rpca_ah$data$ProportionExplained
rpca_ah <- rpca_ah$data$Vectors %>% dplyr::select(SampleID,PC1,PC2)
rpca_ah <- rpca_ah %>% mutate(sample_id = SampleID)

str(rpca_ah)


#classify each sample as to whether it has > 50% elizabethkingia - might as well use that one specific ASV! 

#get relabund 
ps.relabund <- transform_sample_counts(ps.filled, function(x) { x/sum(x) })
ps.eliz <- prune_taxa(
  taxa_names(ps.relabund) == "7e18a39dd964eb207545655ebe71a459",
  ps.relabund
)


#make a new column in ps.filled to show elizabethkingia abundance 
half_eliz <- sample_sums(ps.eliz) %>% data.frame() %>% rownames_to_column()
colnames(half_eliz) <- c("sample_id","perc_eliz") 


sample_data(ps.filled)$perc_eliz <-
  half_eliz$perc_eliz[
    match(sample_names(ps.filled), half_eliz$sample_id)
  ]



sd <- sample_data(ps.filled)

#halves
sd$eliz_bin <- cut(
  sd$perc_eliz,
  breaks = c(0, 1/2, 1),
  include.lowest = TRUE,
  labels = c("<1/2", ">1/2")
)
# 
# #thirds
# sd$eliz_bin <- cut(
#   sd$perc_eliz,
#   breaks = c(0, 1/3, 2/3, 1),
#   include.lowest = TRUE,
#   labels = c("<1/3", "1/3–2/3", "2/3–1")
# )

#quarters
# sd$eliz_bin <- cut(
#   sd$perc_eliz,
#   breaks = c(0, 1/4, 1/2, 3/4, 1),
#   include.lowest = TRUE,
#   labels = c("<1/4", "1/4–1/2", "1/2–3/4", "3/4-1")
# )
# 



sample_data(ps.filled) <- sd

met <- sample_data(ps.filled) %>% as.matrix() %>% data.frame() %>% filter(sample_type == "adult" | sample_type == "homogenate")

met <- met %>% 
  mutate(SampleID = sample_id) #fix sample_id vs SampleID differene before merging
#metadata2 <- subset(metadat, select = -c(sample_id) ) #not sure what this should do...? 


#merge metadata and ordination
rpca_md <- merge(met, rpca_ah, by = "SampleID")
head(rpca_md)





ggplot(rpca_md, aes(x=PC1, y=PC2, color = eliz_bin)) + geom_point(aes(fill = eliz_bin),pch=21, size=2)+
  ggtitle("RPCA - Robust Aitchison")+
  scale_color_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  scale_fill_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  theme_cowplot()+
    xlab(paste0("PC1 (", signif((rpca_ah_axes$PC1) * 100, 4), "%)")) + 
    ylab(paste0("PC2 (", signif((rpca_ah_axes$PC2) * 100, 4), "%)")) +
 # labs_pubr()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16))+
  stat_ellipse( type = "norm")


#wow - this is a pretty strong effect! 

```


# Phylo-RPCA by Eliz 

Still clustering, but proceed with RPCA because the Eliz-based clustering is stronger there 


```{r}

#read in phylo-RPCA ordination
phylo_rpca_ah <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/phylo_rpca_ordination.qza")
rpca_ah_axes <- phylo_rpca_ah$data$ProportionExplained
phylo_rpca_ah <- phylo_rpca_ah$data$Vectors %>% dplyr::select(SampleID,PC1,PC2)
phylo_rpca_ah <- phylo_rpca_ah %>% mutate(sample_id = SampleID) #add new column to match metadata



#classify each sample as to whether it has > 50% elizabethkingia - might as well use that one specific ASV! 



sd <- sample_data(ps.filled)

# #halves
# sd$eliz_bin <- cut(
#   sd$perc_eliz,
#   breaks = c(0, 1/2, 1),
#   include.lowest = TRUE,
#   labels = c("<1/2", ">1/2")
# )
# # 
#thirds
sd$eliz_bin <- cut(
  sd$perc_eliz,
  breaks = c(0, 1/3, 2/3, 1),
  include.lowest = TRUE,
  labels = c("<1/3", "1/3–2/3", "2/3–1")
)

#quarters
# sd$eliz_bin <- cut(
#   sd$perc_eliz,
#   breaks = c(0, 1/4, 1/2, 3/4, 1),
#   include.lowest = TRUE,
#   labels = c("<1/4", "1/4–1/2", "1/2–3/4", "3/4-1")
# )
# 



sample_data(ps.filled) <- sd

met <- sample_data(ps.filled) %>% as.matrix() %>% data.frame() %>% filter(sample_type == "adult" | sample_type == "homogenate")

met <- met %>% 
  mutate(SampleID = sample_id) #fix sample_id vs SampleID differene before merging
#metadata2 <- subset(metadat, select = -c(sample_id) ) #not sure what this should do...? 


#merge metadata and ordination
phylo_rpca_md <- merge(phylo_rpca_ah, met, by = "SampleID")
head(phylo_rpca_ah)





ggplot(phylo_rpca_md, aes(x=PC1, y=PC2, color = eliz_bin)) + geom_point(aes(fill = eliz_bin),pch=21, size=2)+
  ggtitle("RPCA - Robust Aitchison")+
  scale_color_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  scale_fill_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  theme_cowplot()+
    xlab(paste0("PC1 (", signif((rpca_ah_axes$PC1) * 100, 4), "%)")) + 
    ylab(paste0("PC2 (", signif((rpca_ah_axes$PC2) * 100, 4), "%)")) +
 # labs_pubr()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16))+
  stat_ellipse( type = "norm")

```

##  Implementation of PAM clustering  

Let's get some biotypes to understand underlying differences in the adult samples agnostic of all our metadata factors. Then we can make a new metadata column (1,2,3,4, or however many groups) and feed those clusters into aldex2 to see which taxa are differentially abudnant between those two clusters. 



# Get a RPCA dist with just the adult samples (will need qiime for this) - export a trimmed metadata table, import the distance boject


```{r}

######### EXPORT 
ps.1042 <- prune_samples(sample_sums(ps.filled) > 1041, ps.filled ) %>% prune_taxa(taxa_sums(.) >0,.) #onlydo the samples that we also used for alpha div (DONT" WORRY, THIS IS NOT RAREFIED!!)
ps.adult <- subset_samples(ps.1042, sample_type == "adult") %>% prune_taxa(taxa_sums(.)>0,.) #4005, 462 samples

otu.df <- data.frame(otu_table(ps.adult)) #will need this for PAM clustering later

metadat <- sample_data(ps.adult) %>%
  data.frame() %>%  # Strip sample_data attributes completely
  rownames_to_column("sample-id") #rename for qiime 

write.table(metadat, "../data/qiime_gemelli/all_adults/adult_metadata.tsv", quote=FALSE, sep='\t', row.names = FALSE)
######### 

######go back to qiime, run gemelli on samples, then return to import


############# IMPORT: Get dist matrix ready 
rpcadist <- read_qza("../data/qiime_gemelli/all_adults/gemelli/rpca_distance.qza")
rpcadist <- rpcadist$data
#############


```

#PAM clustering 

Try multiple numbers of clusters - but remmber that fewer clusters is generally a better idea

```{r}


# setting function to do the pam clustering
pam.clustering=function(x,k) { # x is a distance matrix and k the number of clusters
  require(cluster)
  cluster = as.vector(pam(as.dist(x), k, diss=TRUE)$clustering)
  return(cluster)
}



dist_matrix <- rpcadist
# how many clusters needed
data.cluster=pam.clustering(dist_matrix, k=3) # start with any number of clusters
nclusters = index.G1(t(otu.df), data.cluster, d = dist_matrix, centrotypes = "medoids") 
#Error in dimnames(x) <- dn : length of 'dimnames' [1] not equal to array extent
nclusters #40.42
nclusters=NULL

for (k in 1:20) { 
  if (k==1) {
    nclusters[k]=NA 
  } else {
    data.cluster_temp=pam.clustering(dist_matrix, k)
    nclusters[k]=index.G1(t(otu.df),data.cluster_temp,  d = dist_matrix,
                          centrotypes = "medoids")
  }
}
plot(nclusters, type="h", xlab="k clusters", ylab="CH index")


#############TWO CLUSTERS!!!

# change number of clusters based on plot
data.cluster=pam.clustering(dist_matrix, k=2)
f=mean(silhouette(data.cluster, dist_matrix)[,3])
#0.3494798 weak

# plotting within clustering
obs.pcoa=dudi.pco(dist_matrix, scannf=F, nf=3) #scannf= whether to display eigenvalues bar plot, nf= axes to keep if scannf is false
s.class(obs.pcoa$li, fac=as.factor(data.cluster), grid=F) 
s.class(obs.pcoa$li, fac=as.factor(data.cluster), grid=F, cell=0, cstar=0, col=c(3,2,4,7)) 

# retrieving assigned cluster
cluster.assignments_2 <- cbind(data.cluster,obs.pcoa$li)


############# THREE CLUSTERS!!!

# change number of clusters based on plot
data.cluster=pam.clustering(dist_matrix, k=3)
obs.silhouette=mean(silhouette(data.cluster, dist_matrix)[,3])
#0.3789591 slighlty better, but not compelling

# plotting within clustering
obs.pcoa=dudi.pco(dist_matrix, scannf=F, nf=3) #scannf= whether to display eigenvalues bar plot, nf= axes to keep if scannf is false
s.class(obs.pcoa$li, fac=as.factor(data.cluster), grid=F) 
s.class(obs.pcoa$li, fac=as.factor(data.cluster), grid=F, cell=0, cstar=0, col=c(3,2,4,7)) 

# retrieving assigned cluster
cluster.assignments_3 <- cbind(data.cluster,obs.pcoa$li)



########## FOUR CLUSTERS!!!

# change number of clusters based on plot
data.cluster=pam.clustering(dist_matrix, k=4)
obs.silhouette=mean(silhouette(data.cluster, dist_matrix)[,3])
#0.2651989 - worst

# plotting within clustering
obs.pcoa=dudi.pco(dist_matrix, scannf=F, nf=3) #scannf= whether to display eigenvalues bar plot, nf= axes to keep if scannf is false
s.class(obs.pcoa$li, fac=as.factor(data.cluster), grid=F) 
s.class(obs.pcoa$li, fac=as.factor(data.cluster), grid=F, cell=0, cstar=0, col=c(3,2,4,7)) 

# retrieving assigned cluster
cluster.assignments_3 <- cbind(data.cluster,obs.pcoa$li)






```


#Plot ordination with both biotypes and >50% elizabethkingia highlighted

Biotype2 is overwhelmingly >50% elizabethkingia!

AF12B

```{r}



#merge metadata and ordination
phylo_rpca_md <- merge(phylo_rpca_ah, met, by = "SampleID")
head(phylo_rpca_ah)





sd <- sample_data(ps.filled)

#halves
sd$eliz_bin <- cut(
  sd$perc_eliz,
  breaks = c(0, 1/2, 1),
  include.lowest = TRUE,
  labels = c("<1/2", ">1/2")
)
#
# #thirds
# sd$eliz_bin <- cut(
#   sd$perc_eliz,
#   breaks = c(0, 1/3, 2/3, 1),
#   include.lowest = TRUE,
#   labels = c("<1/3", "1/3–2/3", "2/3–1")
# )

#quarters
# sd$eliz_bin <- cut(
#   sd$perc_eliz,
#   breaks = c(0, 1/4, 1/2, 3/4, 1),
#   include.lowest = TRUE,
#   labels = c("<1/4", "1/4–1/2", "1/2–3/4", "3/4-1")
# )
# 



sample_data(ps.filled) <- sd

met <- sample_data(ps.filled) %>% as.matrix() %>% data.frame() %>% filter(sample_type == "adult" | sample_type == "homogenate")

met <- met %>% 
  mutate(SampleID = sample_id) #fix sample_id vs SampleID differene before merging
#metadata2 <- subset(metadat, select = -c(sample_id) ) #not sure what this should do...? 


#merge metadata and ordination
phylo_rpca_md <- merge(phylo_rpca_ah, met, by = "SampleID")
ca2 <- cluster.assignments_2 %>% rownames_to_column("SampleID")

phylo_rpca_md <- merge(phylo_rpca_md, ca2)




ggplot(phylo_rpca_md, aes(x=PC1, y=PC2, color = as.character(data.cluster))) + geom_point(aes(fill = as.character(data.cluster), shape = eliz_bin), size=2)+
  ggtitle("RPCA - Robust Aitchison")+
  scale_color_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  scale_fill_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  theme_cowplot()+
    xlab(paste0("PC1 (", signif((rpca_ah_axes$PC1) * 100, 4), "%)")) + 
    ylab(paste0("PC2 (", signif((rpca_ah_axes$PC2) * 100, 4), "%)")) +
 # labs_pubr()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16))+
  stat_ellipse( type = "norm")


#ALright! So it's not entirley 1:1, but very close agreement between biotype and how much Eliz they have. Biotype2 is really dominated by samples with >50% of reads as elizabethkingia. So now we can run aldex2 next. 
```

# Look at relabund for biotype! 

AF12A


```{r}


#define a function to plot various phyloseq objects

ps.1042 <- prune_samples(sample_sums(ps.filled) > 1041, ps.filled ) %>% prune_taxa(taxa_sums(.) >0,.) #onlydo the samples that we also used for alpha div (DONT" WORRY, THIS IS NOT RAREFIED!!)
ps.adult <- subset_samples(ps.1042, sample_type == "adult") %>% prune_taxa(taxa_sums(.)>0,.) #4005, 462 samples

make_microshades_plot <- function(ps,
                                  facet_var,
                                  nrow = 2,
                                  title = "Microshades Plot") {
  # prepare melted data
  mdf_prep <- ps %>%
    phyloseq::transform_sample_counts(function(x) { x/sum(x) }) %>%
    psmelt() %>%
    filter(Abundance > 0)
  
  # create color object
  color_obj <- microshades::create_color_dfs(
    mdf_prep,
    group_level = "Phylum",
    subgroup_level = "Genus",
    selected_groups = c("Proteobacteria", 
                        "Actinobacteriota",
                        "Bacteroidota",
                        "Firmicutes_D")
  )
  
  # extract dfs
  mdf <- color_obj$mdf
  cdf <- color_obj$cdf
  
 cdf <- color_reassign(cdf,
                           group_assignment = c("Proteobacteria", "Actinobacteriota","Bacteroidota","Firmicutes_D"),
                           color_assignment = c("micro_green", "micro_orange","micro_blue","micro_brown"))


  new_groups <- extend_group(mdf, cdf, "Phylum", "Genus", "Proteobacteria", existing_palette = "micro_green", new_palette = "micro_purple", n_add = 5)
  #add 5 more slots to proteo
  
  
  # base plot and legend
  base_plot <- plot_microshades(new_groups$mdf, new_groups$cdf)
  legend <- custom_legend(new_groups$mdf, new_groups$cdf)
  
  # build final plot
  plot <- base_plot +
    scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
    theme(legend.position = "none",
          axis.text.x = element_text(size = 6),
          plot.margin = margin(6,6,6,6),
    # REMOVE GREY BACKGROUND!!
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank()
    ) +
    facet_wrap(as.formula(paste("~", facet_var)),
               scales = "free_x",
               nrow = nrow) +
    ggtitle(title)
  
  # combine plot + legend
  plot_grid(plot, legend, rel_widths = c(1, .1))
}



#Need to add data.cluster to the psobj itself
ca2 <- mutate(ca2, "sample_id" = SampleID)


#now add data.cluster to the 

# Pull sample_data out as a data frame
sd <- sample_data(ps.adult) %>%
  data.frame() %>%
  rownames_to_column("sample_name")
sd2 <- sd %>%
  left_join(
    ca2,
    by = "sample_id"
  )
sample_data(ps.adult) <- sample_data(
  sd2 %>% column_to_rownames("sample_name")
)



plot <- make_microshades_plot(
  ps = ps.adult, 
  facet_var = "~data.cluster",
  nrow = 1,
  title = "Adults from biotypes"
)
plot


#ggsave("./plots/biotypes/biotype_relabund.pdf", height = 20, width=80, device="pdf", limitsize = FALSE)



```


# Metadata characteristics of Biotype1 vs Biotype2

```{r}

phylo_rpca_md

biotype1 <- dplyr::filter(phylo_rpca_md, data.cluster == "1")
biotype2 <- dplyr::filter(phylo_rpca_md, data.cluster == "2")


table(biotype1$eliz_bin) #all less tahn 50% eliz

table(biotype2$eliz_bin) #55 <1/2, 70 > 1/2 

#boxplot of perc_eliz

class(phylo_rpca_md$perc_eliz)


plot <- phylo_rpca_md %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=data.cluster, y=as.numeric(perc_eliz), fill=as.factor(data.cluster)) +
  #scale_fill_manual(values = c("#D0352B","#3C76B0", "#BBDE94"))+
  #oord_cartesian(ylim=c(0,5)) + # adjust y-axis
  xlab("Biotype") +
  ylab("Relative Abundnace of Elizabethkingia")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") #+  #remove the legend as it isn't needed
 # ggtitle("UW homogenate comparisons")

plot





```



# Aldex2 on biotype

We obviously expect Elizabethkingia to be different between the two - but other taxa might be driving this difference? 


```{r}


#

sample_data(ps.adult)$data.cluster <-as.factor(sample_data(ps.adult)$data.cluster) #make uni_tx a factor




#this runs quickly
aldex.clr.biotype<-aldex.clr((as.data.frame(otu_table(ps.adult))),
                             sample_data(ps.adult)$data.cluster,
                             mc.samples=128,
                             denom="all",
                             verbose=TRUE,
                             useMC=FALSE)
#recommedned that DMC (128) * number of samples in smallest group be at least 1000
#quick math - need to have at least 7.8 samples in the smallest group to do this test
table(sample_data(ps.adult)$data.cluster)
#259 : 1, 125 : 2




ttest<-aldex.ttest(aldex.clr.biotype, paired.test = FALSE, hist.plot = FALSE, verbose = FALSE)
effect <- aldex.effect(aldex.clr.biotype, CI=T, verbose=FALSE)

aldex.df <- data.frame(ttest,effect) #make new df from two metrics (ttest and effect size)

par(mfrow=c(1,2))
aldex.plot(aldex.df, type="MA", test="welch")
aldex.plot(aldex.df, type="MW", test="welch")


#if ttest and effect size both significant, put "both" in a new column
aldex.df$one_or_two<-"one"
aldex.df$one_or_two[which(aldex.df$we.eBH < 0.05 & aldex.df$wi.eBH < 0.05)] <-"both"


#get taxa info from original phyloseq object
taxa_info <- data.frame(tax_table(ps.adult))
taxa_info <- taxa_info %>% rownames_to_column(var = "ASV")


aldex.df<-aldex.df %>% rownames_to_column(var = "ASV")

aldex.df <- left_join(aldex.df, taxa_info) #add data info to aldexdf

sig_aldex2 <- aldex.df %>% filter(one_or_two=="both")#(not saved, just for looking - filter for just significant ASVs
#21 taxa


table(sig_aldex2$Class) #8 gammas and 1 bacteroidia
table(sig_aldex2$Order)
table(sig_aldex2$Family)
table(sig_aldex2$Genus) 
#use family 


aldex.df.edit<- aldex.df %>% mutate(pointlabel= if_else(one_or_two == "both",  Genus,"")) %>%
                               mutate(size= if_else(one_or_two == "both",  3,2)) %>%
                               mutate(color = if_else(one_or_two == "both" & effect > 0,  "red",
                                                      if_else(one_or_two == "both" & effect < 0,  "yellow","grey"))) %>%
                        mutate(famcolor = ifelse(one_or_two == "both", 
                        case_when(
                          Family == "Burkholderiaceae_A_580492" ~ "#B34657",
                          Family == "Burkholderiaceae_A_592522" ~ "#B34657",
                          Family == "Burkholderiaceae_A_595422" ~ "#B34657",
                          Family == "Moraxellaceae" ~ "#87C080",
                          Family == "Pseudomonadaceae" ~ "#45884E",
                          Family == "Weeksellaceae" ~ "#588FBF",
                          Family == "Xanthomonadaceae_616009" ~ "#E1E04B",
                          TRUE ~ "black"
                        ), 
                        "grey"))



volcano_biotype <-ggplot(aldex.df.edit, aes(x=diff.win,y=diff.btw)) +
  xlab("Log2 Dispersion")+ ylab("Log2 Difference")+
  geom_text_repel(aes(label = pointlabel), max.overlaps = 10000) +
geom_point(aes(size= size, color= famcolor))+
  scale_color_identity()+
  scale_size_identity()+
  theme_cowplot() +  theme(
  text=element_text(size=18),
  legend.title = element_text(size=18),
  legend.text = element_text(size=18),
  legend.margin = margin(6, 6, 6, 6),legend.box.background = element_rect(color="grey70",size=0.1),
  panel.grid.major = element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank()) +  guides(size = "none") + geom_abline(slope=1, intercept=0, linetype="dashed") + geom_abline(slope=-1, intercept=0, linetype="dashed")+
    scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits = c(-10,13))
volcano_biotype



sig_aldex2
#ok, probs the effect is the one that distinguishes it... but in what direction? 





## plot 

ps_rel <- transform_sample_counts(ps.adult, function(x) x/sum(x))
  melted <- psmelt(ps_rel)

  selected <- melted %>% filter(OTU %in% sig_aldex2$ASV)

  plot_list <- list()

  for (i in seq_len(nrow(sig_aldex2))) {
    asv <- sig_aldex2$ASV[i]
    genus <- sig_aldex2$Genus[i]

    taxon_data <- selected %>%
  filter(OTU == asv) %>%
  mutate(data.cluster = droplevels(factor(data.cluster)))


   p <- ggplot(
  taxon_data,
  aes(x = data.cluster, y = Abundance, color = data.cluster)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.2) +
  theme_pubr() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, vjust = 0.5)
  ) +
  labs(
    title = genus,
    subtitle = asv,
    y = "Relative Abundance",
    x = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

    plot_list[[i]] <- p
  }
  

   grid_plot <- plot_grid(plotlist = plot_list, ncol = 5)



sig_aldex2

# Deevosia = more in biotype 2, soif effect is +, it's more in biptype 2. 

biotype1_indicators <- filter(sig_aldex2, effect <0) #12 of them! 
biotype2_indicators <- filter(sig_aldex2, effect > 0) #9 of them! 



```


# Plotting 


```{r}

ps_rel <- transform_sample_counts(ps.adult, function(x) x/sum(x))
melted <- psmelt(ps_rel)

selected <- melted %>% filter(OTU %in% sig_aldex2$ASV)

# Initialize two plot lists
plot_biotype1 <- list()  # effect < 0
plot_biotype2 <- list()  # effect > 0

idx1 <- 1
idx2 <- 1

for (i in seq_len(nrow(sig_aldex2))) {

  asv    <- sig_aldex2$ASV[i]
  genus  <- sig_aldex2$Genus[i]
  effect <- sig_aldex2$effect[i]   # <-- adjust column name if needed

  taxon_data <- selected %>%
    filter(OTU == asv) %>%
    mutate(data.cluster = droplevels(factor(data.cluster)))

  p <- ggplot(
    taxon_data,
    aes(x = data.cluster, y = Abundance, color = data.cluster)
  ) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, size = 1.2) +
    theme_pubr() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, vjust = 0.5)
    ) +
    labs(
      title = genus,
      subtitle = asv,
      y = "Relative Abundance",
      x = NULL
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    )

  # Split plots by effect direction, add one to counter
  if (effect > 0) {
    plot_biotype2[[idx2]] <- p
    idx2 <- idx2 + 1
  } else if (effect < 0) {
    plot_biotype1[[idx1]] <- p
    idx1 <- idx1 + 1
  }
}

plot_biotype1

plot_grid(plotlist = plot_biotype1, ncol = 3)
plot_grid(plotlist = plot_biotype2, ncol = 3)


```

# What type of samples are in biotype1 vs biotype2? 


```{r}

met_big <- left_join(met, ca2) %>% filter(sample_type == "adult") %>% mutate(biotype = paste0("biotype",data.cluster)) %>% filter(biotype != "biotypeNA") %>% filter(wing_length != "NA")
#Need to remove the 5 adult samples with <1042 reads, do this by fitlering out to just those that got assigned a biotype 

met_big$wing_length <- as.numeric(met_big$wing_length)



# uni_tx counts
uni_tx_counts <- met_big %>%
  group_by(biotype, uni_tx) %>%
  summarise(n = n(), .groups = "drop") %>%
    pivot_wider(
    names_from = uni_tx,
    values_from = n,
    values_fill = 0
  )
#UW match is overwhelmingly in biotype 1, LSTM-match is split almost perfeclty in half
#UW mismatch is also split almost perfeclry in half
#LSTM mismatch also split almost in half, but with 43 in biotype1 and 33 inbiotype 2. So really LSTM-match is the only one different between the two! 

# timepoint counts

timepoint_biotype_counts <- met_big %>%
  count(timepoint, biotype) %>%        # count samples per timepoint × biotype
  pivot_wider(
    names_from = biotype,
    values_from = n,
    values_fill = 0)

timepoint_biotype_counts
#interesting - seems liek the unfrozen timepoints really lean towrads biotype1, 


wing_summary <- met_big %>%
  group_by(biotype, uni_tx) %>%
  summarise(
    mean_wing_length = mean(wing_length, na.rm = TRUE),
    sd_wing_length   = sd(wing_length, na.rm = TRUE),
    n = n(),
    .groups = "drop")

#oh, what if instead i made a boxplot of that 




############### uni
plot_uni <- met_big %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=interaction(biotype,uni), y=wing_length, fill=biotype) +
  #scale_fill_manual(values = c("#D0352B","#3C76B0", "#BBDE94"))+
  #oord_cartesian(ylim=c(0,5)) + # adjust y-axis
  xlab("Biotype") +
  ylab("Wing Length")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") #+  #remove the legend as it isn't needed
 # ggtitle("UW homogenate comparisons")

plot_uni
#ok, so it really doesn't seem like biotype1 or biotype2 dictates wing length - but how does match/mismatch interact with that? 


LSTM_wings <- filter(met_big, uni == "LSTM") 
ggqqplot(LSTM_wings$wing_length) #tbh not bad 
t.test(wing_length~biotype,LSTM_wings, var.equal=FALSE) #nope

UW_wings <- filter(met_big, uni == "UW") 
ggqqplot(UW_wings$wing_length) #tbh not bad 
t.test(wing_length~biotype,UW_wings, var.equal=FALSE) #nope

############### uni_tx
plot <- met_big %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=interaction(biotype,uni_tx), y=wing_length, fill=biotype) +
  #scale_fill_manual(values = c("#D0352B","#3C76B0", "#BBDE94"))+
  #oord_cartesian(ylim=c(0,5)) + # adjust y-axis
  xlab("Biotype") +
  ylab("Wing Length")  +
  theme_pubr() +
  facet_wrap(~uni, scales = "free_x")+
  theme(axis.text.x = element_text(angle=90, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") #+  #remove the legend as it isn't needed
 # ggtitle("UW homogenate comparisons")

plot

#intersting - some tiny trends could potentially bebigger biotype1 in LSTM-match and bigger UW-mismatch and LSTm-mismatch in biotype 2? Test each separately. 



UWMis_wings <- filter(met_big, uni_tx == "UW-mismatch") 
ggqqplot(UWMis_wings$wing_length) #tbh not bad 
t.test(wing_length~biotype,UWMis_wings, var.equal=FALSE) #0.051 NS


LSTMmis_wings <- filter(met_big, uni_tx == "LSTM-mismatch") 
ggqqplot(LSTMmis_wings$wing_length) #tbh not bad 
t.test(wing_length~biotype,LSTMmis_wings, var.equal=FALSE) #0.1033 NS

LSTMM_wings <- filter(met_big, uni_tx == "LSTM-match") 
ggqqplot(LSTMM_wings$wing_length) #tbh not bad 
t.test(wing_length~biotype,LSTMM_wings, var.equal=FALSE) #0.029 #techinicall significant - but do i need to adjust for doing 3 comparisons? point is, none of these are partiucarly strong comparisons 




```

