---
title: "cryo_diversity_and_ordinations"
author: "Holly Nichols"
date: "2024-01-17"
output: html_document
---


# Beta Diversity and Ordinations 


### Goals

Do samples in the same treatment (match/mismatch) or from the same university cluster together? 
How closely do the reicpient adults mirror the donor pool they receive? 
How distinct are 
Does beta diversity diverge once donor pool is stored in the freezer for increasing time? 
How distinct are the inital donor pools? 




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(RColorBrewer)
library(MetBrewer)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(ape)
library(vegan)
library(ggpubr) #for gghistogram
library(rstatix) #for wilcox_test
#library(philr) #for philR transformation adn plotting 
library(microbiome) #for CLR transformation 
#library(pairwiseAdonis) #for pairwise.adonis2
library(ranacapa) #for pairwise_adonis, which has built-in p.adjust
library(stats) #for p.adjust after pairwise.adonis2
library(cowplot) #to make the ordination plots pretty 
library(dunn.test) #for post-hoc test on alpha diversity 
library(multcompView) #for adding letters to plots after ANOVA 
#devtools::install_github("jbisanz/qiime2R", auth_token = "AUTH_GOES_HERE")
library(qiime2R) #to import/export from qiime
library(tibble) #for rownames_to_column
library(tidyverse)


ps.filled = readRDS("../data/phyloseq/crPSFilled")
ps.filled <- prune_taxa(taxa_sums(ps.filled) > 0, ps.filled)
#4493 taxa and 595 samples 12/12/2025

```




#Beta diversity 


## RPCA from Gemelli


First, export cleaned phyloseq object to qiime (ps.filled). Start with full object, can subset in qiime using a trimmed metadata object

```{r}

#only use samples with >1041 reads to match samples used in alpha div 
ps.1042 <- prune_samples(sample_sums(ps.filled) > 1041, ps.filled ) %>% prune_taxa(taxa_sums(.) >0,.)
temp <- as.data.frame(as.matrix((sample_data(ps.1042))))
sample_data(ps.1042) <- mutate(temp, graphvar = paste(sample_type,uni_tx,uni, sep = "_")) #create new column that will make it easy to graph adults and homogenates on same graph 
#4444 asvs, 583 samples
#12/12/2025


# Export taxonomy table as "tax.txt"


# Export taxonomy table as text file
tax<-as(tax_table(ps.1042),"matrix")
tax_cols <- colnames(tax)
tax<-as.data.frame(tax)

# Add prefixes to each column, qiime wants those 
prefixes <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__", "")
taxonomy_prefixed <- tax %>%
  mutate(across(
    everything(), 
    .fns = ~ paste0(prefixes[which(names(tax) == cur_column())], .)
  ))
taxonomy_prefixed$taxonomy<-do.call(paste, c(taxonomy_prefixed[tax_cols], sep=";"))#this will create a new column with all tax info separated by ";"
for(co in tax_cols) taxonomy_prefixed[co]<-NULL #remove all non-concat columns
write.table(taxonomy_prefixed, "qiime_gemelli/full_objects/full_tax.txt", quote=FALSE, col.names=FALSE, sep="\t")


# Export feature/OTU table as otu_biom.biom

library(biomformat);packageVersion("biomformat")

otu<-(as(otu_table(ps.1042),"matrix")) # add 't' to transform if taxa_are_rows=FALSE
otu_biom<-make_biom(data=otu)
write_biom(otu_biom,"qiime_gemelli/full_objects/full_table.biom")



# Export metadata to sample-metadata.txt


metadat <- sample_data(ps.1042) %>%
  data.frame() %>%  # Strip sample_data attributes completely
  rownames_to_column("sample-id") #rename for qiime 
write.table(metadat, "qiime_gemelli/r2q/fill_metadata.tsv", quote=FALSE, sep='\t', row.names = FALSE)

#Finally, export the tree as a Newick file (necessary for phylo-rpca). 
tree <- phy_tree(ps.1042)
write.tree(tree, file = "qiime_gemelli/full_objects/full_tree.nwk")








```


#Export adult + homogenate samples to qiime (gemelli)

```{r adulth}


ps.adult_h <- subset_samples(ps.1042, sample_type == "adult" | sample_type == "homogenate")
ps.adult_h <- prune_taxa(taxa_sums(ps.adult_h) > 0, ps.adult_h)

# Export feature/OTU table as otu_biom.biom

otu<-(as(otu_table(ps.adult_h),"matrix")) 
otu_biom<-make_biom(data=otu)
write_biom(otu_biom,"qiime_gemelli/adults_and_donors/adult_h_table.biom")

# Export taxonomy table as text file

tax<-as(tax_table(ps.adult_h),"matrix")
tax_cols <- colnames(tax)
tax<-as.data.frame(tax)

# Add prefixes to each column, qiime wants those 
prefixes <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__", "")
taxonomy_prefixed <- tax %>%
  mutate(across(
    everything(), 
    .fns = ~ paste0(prefixes[which(names(tax) == cur_column())], .)
  ))
taxonomy_prefixed$taxonomy<-do.call(paste, c(taxonomy_prefixed[tax_cols], sep=";"))#this will create a new column with all tax info separateed by ";"
for(co in tax_cols) taxonomy_prefixed[co]<-NULL #remove all non-concat columns
write.table(taxonomy_prefixed, "qiime_gemelli/adults_and_donors/adult_h_tax.txt", quote=FALSE, col.names=FALSE, sep="\t")



# Export metadata to sample-metadata.txt

metadat <- sample_data(ps.adult_h) %>%
  data.frame() %>%  # Strip sample_data attributes completely
  rownames_to_column("sample-id") #rename for qiime 

write.table(metadat, "qiime_gemelli/adults_and_donors/adult_h_sample-metadata.tsv", quote=FALSE, sep='\t', row.names = FALSE)

#Finally, export the tree as a Newick file (necessary for phylo-rpca). 
tree <- phy_tree(ps.adult_h)
write.tree(tree, file = "qiime_gemelli/adults_and_donors/tree.nwk")



```


# Next, move to qiime to make ordinations + distance matrices 

#Bring RPCA back go R for plotting (Fig 4A)


```{r plot-rpca}


#read in RPCA ordination
rpca_ah <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/rpca_ordination.qza")
rpca_ah_axes <- rpca_ah$data$ProportionExplained
rpca_ah <- rpca_ah$data$Vectors %>% dplyr::select(SampleID,PC1,PC2)
rpca_ah <- rpca_ah %>% mutate(sample_id = SampleID)

str(rpca_ah)

metadat <- metadat %>% 
  mutate(SampleID = sample_id) #fix sample_id vs SampleID differene before merging
#metadata2 <- subset(metadat, select = -c(sample_id) ) #not sure what this should do...? 


#merge metadata and ordination
rpca_md <- merge(metadat, rpca_ah, by = "SampleID")
head(rpca_md)




ggplot(rpca_md, aes(x=PC1, y=PC2, color = graphvar)) + geom_point(aes(fill = graphvar),pch=21, size=2)+
  ggtitle("RPCA - Robust Aitchison")+
  scale_color_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  scale_fill_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ #placeholders for now 
  theme_cowplot()+
    xlab(paste0("PC1 (", signif((rpca_ah_axes$PC1) * 100, 4), "%)")) + 
    ylab(paste0("PC2 (", signif((rpca_ah_axes$PC2) * 100, 4), "%)")) +
 # labs_pubr()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16))+
  stat_ellipse( type = "norm")


#ggsave("./plots/RPCA_adults_homogenates.pdf", height=5, width=9, device="pdf") 
```


https://forum.qiime2.org/t/to-remove-a-sample-from-alpha-beta-diversity-files-remove-from-matrix-or-from-metadata-and-count-table/18309/3
PCoA must be re-calculated if you remove samples because the pcoa is dependent on pairs of samples. 





#Bring phylo-rpca back for plotting (Fig 4B)

```{r plot-phylo-rpca}


phylo_rpca_ah <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/phylo_rpca_ordination.qza")
rpca_ah_axes <- phylo_rpca_ah$data$ProportionExplained
phylo_rpca_ah <- phylo_rpca_ah$data$Vectors %>% select(SampleID,PC1,PC2)
phylo_rpca_ah <- phylo_rpca_ah %>% mutate(sample_id = SampleID) #add new column to match metadata


head(rpca_ah)


#merge metadata and ordination
phylo_rpca_md <- merge(metadat, phylo_rpca_ah, by = "sample_id")
head(phylo_rpca_md)




ggplot(phylo_rpca_md, aes(x=PC1, y=PC2, color = graphvar)) + geom_point(aes(fill = graphvar),pch=21, size=2)+
  ggtitle("phylo-RPCA")+
  scale_color_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ 
  scale_fill_manual(values = c("#AECDE1","#3C76B0","#BBDE94","#549E3F","#ED9F9C","#D0352B"))+ 
  theme_cowplot()+
    xlab(paste0("PC1 (", signif((rpca_ah_axes$PC1) * 100, 4), "%)")) + 
    ylab(paste0("PC2 (", signif((rpca_ah_axes$PC2) * 100, 4), "%)")) +
 # labs_pubr()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text = element_text(size = 16))+
  stat_ellipse( type = "norm")

#ggsave("./plots/phylo_RPCA_adults_homogenates.pdf", height=5, width=9, device="pdf") 


```





# Stats on rpca and phylo-rpca: 

Are these groups significantly different (PERMANOVA)?
Which taxa affect the clustering we see? 

An example of a paper this did use rpca and do adonis2 
https://academic.oup.com/biomedgerontology/article/78/10/1925/6992157


PERMANOVA
```{r permanova-rpca}


rpcadist <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/rpca_distance.qza")
rpcadist <- rpcadist$data


#make a header for the first column/rownames of the dist matrix 

#To ensure that I'm correctly pairing the sample_type with the order of the dist object,
#Make a df that joins the dist and the unordered metadata object 
rpcadist_df <- rpcadist %>% as.matrix() %>% as.data.frame() 
rpcadist_df <- cbind(rownames(rpcadist_df), data.frame(rpcadist_df, row.names =NULL)) 

colnames(rpcadist_df)[1] <- "SampleID"
joined <- inner_join(rpcadist_df,metadat, by = "SampleID")

adonis2(rpcadist~joined$graphvar) # p = 0.001, significant! 
#                  Df SumOfSqs      R2      F Pr(>F)    
# joined$graphvar   5   406.91 0.38691 57.555  0.001 ***
# Residual        456   644.79 0.61309                  
# Total           461  1051.70 1.00000                      


joined$graphvar <- as.vector(joined$graphvar)
ps.disper <- betadisper(rpcadist,joined$graphvar) 
ps.disper
#    adult_LSTM-match_LSTM adult_LSTM-mismatch_LSTM        adult_UW-match_UW 
#                   1.3073                   1.4741                   0.9226 
#     adult_UW-mismatch_UW         homogenate__LSTM           homogenate__UW 
#                   0.9240                   0.7799                   0.4442 
# 
# Eigenvalues for PCoA axes:
# PCoA1 PCoA2 PCoA3 
# 472.3 467.1 112.3 

permutest(ps.disper) #significant, proceed with pairwise 
permutest(ps.disper,pairwise=TRUE)
summary(permustats(permutest(ps.disper, pairwise=TRUE)))
#listing comparisons that are NOT significant (aka have the same dispersion)
#adult uw match vs adult UW mismatch
#adult_UW-mismatch vs homogenate__LSTM


library(pairwiseAdonis)
pairwise.adonis(joined[ ,2:463], joined$graphvar, p.adjust.m = "bonferroni")



#not significant: 
#adult_LSTM-match_LSTM vs adult_LSTM-mismatch_LSTM


```

#phylo-rpca permanova 

```{r}

#try phylo-rpca, 
phylo_rpcadist <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/phylo_distance.qza")
phylo_rpcadist <- phylo_rpcadist$data
phylo_rpcadist_df <- phylo_rpcadist %>% as.matrix() %>% as.data.frame() 
phylo_rpcadist_df <- cbind(SampleID = rownames(phylo_rpcadist_df), data.frame(phylo_rpcadist_df, row.names =NULL)) 

colnames(phylo_rpcadist_df)[1] <- "SampleID"
joined2 <- inner_join(phylo_rpcadist_df,metadat, by = "SampleID")




adonis2(phylo_rpcadist~joined2$graphvar) # p = 0.001, significant! 
#                  Df SumOfSqs      R2      F Pr(>F)    
#joined2$graphvar   5   412.50 0.44279 72.471  0.001 ***


joined2$graphvar <- as.vector(joined2$graphvar)
ps.disper <- betadisper(phylo_rpcadist,joined2$graphvar) 
ps.disper # similar trends to the rpca, with tighter clustering of UW-match and UW mismatch. 

#    adult_LSTM-match_LSTM adult_LSTM-mismatch_LSTM        adult_UW-match_UW 
#                   1.2392                   1.2006                   0.7010 
#     adult_UW-mismatch_UW         homogenate__LSTM           homogenate__UW 
#                   0.6031                   0.9968                   0.4745 
# 
# Eigenvalues for PCoA axes:
#   PCoA1   PCoA2   PCoA3 
# 465.919 462.482   3.206 


permutest(ps.disper) #significant, proceed with pairwise 
permutest(ps.disper,pairwise=TRUE)
summary(permustats(permutest(ps.disper, pairwise=TRUE)))
#listing comparisons that are NOT significant (aka have the same dispersion)
#adult uw match vs adult UW mismatch
#adult_UW-mismatch_UW-homogenate__LSTM


pairwise.adonis(joined2[ ,2:463], joined$graphvar, p.adjust.m = "bonferroni")

```


# Next, move back to qiime and generate separate distance matrices for LSTM-match, LSTM-mismatch, etc... to analyze differences by *life stage* within each treatment 


# Create function for importing back from qiime and plotting ordinations

Go back to qiime to subset to each value of uni_tx and get rpca/phylo-rpca by sample_type
Note that this currenlty can handle 3 or fewer groups due to the way I've coded colors, would need to expand it for other projects

```{r}



met_all <- sample_data(ps.filled) %>% as.matrix() %>% as.data.frame()

# Define the function
create_rpca_plot <- function(ordination_path, metadata_filter, plot_title) {
  # Read ordination object
  rpca <- read_qza(ordination_path)

  # Extract proportion explained and vectors
  axes <- rpca$data$ProportionExplained
  rpca_vectors <- rpca$data$Vectors %>% select(SampleID, PC1, PC2)

  # Adjust for SampleID in metadata
  metadata_filtered <- met_all %>%
    mutate(SampleID = sample_id) %>%
    filter(uni_tx == metadata_filter)

  # Merge metadata and ordination vectors
  merged_data <- merge(metadata_filtered, rpca_vectors, by = "SampleID")

  # Create the ggplot object
  plot <- ggplot(merged_data, aes(x = PC1, y = PC2, color = sample_type)) +
    geom_point(aes(fill = sample_type), pch = 21, size = 2) +
    ggtitle(plot_title) +
    scale_color_manual(values = c("#AECDE1", "#BBDE94", "#3C76B0")) + 
    scale_fill_manual(values = c("#AECDE1", "#BBDE94", "#3C76B0")) + 
    theme_cowplot() +
    xlab(paste0("PC1 (", signif((axes$PC1) * 100, 4), "%)")) + 
    ylab(paste0("PC2 (", signif((axes$PC2) * 100, 4), "%)")) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.title = element_blank(),
      legend.text = element_text(size = 16),
      axis.text = element_text(size = 16),
      legend.position = "none"
    ) +
    stat_ellipse(type = "norm")

  # Return the plot
  return(plot)
}




```

# Plot by Sample Type, one panel for each uni_tx (RPCA)
Additional File 16

```{r}



plot_uwm<- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/uw_match/gemelli/rpca_ordination.qza",
  metadata_filter = "UW-match",
  plot_title = "RPCA - UW-match"
)

plot_uwMis <- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/uw_mismatch/gemelli/rpca_ordination.qza",
  metadata_filter = "UW-mismatch",
  plot_title = "RPCA - UW-mismatch"
)

plot_lstmM<- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/lstm_match/gemelli/rpca_ordination.qza",
  metadata_filter = "LSTM-match",
  plot_title = "RPCA - LSTM-match"
)


plot_lstmMis <- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/lstm_mismatch/gemelli/rpca_ordination.qza",
  metadata_filter = "LSTM-mismatch",
  plot_title = "RPCA - LSTM-mismatch"
)

ggarrange(plot_uwm, plot_lstmM, plot_uwMis, plot_lstmMis)
#ggsave("./plots/RPCA_uni_tx_sample_type.pdf", height=6, width=9, device="pdf") 

#double check that the colors are approrpiately labeled 
#adult = light blue, water = dark blue, larvae = green
plot_uwm + aes(label = sample_type) + geom_text()
plot_uwMis + aes(label = sample_type) + geom_text()
plot_lstmM + aes(label = sample_type) + geom_text()
plot_lstmMis +  aes(label = sample_type) + geom_text()


```

# Plot by Sample Type, one panel for each uni_tx (phylo-RPCA)
Additional File 17

```{r}


plot_phylo_uwm<- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/uw_match/gemelli/phylo_rpca_ordination.qza",
  metadata_filter = "UW-match",
  plot_title = "phylo-RPCA - UW-match"
)

plot_phylo_uwMis <- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/uw_mismatch/gemelli/phylo_rpca_ordination.qza",
  metadata_filter = "UW-mismatch",
  plot_title = "phylo-RPCA - UW-mismatch"
)

plot_phylo_lstmM<- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/lstm_match/gemelli/phylo_rpca_ordination.qza",
  metadata_filter = "LSTM-match",
  plot_title = "phylo-RPCA - LSTM-match"
)


plot_phylo_lstmMis <- create_rpca_plot(
  ordination_path = "../data/qiime_gemelli/trimmed_objects/lstm_mismatch/gemelli/phylo_rpca_ordination.qza",
  metadata_filter = "LSTM-mismatch",
  plot_title = "phylo-RPCA - LSTM-mismatch"
)

ggarrange(plot_phylo_uwm, plot_phylo_lstmM, plot_phylo_uwMis, plot_phylo_lstmMis)
#ggsave("./plots/phylo_RPCA_uni_tx_sample_type.pdf", height=6, width=9, device="pdf") 



#double check that the colors match the sampletype I expect - yep!
plot_phylo_uwm+ aes(label = sample_type) + geom_text()
plot_phylo_uwMis + aes(label = sample_type) + geom_text()
plot_phylo_lstmM+ aes(label = sample_type) + geom_text()
plot_phylo_lstmMis+ aes(label = sample_type) + geom_text()

```


#stats - sample type, RPCA 
Additional File 16 + 17 stats


```{r}




# Function to perform pairwise Adonis test
run_pairwise_adonis <- function(distance_path, metadata_filter, grouping_var) {
  # Read the distance object
  distance_obj <- read_qza(distance_path)
  distance_matrix <- distance_obj$data

  # Convert distance matrix to a data frame
  distance_df <- distance_matrix %>%
    as.matrix() %>%
    as.data.frame() %>%
    rownames_to_column(var = "SampleID")

  # Filter metadata
  metadata_filtered <- met_all %>%
    mutate(SampleID = sample_id) %>%
    filter(uni_tx == metadata_filter)

  # Join distance data with filtered metadata
  joined_data <- inner_join(distance_df, metadata_filtered, by = "SampleID")

  # Perform global Adonis test
  global_adonis <- adonis2(as.dist(distance_matrix) ~ joined_data[[grouping_var]])
  global_p_value <- global_adonis$`Pr(>F)`[1]

  # Check if global Adonis test is significant
  if (!is.na(global_p_value) && global_p_value >= 0.05) {
    return(paste("adonis2 found p =", signif(global_p_value, 4)))
  }

  # Perform pairwise Adonis test
  adonis_results <- pairwise.adonis(joined_data[ ,2:length(distance_df)],
                                    joined_data[[grouping_var]],
                                    p.adjust.m = "bonferroni"
                                    )
  # Return results
  return(list(global_adonis,adonis_results))
}



####################
#rpca PERMANOVA
####################



pwa_uwm <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/uw_match/gemelli/rpca_distance.qza",
  metadata_filter = "UW-match",
  grouping_var = "sample_type"
)
#water + larvae the same, both diff from adults 


pwa_uwmis <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/uw_mismatch/gemelli/rpca_distance.qza",
  metadata_filter = "UW-mismatch",
  grouping_var = "sample_type"
)
#all significantly different, water + larvae p = 0.006


pwa_lstmM <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/lstm_match/gemelli/rpca_distance.qza",
  metadata_filter = "LSTM-match",
  grouping_var = "sample_type"
)
#adult vs larvae is only significant difference, p = 0.003. 
#adults vs water is p = 0.072


pwa_lstmMis <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/lstm_mismatch/gemelli/rpca_distance.qza",
  metadata_filter = "LSTM-mismatch",
  grouping_var = "sample_type"
)
# adult and water dif, adult and larvae dif, water and larvae same. 



####################
#phylo-rpca PERMANOVA
####################

pwa_uwm_phylo <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/uw_match/gemelli/phylo_distance.qza",
  metadata_filter = "UW-match",
  grouping_var = "sample_type"
)

# adults different from water and larvae, water v larvae = p = 0.144

pwa_uwmis_phylo <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/uw_mismatch/gemelli/phylo_distance.qza",
  metadata_filter = "UW-mismatch",
  grouping_var = "sample_type"
)

# adults different from water and larvae, water v larvae = p = 0.291


pwa_lstmM_phylo <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/lstm_match/gemelli/phylo_distance.qza",
  metadata_filter = "LSTM-match",
  grouping_var = "sample_type"
)

#adults vs water is different, adult vs larvae = 0.249, larvae vs water p = 1.00

pwa_lstmMis_phylo <- run_pairwise_adonis(
  distance_path = "../data/qiime_gemelli/trimmed_objects/lstm_mismatch/gemelli/phylo_distance.qza",
  metadata_filter = "LSTM-mismatch",
  grouping_var = "sample_type"
)
#adult dif from water and larvae, larvae and water are the same. 





```



#Next - Boxplots! 



## Which adult samples are closest to the donor pool they recieved (boxplots)? 


Using the distance matrix with adults and donor pools from gemelli 


Plot generated here is a preview - not included in manuscript

```{r}
#----- PAIRWISE DISTANCE BOXPLOTS USING phylo-RPCA  DISTANCES -----#

set.seed(123456)


#first, melt the distance matrix so that it's long/tidy data with Var1 and Var2 being column names for the two libraries being compared, and 
#"value" being a column of aitchison distance between the two samples. 

prpca_Dist_melt <- subset(reshape2::melt(as.matrix(phylo_rpcadist), 
                                                value.name = "value"),
                                 				value != 0)



metadat_Var1 <- metadat %>% rename(Var1 = "sample-id")

#join the melted df with metadata
longdist <- prpca_Dist_melt %>% full_join(metadat_Var1, by = "Var1") %>%
           dplyr::select(Var1, Var2, value, sample_type, uni_tx, timepoint, stock_origin, uni, tx)

#rename metadata again, and join again by Var2 this time (need to have metadata for both sample being compared)
metadat_Var2 <- metadat %>% rename(Var2 = "sample-id")
longdist <- longdist %>% full_join(metadat_Var2, by = "Var2") %>% dplyr::select(Var1, Var2, value, sample_type.x, sample_type.y, uni_tx.x, uni_tx.y, timepoint.x, timepoint.y, stock_origin.x, stock_origin.y, uni.x, uni.y, tx.x, tx.y)
head(longdist)


#rename Var1 and Var2 to Var.x and Var.y
longdist <- longdist %>% rename(Var.x = Var1, Var.y = Var2) #and it's done!! 

#now subset to just the UW homogenate, UW-match, and LSTM-mismatch 
#Should be able to subset by stock_origin 

longdist_UWstock <- filter(longdist, stock_origin.x == "UW" & stock_origin.y == "UW")

# create new column to sort out just the desired pairs 
#THEN, filter for just the pairs we want (homogenate - homogneate or homogenate - adult pairs)

longdist_filtered <- longdist_UWstock %>%  mutate(UniqueSamtype = paste(sample_type.x, sample_type.y, sep = "_")) %>% 
        filter(UniqueSamtype  == "homogenate_homogenate" |
              UniqueSamtype  == "adult_homogenate" |
              UniqueSamtype  == "homogenate_adult")



#next, make a new column that defines the "lumped" groups we want to graph together - notably compare within UW homogenate, between homogenate and UW-match, and between homogenate and LSTM-mismatch. 


LumpConditions_UW <- function(df) {
  df <- df %>%
    mutate(
      graph_var = case_when(
        UniqueSamtype == "homogenate_homogenate" ~ "b/w UW Homogenate",
        uni_tx.x == "UW-match" | uni_tx.y == "UW-match" ~ "UW Homogenate to UW-match",
        TRUE ~ "UW Homogenate to LSTM-mismatch"
      )
    )
  return(df)
}



graphMe <- LumpConditions_UW(longdist_filtered)
DFNoDupes_UW <- graphMe[!duplicated(graphMe$value), ] #get rid of duplicate values - goes from 17080 to 8540 observations which is exactly half!




plot <- DFNoDupes_UW %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=graph_var, y=value, fill=graph_var) +
  scale_fill_manual(values = c("#D0352B","#3C76B0", "#BBDE94"))+
  coord_cartesian(ylim=c(0,5)) + # adjust y-axis
  xlab("Comparisons") +
  ylab("Beta-diversity Distance: phyloRPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("UW homogenate comparisons")

plot

#On average, UW-match adult samples are closer to donor pool than LSTM-mismatch samples are. 

```



# Repeat the above, but compare to LSTM homogenate (phylo-rpca)

Also not in manuscript 


```{r}
#now subset to just the UW homogenate, UW-match, and LSTM-mismatch 
#Subset by stock_origin 

longdist_LSTMstock <- filter(longdist, stock_origin.x == "LSTM" & stock_origin.y == "LSTM")

# create new column to sort out just the desired pairs 
#THEN, filter for just the pairs we want (homogenate - homogneate or homogenate - adult pairs)

longdist_filtered_LSTM <- longdist_LSTMstock %>%  mutate(UniqueSamtype = paste(sample_type.x, sample_type.y, sep = "_")) %>% 
        filter(UniqueSamtype  == "homogenate_homogenate" |
              UniqueSamtype  == "adult_homogenate" |
              UniqueSamtype  == "homogenate_adult")



#next, make a new column that defines the "lumped" groups we want to graph together - notably compare within UW homogenate, between homogenate and UW-match, and between homogenate and LSTM-mismatch. 

LumpConditions_LSTM <- function(df) {
  df <- df %>%
    mutate(
      graph_var = case_when(
        UniqueSamtype == "homogenate_homogenate" ~ "b/w LSTM Homogenate",
        uni_tx.x == "LSTM-match" | uni_tx.y == "LSTM-match" ~ "LSTM Homogenate to LSTM-match",
        TRUE ~ "LSTM Homogenate to UW-mismatch"
      )
    )
  return(df)
}



graphMe_L <- LumpConditions_LSTM(longdist_filtered_LSTM)
DFNoDupes_LSTM <- graphMe_L[!duplicated(graphMe_L$value), ] #get rid of duplicate values - goes from 17080 to 8540 observations which is exactly half!




plot2 <- DFNoDupes_LSTM %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=graph_var, y=value, fill=graph_var) +
  scale_fill_manual(values = c("#ED9F9C","#549E3F", "#AECDE1"))+
  coord_cartesian(ylim=c(0,5)) + # adjust y-axis
  xlab("Comparisons") +
  ylab("Beta-Diversity Distance: phylo-RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("LSTM homogenate comparisons")
#hm - looks like HOmogenate to LSTM-mismatch is lower than homogneate to UW-match? surprised by this 

plot2



ggarrange(plot,plot2)
#wow, these colors are terrible 

#recall the question:
#Are UW or LSTM mosquitoes farther from their original homogenate? 

#Seems like UW-match is closer to its homogenate 


```







# Phylo-rpca - Just the adults, no homogenate, distance from donor pool by uni_tx

This is the version I'll use in manuscript - Figure 4D

```{r}
DFNoDupes_UW$graph_var <- as.factor(DFNoDupes_UW$graph_var)
DFNoDupes_LSTM$graph_var <- as.factor(DFNoDupes_LSTM$graph_var)
all_dist_combos <- bind_rows(DFNoDupes_LSTM, DFNoDupes_UW)
#ok make a version with just the adults 

all_dist_combos_adults <- filter(all_dist_combos, UniqueSamtype != "homogenate_homogenate" ) #15,020 samples, match checks out


plot_adults <- all_dist_combos_adults %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=graph_var, y=value, fill=graph_var) +
  scale_fill_manual(values = c("#AECDE1","#549E3F", "#3C76B0", "#BBDE94"))+
  coord_cartesian(ylim=c(0,5.5)) + # adjust y-axis to match RPCA
  xlab("Comparisons") +
  ylab("Beta-diversity Distance: phylo-RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Adults compared to donor pool (phylo-RPCA)")

plot_adults

#ggsave("./plots/phylo_RPCA_distance_adults.pdf", height=7, width=4, device="pdf") 


underscores_for_stats_phylo_rpca <- all_dist_combos_adults
underscores_for_stats_phylo_rpca$graph_var <- gsub("-", "_", underscores_for_stats_phylo_rpca$graph_var)

anova_adults <- aov(lm(value ~ graph_var, data = underscores_for_stats_phylo_rpca))
summary(anova_adults) #p = < 2e-16 ***, proceed

tukey <- TukeyHSD(anova_adults, conf.level = 0.95) # 
multcompLetters4(anova_adults, tukey)

#try a model with uni, tx, and time point 
#maybe if it's homogenate_adult>> then use teh .y vale for pasting/mutating 
#and if it's adult_homogenate, use teh y value for pasting 


#LSTM Homogenate to UW_mismatch  LSTM Homogenate to LSTM_match UW Homogenate to LSTM_mismatch      UW Homogenate to UW_match 
#                           "a"                            "b"                            "c"                            "d" 




#generate new df to use for linear model 
head(all_dist_combos_adults)

df <- all_dist_combos_adults %>%
  #mutate(uni = paste0(uni.x, uni.y)) %>%
  mutate(uni_tx = paste0(uni_tx.x,uni_tx.y)) %>%
  mutate(timepoint = paste0(timepoint.x, timepoint.y)) %>%
  mutate(tx = paste0(tx.x, tx.y)) %>% 
  mutate(uni = case_when(
    UniqueSamtype == "adult_homogenate" ~ uni.x,
      TRUE ~ uni.y ))

head(df)




```




#phylo-rpca - distance from homogenate by TIMEPOINT (boxplots)

Additional File 4

```{r}

#must run chunks above to have correct df 


#note that this compresses uni_tx together to see any broad differences across time points in distance from homogenate
anova_time <- aov(lm(value ~ timepoint, data = df))
summary(anova_time) # p < 2e-16, proceed
tukey_time <- TukeyHSD(anova_time, conf.level = 0.95) 
#interesting - 24h same as 0h, all others are different 
multcompLetters4(anova_time, tukey_time)

plot_time <- df %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=timepoint, y=value, fill=timepoint) +
  #scale_fill_manual(values = c("#AECDE1","#549E3F", "#3C76B0", "#BBDE94"))+
  coord_cartesian(ylim=c(0,5.5)) + # adjust y-axis to match RPCA
  xlab("Comparisons") +
  ylab("Beta-diversity Distance: phylo-RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Adults compared to donor pool (phylo-RPCA)")

plot_time

#whoah! wtf, this is the first clear signal that there's an increase in distance from donor pool as time progresses. 
#note that this collapses variation. Let's look at just UW-match and LSTM-match 
#Wonder if this holds true for rpca and phylo-rpca, and what it means if not!!



head(df)

df_UWM <- filter(df, uni_tx == "UW-match")

#stats
anova_time_UWM <- aov(lm(value ~ timepoint, data = df_UWM))
summary(anova_time_UWM) # p < 2e-16, proceed
tukey_time_UWM <- TukeyHSD(anova_time_UWM, conf.level = 0.95)
#ok this one deserves letters 
multcompLetters4(anova_time_UWM, tukey_time_UWM)
#d10 24h mo6 d16  0h d30 
#"a" "a" "a" "b" "c" "d" 
#aligns with other analyses that there is NOT a consistent divergence from donor pool over time, but what about LSTM? 

#mo6-0h   0.27732971  0.20067296  0.35398646 0.0000000

#plot
plot_time_UW <- df_UWM %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=timepoint, y=value, fill=timepoint) +
  scale_fill_manual(values = c("#E5F6FF", "#B5DEFF", "#85C6FF", "#559EFF", "#2F74F5", "#153FBF"))+
  coord_cartesian(ylim=c(0,4.5)) + # adjust y-axis to match other panels 
  #xlab("Comparisons") +
  ylab("Beta-diversity Distance: phylo-RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("UW-match Adults compared to donor pool (phylo-RPCA)")





#LSTM 
df_LSTMM <- filter(df, uni_tx == "LSTM-match")

anova_time_LSTMM <- aov(lm(value ~ timepoint, data = df_LSTMM))
summary(anova_time_LSTMM) # p < 2e-16, proceed
tukey_time_LSTMM <- TukeyHSD(anova_time_LSTMM, conf.level = 0.95)
#ok this one deserves lettters 
multcompLetters4(anova_time_LSTMM, tukey_time_LSTMM)
#mo6 d30 d16 d10 24h  0h 
#"a" "b" "c" "d" "e" "e" 
#

#0h vs mo6 
            # diff      lwr           upr       p adj
#mo6-0h   0.71351845  0.59749688 0.82954003 0.0000000


#plot
plot_time_LSTM <- df_LSTMM %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=timepoint, y=value, fill=timepoint) +
  scale_fill_manual(values = c("#E5F6FF", "#B5DEFF", "#85C6FF", "#559EFF", "#2F74F5", "#153FBF"))+
  coord_cartesian(ylim=c(0,4.5)) + # adjust y-axis to match other panels
  #xlab("Comparisons") +
  ylab("Beta-diversity Distance: phylo-RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("LSTM-match Adults compared to donor pool (phylo-RPCA)")







ggarrange(plot_time_LSTM,plot_time_UW)
#ggsave("./plots/phylo-rpca-boxplots-timepoint.pdf", height=4, width=7, device="pdf") 


#conclusion -LSTM-match consistently moves farther for donor homogenate as time progresses, while UW-match does not have this consistent pattern. 
#also notable from the graph - LSTM have wider spread. 




```





#RPCA distance boxplots and stats (boxplots)

Figure 4C


```{r}

rpcadist <- read_qza("../data/qiime_gemelli/adults_and_donors/gemelli/rpca_distance.qza")
rpcadist <- rpcadist$data


rpca_Dist_melt <- subset(reshape2::melt(as.matrix(rpcadist), 
                                                value.name = "value"),
                                 				value != 0)


head(metadat)


metadat_Var1 <- metadat %>% rename(Var1 = "sample-id")

#join the melted df with metadata - will give metadata info on var1
longdist_rpca <- rpca_Dist_melt %>% full_join(metadat_Var1, by = "Var1") %>%
           select(Var1, Var2, value, sample_type, uni_tx, timepoint, stock_origin, uni, tx)

#rename metadata again, and join again by Var2 this time (need to have metadata for both sample being compared)
metadat_Var2 <- metadat %>% rename(Var2 = "sample-id")
longdist_rpca <- longdist_rpca %>% full_join(metadat_Var2, by = "Var2") %>% select(Var1, Var2, value, sample_type.x, sample_type.y, uni_tx.x, uni_tx.y, timepoint.x, timepoint.y, stock_origin.x, stock_origin.y, uni.x, uni.y, tx.x, tx.y)



#rename Var1 and Var2 to Var.x and Var.y
longdist_rpca <- longdist_rpca %>% rename(Var.x = Var1, Var.y = Var2) #and it's done!! 



# create new column to sort out just the desired pairs 
#THEN, filter for just the pairs we want (homogenate - homogneate or homogenate - adult pairs)



longdist_filtered_rpca <- longdist_rpca %>%  mutate(UniqueSamtype = paste(sample_type.x, sample_type.y, sep = "_")) %>% 
        filter(UniqueSamtype  == "homogenate_homogenate" |
              UniqueSamtype  == "adult_homogenate" |
              UniqueSamtype  == "homogenate_adult")


longdist_LSTM_rpca <- filter(longdist_filtered_rpca, stock_origin.x == "LSTM" & stock_origin.y == "LSTM")
longdist_UW_rpca <- filter(longdist_filtered_rpca, stock_origin.x == "UW" & stock_origin.y == "UW")




#next, make a new column that defines the "lumped" groups we want to graph together - notably compare within UW homogenate, between homogenate and UW-match, and between homogenate and LSTM-mismatch. 




graphMe_rpca_UW <- LumpConditions_UW(longdist_UW_rpca)
graphMe_rpca_LSTM <- LumpConditions_LSTM(longdist_LSTM_rpca)

DFNoDupes_UW_rpca <- graphMe_rpca_UW[!duplicated(graphMe_rpca_UW$value), ]
DFNoDupes_LSTM_rpca <- graphMe_rpca_LSTM[!duplicated(graphMe_rpca_LSTM$value), ]#get rid of duplicate values





all_dist_combos_rpca <- bind_rows(DFNoDupes_UW_rpca, DFNoDupes_LSTM_rpca)

plot <- all_dist_combos_rpca %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=graph_var, y=value, fill=graph_var) +
  #scale_fill_manual(values = c("#D0352B","#3C76B0", "#BBDE94", "#000000"))+
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  xlab("Comparisons") +
  ylab("Beta-diversity Distance: RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("all homogenate comparisons")
#hm - looks like HOmogenate to LSTM-mismatch is lower than homogenate to UW-match? surprised by this 

plot


#now remove the homogenate samples
all_dist_combos_adults_rpca <- filter(all_dist_combos_rpca, UniqueSamtype != "homogenate_homogenate" )

plot5 <- all_dist_combos_adults_rpca %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=graph_var, y=value, fill=graph_var) +
  scale_fill_manual(values = c("#AECDE1","#549E3F", "#3C76B0", "#BBDE94"))+
  coord_cartesian(ylim=c(0,5.5)) + # adjust y-axis
  xlab("Comparisons") +
  ylab("Beta-diversity Distance: RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Adults compared to homogenate (RPCA)")
#hm - looks like HOmogenate to LSTM-mismatch is lower than homogneate to UW-match? surprised by this 

plot5


#ggsave("./plots/RPCA_distance_adults.pdf", height=7, width=4, device="pdf") 


###########stats 


underscores_for_stats_rpca <- all_dist_combos_adults_rpca
underscores_for_stats_rpca$graph_var <- gsub("-", "_", underscores_for_stats_rpca$graph_var)

anova_adults2 <- aov(lm(value ~ graph_var, data = underscores_for_stats_rpca))
summary(anova_adults2) #p = < 2e-16 ***, proceed


tukey <- TukeyHSD(anova_adults2, conf.level = 0.95) # All essentially zero, LSTM-match adn UW-mismatch are the only ones that have a p adj not 0  (still 1e-07 though )
multcompLetters4(anova_adults2,tukey)
#UW Homogenate to LSTM_mismatch      UW Homogenate to UW_match  LSTM Homogenate to LSTM_match LSTM Homogenate to UW_mismatch 
#                           "a"                            "b"                            "c"                            "d" 



```



#RPCA - distance from homogenate - by TIMEPOINT (boxplots)

Additional File 4

```{r}

df_rpca <- all_dist_combos_adults_rpca %>%
  #mutate(uni = paste0(uni.x, uni.y)) %>%
  mutate(uni_tx = paste0(uni_tx.x,uni_tx.y)) %>%
  mutate(timepoint = paste0(timepoint.x, timepoint.y)) %>%
  mutate(tx = paste0(tx.x, tx.y)) %>% 
  mutate(uni = case_when(
    UniqueSamtype == "adult_homogenate" ~ uni.x,
      TRUE ~ uni.y ))





#must run chunk above to have correct df_rpca 


#note that this compresses uni_tx together to see any broad differences across time points in distance from homogenate
anova_time <- aov(lm(value ~ timepoint, data = df_rpca))
summary(anova_time) # p < 2e-16, proceed
tukey_time <- TukeyHSD(anova_time, conf.level = 0.95) 
multcompLetters4(anova_time, tukey_time)
#d30 mo6 24h d16  0h d10 
#"a" "b" "c" "d" "d" "d"
#interesting - unlike the phylo-rpca, no overall trend away from donor pool over time 
#d30 is the farthest from donor pool, then mo6, then 24h 



plot_time <- df_rpca %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=timepoint, y=value, fill=timepoint) +
  #scale_fill_manual(values = c("#AECDE1","#549E3F", "#3C76B0", "#BBDE94"))+
  coord_cartesian(ylim=c(0,5.5)) + # adjust y-axis to match RPCA
  xlab("Comparisons") +
  ylab("Beta-diversity Distance: RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Adults compared to donor pool (RPCA)")

plot_time

#whoah! wtf, this is the first clear signal that there's an increase in distance from donor pool as time progresses. 
#note that this collapses variation. Let's look at just UW-match and LSTM-match 
#Wonder if this holds true for rpca and RPCA, and what it means if not!!





df_rpca_UWM <- filter(df_rpca, uni_tx == "UW-match")

#stats
anova_time_UWM <- aov(lm(value ~ timepoint, data = df_rpca_UWM))
summary(anova_time_UWM) # p < 2e-16, proceed
tukey_time_UWM <- TukeyHSD(anova_time_UWM, conf.level = 0.95)
#ok this one deserves lettters 
multcompLetters4(anova_time_UWM, tukey_time_UWM)
#d30 mo6 24h d10  0h d16 
#"a" "b" "c" "d" "e" "f" 
#aligns with other analyses that there is NOT a consistent divergence from donor pool over time, but what about LSTM? 



#plot
plot_time_UW <- df_rpca_UWM %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=timepoint, y=value, fill=timepoint) +
  scale_fill_manual(values = c("#E5F6FF", "#B5DEFF", "#85C6FF", "#559EFF", "#2F74F5", "#153FBF"))+
  coord_cartesian(ylim=c(0,4.5)) + # adjust y-axis to match other panels 
  #xlab("Comparisons") +
  ylab("Beta-diversity Distance: RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("UW-match Adults compared to donor pool (RPCA)")





#LSTM 
df_rpca_LSTMM <- filter(df_rpca, uni_tx == "LSTM-match")

anova_time_LSTMM <- aov(lm(value ~ timepoint, data = df_rpca_LSTMM))
summary(anova_time_LSTMM) # p < 2e-16, proceed
tukey_time_LSTMM <- TukeyHSD(anova_time_LSTMM, conf.level = 0.95)
#ok this one deserves lettters 
multcompLetters4(anova_time_LSTMM, tukey_time_LSTMM)
# mo6  d30   0h  24h  d10  d16 
# "a"  "b"  "c"  "d" "de"  "e" 
#mo6 is def the farthest away 
#if antyhing, this one is U-shaped. Intersting to contrast with phylo-RPCA



#plot
plot_time_LSTM <- df_rpca_LSTMM %>%
  ggplot+
  geom_boxplot(outlier.shape = NA) +
  aes(x=timepoint, y=value, fill=timepoint) +
  scale_fill_manual(values = c("#E5F6FF", "#B5DEFF", "#85C6FF", "#559EFF", "#2F74F5", "#153FBF"))+
  coord_cartesian(ylim=c(0,4.5)) + # adjust y-axis to match other panels
  #xlab("Comparisons") +
  ylab("Beta-diversity Distance: RPCA")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("LSTM-match Adults compared to donor pool (RPCA)")







ggarrange(plot_time_LSTM,plot_time_UW)
#ggsave("./plots/rpca-boxplots-timepoint.pdf", height=4, width=7, device="pdf") 




```

# Which factors contribute to beta div of adult samples??


### Get  RPCA + phylo-RPCA dist with just the adult samples (will need qiime for this) - export a trimmed metadata table, import the distance object in code chunks below 



```{r}

######### EXPORT 
ps.1042 <- prune_samples(sample_sums(ps.filled) > 1041, ps.filled ) %>% prune_taxa(taxa_sums(.) >0,.) #onlydo the samples that we also used for alpha div (DONT" WORRY, THIS IS NOT RAREFIED!!)
ps.adult <- subset_samples(ps.1042, sample_type == "adult") %>% prune_taxa(taxa_sums(.)>0,.) #4005, 462 samples

otu.df <- data.frame(otu_table(ps.adult)) #will need this for PAM clustering later

metadat <- sample_data(ps.adult) %>%
  data.frame() %>%  # Strip sample_data attributes completely
  rownames_to_column("sample-id") #rename for qiime 

write.table(metadat, "../data/qiime_gemelli/all_adults/adult_metadata.tsv", quote=FALSE, sep='\t', row.names = FALSE)
######### 

```


# PERMANOVA with nesting - RPCA

Give this a try to compare the effect size of treatment (uni_tx) versus replicate (flaskID)

Additional File 5, sheet 3 

```{r}

rpcadist <- read_qza("../data/qiime_gemelli/all_adults/gemelli/rpca_distance.qza")
rpcadist <- rpcadist$data

#need a metadat obj
metadat_a <- sample_data(ps.filled) %>% as.matrix() %>% data.frame() %>% filter(sample_type == "adult")
metadat_a <- metadat_a %>% 
  mutate(SampleID = sample_id) #fix sample_id vs SampleID differene before merging



#To ensure that I'm correctly pairing the sample_type with the order of the dist object,
#Make a df that joins the dist and the unordered metadata object 
rpcadist_df <- rpcadist %>% as.matrix() %>% as.data.frame() 
rpcadist_df <- cbind(rownames(rpcadist_df), data.frame(rpcadist_df, row.names =NULL)) 

colnames(rpcadist_df)[1] <- "SampleID"
joined <- inner_join(rpcadist_df,metadat_a, by = "SampleID")


#### ok, now here's where we make the relevant columns 

colnames(joined)

# new columns - first one that encompasses all adults in a given combo of timepoitn + treatment
joined <- mutate(joined, timepoint_uni_tx = paste0(timepoint, "_", uni_tx))


#next, one that uniquely identifies each flask 
#Define a function to extract characters from the right side of a string 
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}


joined <- mutate(joined, flaskID = paste(timepoint, uni_tx, substrRight(joined$sorting_var, 1), sep = "_"))
joined$flaskID  <- sub("-", "_", joined$flaskID)

#make into vectors 
joined$uni_tx <- as.vector(joined$uni_tx)
joined$flaskID <- as.vector(joined$flaskID)

adonis2(rpcadist ~ joined$uni_tx* joined$timepoint/ joined$flaskID , permutations = 999) # the slash should mean nesting each flask within uni_tx

#for only uni_tx and flaskID (not used in manuscript, just to compare what it looks like without timepoint added)  
#                               DF    sumSq    R2    F      P
# joined$uni_tx                  3   104.57 0.11572 33.1282  0.001 ***
# joined$uni_tx:joined$flaskID  67   469.75 0.51984  6.6637  0.001 ***
# Residual                     313   329.32 0.36444                   
# Total                        383   903.65 1.00000                   
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#and if we add in timepoint (what actaully ran above / what we used in manuscript) - explains a significant amount, and so does the interaction! Some of what was getting attributed to flask was actaully a timepoint effect
# 
# adonis2(formula = rpcadist ~ joined$uni_tx * joined$timepoint/joined$flaskID, permutations = 999)
#                                                Df SumOfSqs      R2       F Pr(>F)    
# joined$uni_tx                                   3   104.57 0.11572 33.1282  0.001 ***
# joined$timepoint                                5   107.54 0.11900 20.4411  0.001 ***
# joined$uni_tx:joined$timepoint                  9   162.95 0.18032 17.2075  0.001 ***
# joined$uni_tx:joined$timepoint:joined$flaskID  53   199.27 0.22052  3.5735  0.001 ***
# Residual                                      313   329.32 0.36444                   
# Total                                         383   903.65 1.00000    


#how's the dispersion looking? 

anova(betadisper(rpcadist, joined$uni_tx)) # sig
anova(betadisper(rpcadist, joined$timepoint)) #Not sig! 
anova(betadisper(rpcadist, joined$flaskID)) #this is signiicant 



```


# PERMANOVA with nesting - phylo-RPCA

```{r}

phylo_rpcadist <- read_qza("../data/qiime_gemelli/all_adults/gemelli/phylo_distance.qza")
phylo_rpcadist <- phylo_rpcadist$data

#To ensure that I'm correctly pairing the sample_type with the order of the dist object,
#Make a df that joins the dist and the unordered metadata object 
phylo_rpcadist_df <- phylo_rpcadist %>% as.matrix() %>% as.data.frame() 
phylo_rpcadist_df <- cbind(rownames(phylo_rpcadist_df), data.frame(phylo_rpcadist_df, row.names =NULL)) 

colnames(phylo_rpcadist_df)[1] <- "SampleID"
joined_phylo <- inner_join(phylo_rpcadist_df,metadat_a, by = "SampleID")




# new columns - first one that encompasses all adults in a given combo of timepoitn + treatment
joined_phylo <- mutate(joined_phylo, timepoint_uni_tx = paste0(timepoint, "_", uni_tx))


#next, one that uniquely identifies each flask 
joined_phylo <- mutate(joined_phylo, flaskID = paste(timepoint, uni_tx, substrRight(joined_phylo$sorting_var, 1), sep = "_"))
joined_phylo$flaskID  <- sub("-", "_", joined_phylo$flaskID)



joined_phylo$flaskID <- as.vector(joined_phylo$flaskID)


adonis2(phylo_rpcadist ~ joined_phylo$uni_tx* joined_phylo$timepoint/ joined_phylo$flaskID , permutations = 999)

#                                                                 Df SumOfSqs      R2       F Pr(>F)    
# joined_phylo$uni_tx                                               3   198.77 0.25744 58.1706  0.001 ***
# joined_phylo$timepoint                                            5    24.84 0.03218  4.3625  0.001 ***
# joined_phylo$uni_tx:joined_phylo$timepoint                        9    54.64 0.07076  5.3299  0.001 ***
# joined_phylo$uni_tx:joined_phylo$timepoint:joined_phylo$flaskID  53   137.34 0.17788  2.2750  0.001 ***
# Residual                                                        313   356.51 0.46174                   
# Total                                                           383   772.09 1.00000                   
# ---
                                
#same trend as RPCA! now test disper


anova(betadisper(phylo_rpcadist, joined_phylo$uni_tx)) #sig
anova(betadisper(phylo_rpcadist, joined_phylo$timepoint)) #not sig
anova(betadisper(phylo_rpcadist, joined_phylo$flaskID)) #sig 

                                                              
#awesome, every trend is the same in rpca + phylo-rpca, simply report them together and ahve stats available in supplment 

```



