---
title: "cryo_alpha_diversity"
author: "Holly Nichols"
date: "2024-12-03"
output: html_document
---

# Cryopreservation Alpha Diversity 

**Goal**: How is alpha diversity affected by university (UW vs LSTM), treatment (match vs  mismatch), and duration of cryostorage (timepoint)? 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(RColorBrewer)
library(MetBrewer)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(ape)
library(vegan)
library(ggpubr) #for gghistogram
library(rstatix) #for wilcox_test
library(BiocManager)
#BiocManager::install("microbiome")
library(microbiome) #for CLR transformation 
library(devtools)
#Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS"=TRUE)
#install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis", auth_token = "AUTH_GOES_HERE")
library(pairwiseAdonis) #for pairwise.adonis2
#devtols::install_github("gauravsk/ranacapa", auth_token = "AUTH_GOES_HERE")
library(ranacapa) #for pairwise_adonis, which has built-in p.adjust
library(stats) #for p.adjust after pairwise.adonis2
library(cowplot) #to make the ordination plots pretty 
library(dunn.test) #for post-hoc test on alpha diversity 
install.packages("multcompView")
library(multcompView) #for adding letters to plots after ANOVA 




ps.filled = readRDS("../data/phyloseq/crPSdecontam_cutoff_12_11_2025")
#4493 taxa, 595 samples




```


# Summary stats

Overview of reads/ASVs in this sequencing run - and choose rarefaction depth 

```{r}

ps.filled # the phyloseq object with the final samples 
#4493 taxa and 595 samples 


#How many reads per sample, and reads per sample_type? 

samsum <- as.data.frame(sample_sums(ps.filled))
colnames(samsum)[1] = "reads" #change column name to something useful 
met <- as.data.frame(as.matrix(sample_data(ps.filled)))
met <- met %>% rownames_to_column() %>% arrange(rowname)
samsum <- samsum %>% rownames_to_column() %>% arrange(rowname)
sum(samsum$reads)
#18,936,817 total reads


j<- left_join(samsum,met, by= "rowname")

write.csv(j, "./choose_rarefaction.csv")
#Choose 1042 for rarefaction, reduces number of samples lost

reads_by_samtype <- j %>% group_by(sample_type) %>% summarise(avg_reads = mean(reads),
                                               median_reads = median(reads), 
                                               se_reads = sd(reads)/sqrt(n()) )


#                 avg     median    SE
# adult	      28949.72	26324.0	1194.755	
# homogenate	38847.29	39043.5	1375.716	
# larvae	    33876.95	22622.0	6527.530	
# water	      38169.14	38305.0	1811.838	

#mean/median of reads is consistent, but larvae hav high spread compared to others. Fewer reads off adult samples. 

all_summarised <- j %>% summarise(avg = mean(reads), median = median(reads))
#avg        median
#31826.58	  32099	



#What's the average number of ASVs in each sample type? 

asvs_per_sample <- sample_sums(otu_table(ps.filled) > 0)
summary(asvs_per_sample) #median of 22 ASVs overall 

ps.UW <- subset_samples(ps.filled,uni =="UW") %>% prune_taxa(taxa_sums(.) >0, .) #1796 taxa
ps.LSTM <- subset_samples(ps.filled,uni =="LSTM") %>% prune_taxa(taxa_sums(.) >0, .) #3008 taxa



#What are the dominant ASVs? 


#Extract the ASV abundance matrix and convert it to a data frame
asv_table <- as.data.frame(otu_table(ps.filled))  # Get OTU/ASV table
asv_totals <- rowSums(asv_table)          # Calculate total abundance per ASV

#Calculate total reads in the experiment
total_reads <- sum(asv_totals)

#Create a data frame with ASV totals and their contribution
asv_summary <- data.frame(
  ASV = rownames(asv_table),
  TotalAbundance = asv_totals,
  PercentContribution = round((asv_totals / total_reads) * 100, 2) # Rounded to 2 decimal places
)



#Rank ASVs by total abundance (descending)
asv_summary <- asv_summary %>%
  arrange(desc(TotalAbundance))

#17 total ASVs that are >1% of the dataset 

# Identify the  most dominant ASVs
top_asvs <- asv_summary[1:2, ]
top_22_asvs <- asv_summary[1:22, ]

#Elizabethkingia and Family Enterobacteriaceae make up 50% of reads 
#next 20 reads take it up to 91% of reads 


ps.h_UW <- subset_samples(ps.filled, sample_type == "homogenate" & uni == "UW") %>% prune_taxa(taxa_sums(.) > 0, .)
#99 taxa

ps.h_LSTM <- subset_samples(ps.filled, sample_type == "homogenate" & uni == "LSTM") %>% prune_taxa(taxa_sums(.) > 0, .)
#729

ps.all_LSTM <- subset_samples(ps.filled, uni == "LSTM") %>% prune_taxa(taxa_sums(.)>0,.)
#3028 taxa

ps.all_UW <- subset_samples(ps.filled, uni == "UW") %>% prune_taxa(taxa_sums(.)>0,.)
#1797 taxa


```

# Higher taxonomic trends 

What taxa dominate donor and recipient communities? Check out Class or Family level maybe? 

```{r}

#reload the objects from cryo_relabund_decontamfix.rmd 

ps.genus_filled = readRDS("../data/phyloseq/crPSGenus_filled")
ps.family_filled = readRDS("../data/phyloseq/crPSFamily")

#order
ps.order_filled = tax_glom(ps.filled, taxrank = 'Order') #glom to Order level
saveRDS(ps.order_filled, "../data/phyloseq/crPSOrder")
ps.order_filled = readRDS("../data/phyloseq/crPSOrder")
#271 taxa


#Class 
ps.class_filled = tax_glom(ps.filled, taxrank = 'Class') #glom to Order level
saveRDS(ps.class_filled, "../data/phyloseq/crPSClass")
ps.class_filled = readRDS("../data/phyloseq/crPSClass")
#101 taxa 


class_reads <- taxa_sums(ps.class_filled) %>% as.data.frame() %>% rownames_to_column()
colnames(class_reads) <- c("ASV", "reads")

taxa_info <- data.frame(tax_table(ps.filled)) %>% rownames_to_column("ASV")

class_reads %>% arrange(desc(reads)) %>%  # arrange in descending order for water
  left_join(taxa_info) %>% dplyr::select(ASV, reads, Domain, Phylum, Class)

total_reads <- sum(class_reads$reads)
top_5_classes <- filter(class_reads, reads > 100000) 
sum(top_5_classes$reads)/total_reads #that's 99.7% of reads! 


#Top 5 classes - Gammaproteobacteria, Bacteroidia, Alphaproteobacteria, Actinomycetia, Bacilli  - 




```


# Plot higher taxonomic levels 

(Additional File 2)

```{r}

#tax_table(ps.class_filled)

class_relabund <- transform_sample_counts(ps.class_filled, function(x) x / sum(x) )
melt_class <- psmelt(class_relabund)

  
melt_class_plot <- melt_class %>%
  mutate(Class_plot = case_when(
    Abundance < 0.1 ~ "< 10%",  #if that row is less than 10% of the reads in its sample, set ASV to  <10%
    TRUE ~ Class                                     # Otherwise, call it the same value from Class
  ))



stacked_bar_class <- melt_class_plot %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Class_plot)) +
   geom_bar(stat = "identity") +
   labs(x = "",y = "Relative Abundance",title = "Class Relative Abundance")+ 
   facet_wrap(~interaction(sample_type, uni), scales = "free_x") +
   theme_cowplot() + 
   #theme(legend.position="none") +
   theme(text = element_text(size=20), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8)) +
   scale_y_continuous(expand = c(0,0.001))+ # move x-axis up so it's nearly touching bars but not obscuring. could use 0,0 for most cases
   coord_cartesian(ylim=c(0,1))

stacked_bar_class

#ggsave("./plots/relabund_CLASS_uni_samtype.pdf", height=7, width=9, device="pdf") 


```






# Rarefaction 

Plot rarefaction curves. 
Commented out parts because it takes a while to run. 

```{r ggrare}

library(ggpubr)
#make a new variable that will distinguish UW donor pool from LSTM donor pool from all the 4 uni_tx combos,all in one variable 

sample_data(ps.filled)$type_uni <- paste(sample_data(ps.filled)$uni_tx,
                                         sample_data(ps.filled)$uni, sep = "_")


#highlight saturationor farefaction at our chosen cutoff
rareplot_type <- ggrare(ps.filled, step = 10, color = "type_uni", se = FALSE) + facet_wrap(~sample_type) + xlim(0,1042)
#add colors 
final_plot <- rareplot_type + scale_color_manual(values = c(
  "LSTM-match_LSTM" = "#a6cee3",
  "LSTM-mismatch_LSTM" = "#1f78b4",
  "UW-match_UW" = "#b2df8a",
  "UW-mismatch_UW" = "#33a02c",
  "_LSTM" = "#fb9a99",
  "_UW" = "#e31a1c"))+ 
  theme_pubr()+
 theme(legend.position="none")   #remove the legend as it isn't needed


#ggsave("./plots/Rarecurve_uni_tx.pdf", height=5, width=7, device="pdf")


###### Also include a longer version so that we can showthat we are saturating diversity with our sequencing - change xlim to the median read count (32099)

rareplot_type_long <- ggrare(ps.filled, step = 100, color = "type_uni", se = FALSE) + facet_wrap(~sample_type) + xlim(0,32099)
#make another rarefaction curve, this time with x axis extending to the median + of reads per sample 

final_plot <- rareplot_type_long + scale_color_manual(values = c(
  "LSTM-match_LSTM" = "#a6cee3",
  "LSTM-mismatch_LSTM" = "#1f78b4",
  "UW-match_UW" = "#b2df8a",
  "UW-mismatch_UW" = "#33a02c",
  "_LSTM" = "#fb9a99",
  "_UW" = "#e31a1c"))+ 
  theme_pubr()+
 theme(legend.position="none")   #remove the legend as it isn't needed

#ggsave("./plots/Rarecurve_uni_tx_long.pdf", height=5, width=7, device="pdf")





#look at rarfaction overall - 


ps.rarefied = rarefy_even_depth(ps.filled, rngseed=1, sample.size=1042, replace=F)
#2854 taxa and 583 samples 
ps.rarefied <- saveRDS(ps.rarefied, "../data/phyloseq/crPSRarefied")

ps.rarefied <- readRDS("../data/phyloseq/crPSRarefied")

```



# Are there major differences in sampling depth by any of our factors of interest? (life stage, uni_tx, timepoint)

If so, rarefaction might be unnecessarily biasing our conclusions. 

Conclusion - no large differences in sequencing depth. Proceed. 


```{r}

sample_sums(ps.filled)

#define a temporary phyloseq object to play with
ps.temp <- ps.filled

# create new column in metadata with read depth
sample_data(ps.temp)$read_depth <- sample_sums(ps.filled)

depth_df <- as.data.frame(as.matrix((sample_data(ps.temp))))
dim(depth_df)

#now plot read depth by various factors of interest

simpler <- depth_df %>% dplyr::select(sample_id, uni_tx, timepoint, sample_type, read_depth, pool_name)
simpler$read_depth <- as.numeric(simpler$read_depth)
simpler <-  mutate(simpler, log_depth = log10(read_depth))


plot <- simpler %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=sample_type, y=log_depth, fill=sample_type) +
  #scale_fill_manual(values = c("#A6CEE3", "#1F78B4"))+ 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  #xlab("Sample Type") +
  #ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  #facet_wrap(~uni, scales = "free_x", nrow = 1) +

  ggtitle("Read Depth by sample type")




plot1 <- simpler %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=uni_tx, y=log_depth, fill=uni_tx) +
  #scale_fill_manual(values = c("#A6CEE3", "#1F78B4"))+ 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  #xlab("Sample Type") +
  #ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  #facet_wrap(~uni, scales = "free_x", nrow = 1) +

  ggtitle("Read Depth by uni_tx")





plot2 <- simpler %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=pool_name, y=log_depth, fill=pool_name) +
  #scale_fill_manual(values = c("#A6CEE3", "#1F78B4"))+ 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  #xlab("Sample Type") +
  #ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  #facet_wrap(~uni, scales = "free_x", nrow = 1) +

  ggtitle("Read Depth by sequencing run ")




plot3 <- simpler %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=timepoint, y=log_depth, fill=timepoint) +
  #scale_fill_manual(values = c("#A6CEE3", "#1F78B4"))+ 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  #xlab("Sample Type") +
  #ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  geom_jitter(width = 0.2, size=1.5) +
  #facet_wrap(~uni, scales = "free_x", nrow = 1) +

  ggtitle("Read Depth by timepoint")



plot
plot1
plot2
plot3

#looks good - no major differnces in read depth. Some variaiton by sample type, as expected, but no huge bias

```





# Alpha diversity metrics 

Goals: plot observed, shannon, simpson. 
Coming in with the rarefied phyloseq object (1042)


## Diversity by sample type (all samples)

Note that this is the only time rarefied phyloseq object should be use - for alpha diversity calculations!

```{r shannon}
#### Alpha Diversity Metrics ####



Richness <- phyloseq::estimate_richness(ps.rarefied, split = TRUE, measures = c("Observed","Shannon", "InvSimpson", "Chao1")) 


Summary <- Richness %>%
  bind_cols(ps.rarefied %>% sample_data %>% as_tibble())
head(Summary) #Richness Estimates
dim(Summary)

#write.csv(Summary,"ShannonTable.csv")

#fix things that should be factors

cols <- c("sample_type", "pool_name", "uni", "uni_tx", "tx", "timepoint", "stock_origin")
Summary[cols] <- lapply(Summary[cols], factor)
#sapply(Summary,class) #check whether metadata values are desired class

#wing_length needs to be a numeric
Summary$wing_length <- as.numeric(Summary$wing_length)



#need to know if data are symmetrically distributed before doing any stats
gghistogram(Summary, x = "Shannon", y = "..density..", 
            fill = "steelblue",bins = 20, add_density = TRUE) + ggtitle("Distribution of Shannon Scores") 
#not very symmetrical, - strong leftwards tilt

gghistogram(Summary, x = "Observed", y = "..density..", 
            fill = "steelblue",bins = 20, add_density = TRUE) + ggtitle("Distribution of Observed Taxa")
#this one is much more symmetrical. Most have 10-40 taxa, with a few samples with >100 taxa

gghistogram(Summary, x = "Chao1", y = "..density..", 
            fill = "steelblue",bins = 20, add_density = TRUE) + ggtitle("Distribution of Chao1")
#Not terrible

gghistogram(Summary, x = "InvSimpson", y = "..density..", 
            fill = "steelblue",bins = 20, add_density = TRUE) + ggtitle("Distribution of InvSimpson")
#wildly biased towards the left here. 



#let's just plot the alpha div right here and now! 


plot <- Summary %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=sample_type, y=Shannon, fill=sample_type) +
  scale_fill_manual(values = met.brewer("Tam", 6)) + 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  xlab("Sample Type") +
  ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Alpha Diversity by Sample Type (rarefied = 1042)")

plot

#homogenate diversity lowest, then water + larvae, then adults 




###Summary of  test on richness (Shannon) - sample_type
kruskal.test(Shannon ~ sample_type, data = Summary) 
#Kruskal-Wallis chi-squared = 64.323, df = 3, p-value = 7.002e-14
dunn.test(Summary$Shannon,Summary$sample_type,method = "bonferroni")
pairwise.wilcox.test(Summary$Shannon, Summary$sample_type, p.adj = "bonf", exact = FALSE)
#Dunn: adult the same as water and larvae, but homogenate is lower than everyting else. 
#pairwise.wilcox: adult the same as water and larvae, but homogenate is lower than everyting else. 



Summary_UW <- filter(Summary, uni == "UW")
kruskal.test(Shannon ~ sample_type, data = Summary_UW) # p < 2.2e-16
dunn.test(Summary_UW$Shannon,Summary_UW$sample_type,method = "bonferroni")
#donor pool lower than everythign else

Summary_LSTM <- filter(Summary, uni == "LSTM")
kruskal.test(Shannon ~ sample_type, data = Summary_LSTM) 
#Kruskal-Wallis chi-squared = 2.5504, df = 3, p-value = 0.4662
#donor pool has higher div, so no differences here





```





# Alpha div of adults Timepoint - faceted by uni_tx - for manuscript (Additional File 3)

Final manuscript version, complete with ice-blue colors. CLD added in Inkscape to represent significant comparisons, see chunk below for details. 


```{r}
Summary_adults_only <- Summary %>% dplyr::filter(sample_type == "adult")


plot <- Summary_adults_only %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=timepoint, y=Shannon, fill=timepoint) +
  #scale_fill_manual(values = met.brewer("Tam", 6)) + 
  scale_fill_manual(values = c("#E5F6FF", "#B5DEFF", "#85C6FF", "#559EFF", "#2F74F5", "#153FBF"))+
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  xlab("Sample Type") +
  ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  theme(legend.position="none") +  #remove the legend as it isn't needed+
  facet_wrap(~uni_tx) +
  ggtitle("Alpha Diversity by Time Point (rarefied = 1042)")

plot

#ggsave("plots/alphadiv_timepoint_faceted_unitx.pdf",height=5, width=7, device="pdf")


#temp to count 
Summary_adults_only %>% filter(uni == "LSTM")



```

# Time alpha div stats for manuscript (AF3)

```{r}

#######stats on just LSTM-match #######

#Rswitch to swap to R version 4.4.0 to be able to instsall rcompanion for compact letter display (CLD)

#install.packages("rcompanion")
library(rcompanion)
#install.packages("FSA")
library(FSA)


###### LSTM match 
Summary_LSTM_match_adults  <- Summary_adults_only %>%filter(uni_tx == "LSTM-match")
kruskal.test(Shannon ~ timepoint, data = Summary_LSTM_match_adults ) 
#Kruskal-Wallis chi-squared = 33.575, df = 5, p-value = 2.893e-06
Phocdunn <- dunnTest(Shannon ~ timepoint, data=Summary_LSTM_match_adults, method="bonferroni")   
Phocdunns <- Phocdunn$res

cld_LSTM_match <- cldList(comparison = Phocdunns$Comparison,
        p.value    = Phocdunns$P.adj,
        threshold  = 0.05)[1:2] 
#Weird behavior of deleting any 0 characters in the Group name, but we can work with that!


#d16 and d30 are lower than 0h
#d16 lower than 24h
#d16 also diff from mo6




########stats on just UW_match #######

Summary_UW_match_adults  <- Summary_adults_only%>%filter(uni_tx == "UW-match")
kruskal.test(Shannon ~ timepoint, data = Summary_UW_match_adults ) 
#Kruskal-Wallis chi-squared = 19.273, df = 5, p-value = 0.001709
Phocdunn <- dunnTest(Shannon ~ timepoint, data=Summary_UW_match_adults, method="bonferroni")   
Phocdunns <- Phocdunn$res
cld_UW_match <- cldList(comparison = Phocdunns$Comparison,
        p.value    = Phocdunns$P.adj,
        threshold  = 0.05)[1:2] 

# 0h is same as all the rest!  
#24h is actaully higher than d10, that's the only diff


#stats on LSTM-mismatch 
Summary_LSTM_mismatch_adults  <- Summary_adults_only%>%filter(uni_tx == "LSTM-mismatch")
kruskal.test(Shannon ~ timepoint, data = Summary_LSTM_mismatch_adults ) 
#Kruskal-Wallis chi-squared = 27.856, df = 2, p-value = 8.935e-07
Phocdunn <- dunnTest(Shannon ~ timepoint, data=Summary_LSTM_mismatch_adults, method="bonferroni")   
Phocdunns <- Phocdunn$res
cld_UW_mismatch <- cldList(comparison = Phocdunns$Comparison,
        p.value    = Phocdunns$P.adj,
        threshold  = 0.05)[1:2] 
#d30 is lower than the others, d6 and mo6 are not different




#stats on UW-mismatch 
Summary_UW_mismatch_adults  <- Summary_adults_only%>%filter(uni_tx == "UW-mismatch")
kruskal.test(Shannon ~ timepoint, data = Summary_UW_mismatch_adults ) 
#Kruskal-Wallis chi-squared = 5.4648, df = 2, p-value = 0.06506
#not sig, do not follow with Dunn's

```









# Adult alpha diversity (rarefied), Figure 3b

For manuscript - adult diversity by uni-tx (no donor pool)


```{r}

summary_adults <- Summary %>% filter(sample_type == "adult")




plot <- summary_adults %>%
  ggplot+
  geom_boxplot() +#outlier.shape = NA) +
  aes(x=uni_tx, y=Shannon, fill=uni) +
  scale_fill_manual(values = c("#A6CEE3", "#1F78B4"))+ 
  #coord_cartesian(ylim=c(0,1)) + # adjust y-axis
  xlab("Sample Type") +
  ylab("Shannon")  +
  theme_pubr() +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  #geom_jitter(width = 0.2, size=1.5) +
  facet_wrap(~uni, scales = "free_x", nrow = 1) +
  theme(legend.position="none") +  #remove the legend as it isn't needed
  ggtitle("Adult Alpha Diversity")

plot

#ggsave("./plots/shannon_unitx.pdf", height=5, width=4, device="pdf") 





```




# Alpha div by uni_tx stats for manuscript 

```{r}

library(stringr)#cld doesn't like dashes in names , fix that first
Summary_adults_only$uni_tx <-
  str_replace_all(Summary_adults_only$uni_tx, "-", "_")

kruskal.test(Shannon ~ uni_tx, data = Summary_adults_only) 
#Kruskal-Wallis chi-squared = 19.264, df = 3, p-value = 0.0002411
Phocdunn <- dunnTest(Shannon ~ uni_tx, data=Summary_adults_only, method="bonferroni")   
Phocdunns <- Phocdunn$res


cld_unitx <- cldList(comparison = Phocdunns$Comparison,
        p.value    = Phocdunns$P.adj,
        threshold  = 0.05)[1:2] 

#a a b ab 
```

