---
title: "cryo_import_Dec_2025_final"
author: "Holly Nichols"
date: "2025-12-12"
output: html_document
---

#Cryo Import


* We are splitting the UW and LSTM samples and decontaming them with just their own extraction/experimental controls (deocontam1)
* Then we decontam them all against the 10 clean PCR controls, but then adding back common ASVs (found at >10% in at least 10 samples at each decontam step) because 10 controls is not enough to determine if extremely common taxa are contaminants against the 600 real samples!
* Final phyloseq object is called ps.decontam_final
* If you want a version that gets rid o samples with <100 reads, use ps.decontam_cutoff. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(qiime2R) # for importing from qiime
library(phyloseq)
library(decontam)
library(phytools)#to read in .nwk files
library(ggplot2)
library(dplyr)
library(cowplot)
library(tibble)

```





# Import from QIIME2

We need the following files from qiime: 

* metadata
* tree
* table (otu_table)
* taxonomy (tax_table)

Note that these will be located in here: 
/Users/HollyNichols1/Desktop/CoonLab/Cryopreservation/analysis/phyloseq

## Read in the metadata

Note that "lower than blank" values on quantas have been set to 0.0001 for the benefit of decontam
As of 1/17/23, the metadata file has been updated to fix the following: 

* CR258 was listed as UW-mismatch, when it is a 24h larvae C which is a UW-match sample. 
* CR341 and 342 were listed as samples, they are actually 6month food and water controls from LSTM. Also removed the LSTM-match label, and changed is.neg to TRUE
* CR565, 566, 567, and 568 were adult controls - removed the LSTM-match label. 
As of 1/24/24, changed CR686 from "UW" stock_origin to "LSTM"
On 1/30/24, I realized that a significant number of adult samples (44 total) were somehow classified as "ctrl" samples. This was fixed in a NEW version called cr-metadata_fix.csv, and was manually fixed in excel. 

* I didn't catch this earlier b/c when I was counting controls in UW and LSTM,i was counting by the "ctrl_or_sample"variable, which has the correct info. only the is.neg column was incorrect here. 
* For some reason it was only mo6 adult samples?? 

**METADATA UPDATE ** 03/01/24 I found metadata issues. Fixed in the cr-metadata_fix_TESTING_ground.xlsx file, saved as a .csv with the date as  "cr-metadata_fix_030124.csv"  then will import new file and propogate through 
Metadata issues: 

*  a few of the d30 water samples were mislabelled as d16. 
* CR264 was missing the LSTM stock_origin label 
* Please for the love of god let this be the last thing I discover wrong with the metadata 


Applies to SILVA and Greengenes2

```{r meta}

metadata_030124 <- read.csv("../data/phyloseq/cr-metadata_fix_030124.csv")
# set the first column with sample names to be row names
meta_rows_030124<-data.frame(metadata_030124, row.names = 1)
# use the phyloseq function to read in sample_data
meta_ps_030124<-phyloseq::sample_data(meta_rows_030124)
```


## Read in phylogenetic tree and re-root it 

Using code from Aldo: More pragmatic rooting upon import. 

```{r rootMe}

####More pragmatic tree rooting upon import####
pick_new_outgroup <- function(tree.unrooted){
  require("magrittr")
  require("data.table")
  require("ape") # ape::Ntip
  # tablify parts of tree that we need.
  treeDT <-
    cbind(
      data.table(tree.unrooted$edge),
      data.table(length = tree.unrooted$edge.length)
    )[1:Ntip(tree.unrooted)] %>%
    cbind(data.table(id = tree.unrooted$tip.label))
  # Take the longest terminal branch as outgroup
  new.outgroup <- treeDT[which.max(length)]$id
  return(new.outgroup)
}
# Read in unrooted tree
unrooted_qiime_default<-read.newick(file="../data/phyloseq/tree-dada2.nwk")
# the line below applies the user-defined function "pick_new_outgroup" to pick a more reasonable tree outgroup
new.outgroup = pick_new_outgroup(unrooted_qiime_default)
# the line below roots the tree using a package called "ape"
rootedTree = ape::root(unrooted_qiime_default, outgroup=new.outgroup, resolve.root=TRUE)
# the function below is the phyloseq function to make phylogenic tree
phy_tree<-phyloseq::read_tree(rootedTree, errorIfNULL=FALSE)


```


## Read in the OTU table


```{r otutab}
###OTU table
#first, converte .txt file from QIIME to a .csv. Make sure to remove the first row that just contains "# Constructed from biom file"
otu_csv<-read.csv("../data/phyloseq/otu_table-dada2.csv")
# set the first column with otu names to be the row names
otu_rows_fixed <- data.frame(otu_csv, row.names = 1)
# use the phyloseq function to read in the otu table making sure to tell phyloseq that each row (as opposed to each column) corresponds to a taxon
OTU_16S<-phyloseq::otu_table(otu_rows_fixed,taxa_are_rows = TRUE)

```



# Import the greengenes2 data

* Extracted reads w/ our primers, trained in-house on 2022.10 release of greengenes2
* used dada2 to denoise this.


```{r importgg }



##Taxonomy File
# IMPORTANT: Must first use Excel to separate the Taxon column into separate columns before importing here. 
# read in csv file
tax_csv_gg<-read.csv("../data/phyloseq/taxonomy_gg2_myprimers.csv")
# set the first column with otu names to be the row names
tax_rows_fixed_gg<-data.frame(tax_csv_gg, row.names = 1)
# the phyloseq function for reading in a taxonomy table expects a matrix:
tax_matrix_gg<-as.matrix(tax_rows_fixed_gg)
# use the phyloseq function to read in a tax_table
TAX_16S_gg<-phyloseq::tax_table(tax_matrix_gg, errorIfNULL=TRUE)


#now make a full phyloseq obj

ps.importgg2<-phyloseq(OTU_16S,TAX_16S_gg,meta_ps_030124,phy_tree)
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 5781 taxa and 696 samples ]
# sample_data() Sample Data:       [ 696 samples by 30 sample variables ]
# tax_table()   Taxonomy Table:    [ 5781 taxa by 8 taxonomic ranks ]
# phy_tree()    Phylogenetic Tree: [ 5781 tips and 5749 internal nodes ]

#saveRDS(ps.importgg2, "phyloseq/crPSimportgg2")

#to load this in the future: 

#ps.importgg2 = readRDS("phyloseq/crPSimportgg2")


length(which(sample_data(ps.importgg2)$is.neg=="TRUE"))
#Confirm there are 97 controls

```





# Remove non-bacterial reads


```{r bact}
##### Next, exclude non-bacterial reads
ps.bactgg2 <- subset_taxa(ps.importgg2, Domain=="Bacteria") #removed 39 taxa
#saveRDS(ps.bactgg2, "phyloseq/crPSbactgg2")


#5732 taxa and 696 samples

```



# Define functions 

to double check the work of decontam by creating a list of "contaminat" ASVs that are high abundance in true samples 
```{r}
RetainCommonASVs <- function(df, keep_list) {

  oldDF <- df

  tested_contaminant <- vector(mode = "character", length = nrow(oldDF))

    for (i in 1:nrow(oldDF)) {
    if (oldDF[i,"contaminant"] == "FALSE") {
      tested_contaminant[i] <- "FALSE"
      } else if (oldDF[i,"contaminant"] == "TRUE" & oldDF[i,"ASV"] %in% keep_list ) {
        tested_contaminant[i] <- "FALSE"
      } else {
        tested_contaminant[i] <- "TRUE"
      }
  }

  oldDF$tested_contaminant <- tested_contaminant

  return(oldDF)

}


ClassifyWhoIsRemoved <- function(df, keep_list) {

  oldDF <- df

  tested_contaminant <- vector(mode = "character", length = nrow(oldDF))

    for (i in 1:nrow(oldDF)) {
    if (oldDF[i,"contaminant"] == "FALSE") {
      tested_contaminant[i] <- "FALSE"
      } else if (oldDF[i,"contaminant"] == "TRUE" & oldDF[i,"ASV"] %in% keep_list ) {
        tested_contaminant[i] <- "SAVED"
      } else {
        tested_contaminant[i] <- "TRUE"
      }
  }

  oldDF$tested_contaminant <- tested_contaminant

  return(oldDF)

}

library(tibble)

ps.relabund <- transform_sample_counts(ps.bactgg2, function(x) { x/sum(x) })
melt_relabund <- psmelt(ps.relabund) #Commented out b/c it takes forever to run 

```


# Separate out UW samples, decontam using the UW experimental controls

```{r}


ps.dUW <- subset_samples(ps.bactgg2, uni == "UW") %>% #50 control samples
          subset_samples(., !(ctrl_or_sample == "PCR-ctrl")) %>% #remove PCR controls for this step (will need to add back in the final decontam step!!)
          prune_taxa(taxa_sums(.) > 0, .)   #remove taxa with 0 observations 
#2827 taxa and 333 samples


sample_data(ps.dUW)$is.neg <- as.logical(sample_data(ps.dUW)$is.neg) # ensure is.neg is a logical

contamdf.prev05_UW <- decontam::isContaminant(ps.dUW, method="prevalence", neg="is.neg", threshold=0.5)  
# make sure to add a new column called ASV to your contaminant dataframe
contamdf.prev05_UW_test <- mutate(contamdf.prev05_UW, ASV = rownames(contamdf.prev05_UW))

table(contamdf.prev05_UW$contaminant) #this kept 2467 taxa and tossed 367
#12/11 - FALSE 2496, TRUE 331


##############Retain common taxa 

ps.dUW_pruned <- prune_taxa(taxa_sums(ps.dUW) > 0, ps.dUW) %>%
  subset_samples(sample_type != "ctrl" ) #down to 283 samples
ps.dUW_pruned <- subset_samples(ps.dUW_pruned, sample_sums(ps.dUW_pruned) > 0  ) #ensure there are no samples with  0 reads (282 samples)


ps.relabundUW <- transform_sample_counts(ps.dUW_pruned, function(x) x / sum(x))# First create relative abundance phyloseq object
relabund_mat <- as(otu_table(ps.relabundUW), "matrix")# get OTU table

# keep taxa that are >X% in at least Y samples
keep_list_UW <- rownames(relabund_mat)[rowSums(relabund_mat > 0.10) >= 5] #25 to keep, but some are likley not contam to begin with 

#make the object
contamdf.prev05_UW_retained <- RetainCommonASVs(contamdf.prev05_UW_test, keep_list_UW)
contamdf.prev05_UW_retained$tested_contaminant <- as.logical(contamdf.prev05_UW_retained$tested_contaminant) #must be logical for filters downstream
#note that the "contaminant" column will still show all as true, but the tested_contaminant column is the new column that includes info after keep_common step
ps.decontam_UW_prev_1 <- prune_taxa(!contamdf.prev05_UW_retained$tested_contaminant, ps.dUW)  #make the phyloseq object!
#2500 taxa and 333 samples 


tax_sums_UW<- taxa_sums(ps.dUW) %>% as.matrix() %>% as.data.frame() %>%rownames_to_column()
colnames(tax_sums_UW) <- c("ASV","reads")

UW_removal <- ClassifyWhoIsRemoved(contamdf.prev05_UW_test, keep_list_UW) %>% left_join(tax_sums_UW, by = "ASV")
table(UW_removal$tested_contaminant)

# FALSE SAVED  TRUE 
#  2496     4   327 

saved_ASVS <- (UW_removal %>% filter(tested_contaminant == "SAVED"))$ASV




```


# Look at taxa kept from retain_common

```{r}

relabund_filtered <- filter(melt_relabund, OTU %in% saved_ASVS) 


bars <- relabund_filtered  %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(interaction(uni,sample_type)), scales = "free_x", ncol = 6) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

bars

removed_ASVS <- (UW_removal %>% filter(tested_contaminant == "TRUE"))$ASV
relabund_filtered <- filter(melt_relabund, OTU %in% removed_ASVS) 


bars_removed <- relabund_filtered  %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = "deeppink2")) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(interaction(uni,sample_type)), scales = "free_x", ncol = 6) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

bars_removed

#aegomonas, chryseobacterium, Escherichia, staph - the latter two are suspicous, investigate further 

```
# Make function to graph every ASV - use to investigate those called contaminatns, and those kept 


```{r}

plot_OTU_bars <- function(physeq_object, 
                          melt_relabund_df,
                          otu_list,
                          output_dir = NULL) {


  # Extract taxonomy table as df
  taxdf <- phyloseq::tax_table(physeq_object) %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column("OTU")

  # Prepare output list
  plot_list <- list()

  # Loop over OTUs
  for (otu in otu_list) {

    # Filter abundance table
    relabund_filtered <- dplyr::filter(melt_relabund_df, OTU == otu)

    # Get genus & title
    genus <- taxdf$Genus[taxdf$OTU == otu]
    genus <- ifelse(is.na(genus) | genus == "", "Unclassified", genus)

    title_text <- paste0(genus, " (", substr(otu, 1, 5), ")")

    # Create bar plot
    p <- relabund_filtered %>%
      ggplot(aes(x = sample_id, y = Abundance, fill = "deeppink3")) +
      geom_bar(stat = "identity") +
      labs(title = title_text,
           x = "",
           y = "Relative Abundance") +
      facet_wrap(~(interaction(uni, sample_type)),
                 scales = "free_x",
                 ncol = 6) +
      theme_cowplot() +
      theme(
        text = element_text(size = 20),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 8),
        legend.position = "none"
      ) +
      scale_y_continuous(expand = c(0, 0.001))

    # Save if requested
    if (!is.null(output_dir)) {
      if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
      outfile <- file.path(output_dir, paste0(otu, "_barplot.pdf"))
      ggsave(outfile, p, width = 10, height = 6)
    }

    # Add to list
    plot_list[[otu]] <- p
  }

  return(plot_list)
}



```

#
# UW - plot contaminants   
```{r}

#will be 327 plots 

contams_UW_plots <- plot_OTU_bars(
  physeq_object = ps.relabund,
  melt_relabund_df = melt_relabund,
  otu_list = removed_ASVS #,
  #output_dir = "otu_barplots"   # optional
)

#contams_UW_plots

#looks good - none that are biased towrads oen sample type in particular 


```


# UW "saved" contaminant

```{r}

#will be 6 plots 

saved_UW_plots <- plot_OTU_bars(
  physeq_object = ps.relabund,
  melt_relabund_df = melt_relabund,
  otu_list = saved_ASVS #,
  #output_dir = "otu_barplots"   # optional
)

#saved_UW_plots


#these two look suspicious, neither Staph or Eschericia are known to be present in moz microbiome, and both ar at high abundance in controls  and not present in donor pools 
#d46e2205f0c6ecf67b51f83d111c509c
#922d33a2963d3e87a3d4df6b82371235

# Remove Eschericia + staph - too common in controls and not known to be mosquito-associated 
#remove suspicous ASVs that were "saved" in UW

manual_contams <- c("922d33a2963d3e87a3d4df6b82371235", "d46e2205f0c6ecf67b51f83d111c509c")

ps.decontam_UW_prev <- prune_taxa(!(taxa_names(ps.decontam_UW_prev_1) %in% manual_contams), ps.decontam_UW_prev_1)  #make the phyloseq object!
#2498 taxa and 333 samples



```





# Separate out LSTM samples, decontam using the LSTM experimental controls


```{r}


ps.dLSTM <- subset_samples(ps.bactgg2, uni == "LSTM") %>% #50 control samples
          subset_samples(., !(ctrl_or_sample == "PCR-ctrl")) %>% #remove PCR controls for this step (will need to add back in the final decontam step!!)
          prune_taxa(taxa_sums(.) > 0, .)   #remove taxa with 0 observations 
#3543 taxa and 344 samples


sample_data(ps.dLSTM)$is.neg <- as.logical(sample_data(ps.dLSTM)$is.neg) # ensure is.neg is a logical

contamdf.prev05_LSTM <- decontam::isContaminant(ps.dLSTM, method="prevalence", neg="is.neg", threshold=0.5)  
# make sure to add a new column called ASV to your contaminant dataframe
contamdf.prev05_LSTM_test <- mutate(contamdf.prev05_LSTM, ASV = rownames(contamdf.prev05_LSTM))

table(contamdf.prev05_LSTM$contaminant) #this kept 2467 taxa and tossed 367
#12/11 - FALSE 3291, TRUE 252


##############Retain common taxa 

#make a new phyloseq object of just the OTUs that are present in at least X% of REAL samples
#need to relabund it first

ps.dLSTM_pruned <- prune_taxa(taxa_sums(ps.dLSTM) > 0, ps.dLSTM) %>%
  subset_samples(sample_type != "ctrl" ) #down to 316 samples
ps.dLSTM_pruned <- subset_samples(ps.dLSTM_pruned, sample_sums(ps.dLSTM_pruned) > 0  ) #ensure there are no samples with  0 reads (282 samples)


ps.relabundLSTM <- transform_sample_counts(ps.dLSTM_pruned, function(x) x / sum(x))# First create relative abundance phyloseq object
relabund_mat <- as(otu_table(ps.relabundLSTM), "matrix")# get OTU table

# keep taxa that are >X% in at least Y samples
keep_list_LSTM <- rownames(relabund_mat)[rowSums(relabund_mat > 0.10) >= 5] #18 taxa

#make the object to filter by
contamdf.prev05_LSTM_retained <- RetainCommonASVs(contamdf.prev05_LSTM_test, keep_list_LSTM)
contamdf.prev05_LSTM_retained$tested_contaminant <- as.logical(contamdf.prev05_LSTM_retained$tested_contaminant) #must be logical for filters downstream
#note that the "contaminant" column will still show all as true, but the tested_contaminant column is the new column that includes info after keep_common step


ps.decontam_LSTM_prev <- prune_taxa(!contamdf.prev05_LSTM_retained$tested_contaminant, ps.dLSTM) 
#3297 taxa and 344 samples


tax_sums_LSTM<- taxa_sums(ps.dLSTM) %>% as.matrix() %>% as.data.frame() %>%rownames_to_column()
colnames(tax_sums_LSTM) <- c("ASV","reads")

LSTM_removal <- ClassifyWhoIsRemoved(contamdf.prev05_LSTM_test, keep_list_LSTM) %>% left_join(tax_sums_LSTM, by = "ASV")
table(LSTM_removal$tested_contaminant)

#10% in 5 samples
# FALSE SAVED  TRUE 
#  3291     6   246 


saved_ASVS_LSTM <- (LSTM_removal %>% filter(tested_contaminant == "SAVED"))$ASV

```

#plot LSTM saved


```{r}
relabund_filtered <- filter(melt_relabund, OTU %in% saved_ASVS_LSTM) 


bars <- relabund_filtered  %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(interaction(uni,sample_type)), scales = "free_x", ncol = 6) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

bars

#keeps unclassifed Enterobacteriaceae (red bars), vibrio (which is clearly real in LSTM donor pool, pedobacter (clealry present in one flask of LSTM larvae) sphingobacteirum (also seems to be water/larval assocateed))


removed_ASVS_LSTM <- (LSTM_removal %>% filter(tested_contaminant == "TRUE"))$ASV
relabund_filtered <- filter(melt_relabund, OTU %in% removed_ASVS_LSTM) 


bars_removed <- relabund_filtered  %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = "deeppink1")) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(interaction(uni,sample_type)), scales = "free_x", ncol = 6) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

bars_removed


```

# LSTM - plot contaminants
```{r}

#will be 267 plots 

contams_LSTM_plots <- plot_OTU_bars(
  physeq_object = ps.relabund,
  melt_relabund_df = melt_relabund,
  otu_list = removed_ASVS_LSTM #,
  #output_dir = "otu_barplots"   # optional
)

#contams_LSTM_plots


#Add achromobater to keep list - it's a low abundacen but is associated consistently with water/larval samples
# keep taxa that are >X% in at least Y samples
keep_list_LSTM <- rownames(relabund_mat)[rowSums(relabund_mat > 0.10) >= 5] #18 taxa
keep_list_LSTM <- c(keep_list_LSTM, "20b3df08cd5fac2be26a2928155bab69")

#make the object to filter by
contamdf.prev05_LSTM_retained <- RetainCommonASVs(contamdf.prev05_LSTM_test, keep_list_LSTM)
contamdf.prev05_LSTM_retained$tested_contaminant <- as.logical(contamdf.prev05_LSTM_retained$tested_contaminant) #must be logical for filters downstream
#note that the "contaminant" column will still show all as true, but the tested_contaminant column is the new column that includes info after keep_common step


ps.decontam_LSTM_prev <- prune_taxa(!contamdf.prev05_LSTM_retained$tested_contaminant, ps.dLSTM) 
#3298 taxa and 344 samples




```




# LSTM "saved" contaminants

```{r}

saved_LSTM_plots <- plot_OTU_bars(
  physeq_object = ps.relabund,
  melt_relabund_df = melt_relabund,
  otu_list = saved_ASVS_LSTM #,
  #output_dir = "otu_barplots"   # optional
)

#saved_LSTM_plots

```




# Cleanup for decontam - phase 2

First need to combine the two decontam'd datasets - nervous about this part
Then need to add the PCR ctrl samples back in 
Then need to remove the shitty PCR ctrl samples 
Then can start decontam again! 


```{r recombine_4_decontam}

#First need a phyloseq object with just the ctrl samples 

ps.PCRctrl <- subset_samples(ps.bactgg2, ctrl_or_sample == "PCR-ctrl") # 19 PCR ctrls, need to remove the bad guys 

#remove the PCR ctrl samples that had > 20,000 reads
ctrl20000 <- c( "CR106","CR147","CR239","CR274","CR317","CR448","CR506","CR552","CR613")# PCR ctrl samples with > 20,000 reads
ps.PCRctrl <- subset_samples(ps.PCRctrl, !(sample_id %in% ctrl20000)) #down to 10 samples 



#deconstruct the PCR ctrl phyloseq object - rownames are still rownames in this batch. 

sd.PCR <- as.data.frame(as.matrix(sample_data(ps.PCRctrl)))
tax.PCR <- as.data.frame(as.matrix(tax_table(ps.PCRctrl))) #huh - this is sorted alphabetically. Is that normal? 
otu.PCR <- as.data.frame(as.matrix(otu_table(ps.PCRctrl)))

### deconstruct the decontam'd UW phyloseq object 
sd.UW <- as.data.frame(as.matrix(sample_data(ps.decontam_UW_prev)))
tax.UW <- as.data.frame(as.matrix(tax_table(ps.decontam_UW_prev)))
otu.UW <- as.data.frame(as.matrix(otu_table(ps.decontam_UW_prev)))

sd.LSTM <- as.data.frame(as.matrix(sample_data(ps.decontam_LSTM_prev)))
tax.LSTM <- as.data.frame(as.matrix((tax_table(ps.decontam_LSTM_prev))))
otu.LSTM <- as.data.frame(as.matrix(otu_table(ps.decontam_LSTM_prev)))


############ SAMPLE DATA ######### 

sd.PCR.rn <- ps.PCRctrl %>% sample_data() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("sample.id")
sd.UW.rn <- ps.decontam_UW_prev %>% sample_data() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("sample.id")
sd.LSTM.rn <- ps.decontam_LSTM_prev %>% sample_data() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("sample.id")


sd <- sd.PCR.rn %>% merge(sd.UW.rn, all = TRUE) %>% merge(sd.LSTM.rn, all = TRUE) #working! need to change sample.id back to rownames
sd_rows <- data.frame(sd, row.names = 1)
SD_D1 <- phyloseq::sample_data(sd_rows)
sample_data(SD_D1)



 ######## TAX TABLES ######
tax.PCR_rn <- ps.PCRctrl %>% tax_table() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("ASV")
tax.UW_rn <- ps.decontam_UW_prev %>% tax_table() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("ASV")
tax.LSTM_rn <- ps.decontam_LSTM_prev %>% tax_table() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("ASV")

tax <- tax.PCR_rn %>% merge(tax.UW_rn, all = TRUE) %>% merge(tax.LSTM_rn, all = TRUE) #working, yay!!! need to change ASV back to rownames 
tax_rows <- data.frame(tax, row.names = 1)
tax_matrix <- as.matrix(tax_rows) #phyloseq expects tax to be a matrix 
TAX_D1 <- phyloseq::tax_table(tax_matrix, errorIfNULL = TRUE )




############ OTU TABLES #########



otu.PCR_rn <- ps.PCRctrl %>% otu_table() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("ASV")
otu.UW_rn <- ps.decontam_UW_prev %>% otu_table() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("ASV")
otu.LSTM_rn <- ps.decontam_LSTM_prev %>% otu_table() %>% as.matrix() %>% as.data.frame %>% tibble::rownames_to_column("ASV")


otu<- otu.PCR_rn %>% merge(otu.UW_rn, all = TRUE) %>% merge(otu.LSTM_rn, all = TRUE) #5732 x 686
otu <- otu %>% replace(is.na(.), 0) #merge will populate with NAs instead of 0s. Fix that. 

otu_rows <- data.frame(otu, row.names = 1)
OTU_D1 <- phyloseq::otu_table(otu_rows, taxa_are_rows = TRUE)



######### INTO A PHYLOSEQ OBJECT ########

ps.decontam1 <-phyloseq(SD_D1,TAX_D1,OTU_D1,phy_tree) #note that the phy_tree is from the original import. 
#5732 taxa, 685 samples 
ps.decontam1 <- prune_taxa(taxa_sums(ps.decontam1) > 0, ps.decontam1) #remove taxa with 0 observations, down to 5457 taxa
ps.decontam1 <- prune_samples(sample_sums(ps.decontam1) > 0, ps.decontam1) #685 samples - two samples had 0 observations
#Bad PCR ctrls have already been removed. Time to continue to the next phase of decontam! 
#5441 taxa - post-SNAFU. was 5429
# 5446 taxa and 685 samples 3/11/24
#5450 taxa and 685 samples 7/15/24 (after the keep_common was reintroduced to the 1st decontam step)

#now - back to 5441, we've kept 1 ASV from UW and 4 from LSTM 


```


# Decontam round 2 - this time consider PCR controls and all samples 

```{r}


sample_data(ps.decontam1)$is.neg <- as.logical(sample_data(ps.decontam1)$is.neg) # ensure is.neg is a logical


#Remove all of the NON-PCR controls from this, becuase those were already accounted for in decontam1
ps.decontam1_rm <- subset_samples(ps.decontam1, sample_type != "ctrl" ) #because PCR-ctrl is a separate metadata variable, this removes all extraction and experimental controls
#608 samples
ps.decontam1_rm <- prune_taxa(taxa_sums(ps.decontam1_rm) > 0, ps.decontam1_rm) #remove taxa with 0 observations, down to 4599
#currenlty getting 4585 3/11/24
#now 4589 taxa and 608 samples 7/15/24
#12/11 4580




contamdf.prev05<- decontam::isContaminant(ps.decontam1_rm, method="prevalence", neg="is.neg", threshold=0.5)  
contamdf.prev05 <- mutate(contamdf.prev05, ASV = rownames(contamdf.prev05))

table(contamdf.prev05$contaminant)
# FALSE  TRUE 
#  4542    38 



##############Retain common taxa 

#make a new phyloseq object of just the OTUs that are present in at least X% of REAL samples
#need to relabund it first

ps.decontam1_pruned <- prune_taxa(taxa_sums(ps.decontam1_rm) > 0, ps.decontam1_rm) %>%
  subset_samples(sample_type != "ctrl" ) #down to 316 samples
ps.decontam1_pruned <- subset_samples(ps.decontam1_pruned, sample_sums(ps.decontam1_pruned) > 0  ) #ensure there are no samples with  0 reads (282 samples)


ps.relabundDecontam1<- transform_sample_counts(ps.decontam1_pruned, function(x) x / sum(x))# First create relative abundance phyloseq object
relabund_mat <- as(otu_table(ps.relabundDecontam1), "matrix")# get OTU table

# keep taxa that are >X% in at least Y samples
keep_list_Decontam1 <- rownames(relabund_mat)[rowSums(relabund_mat > 0.10) >= 10] %>% append("20b3df08cd5fac2be26a2928155bab69") #add back the one that it tried to remove from LSTM 

#make the object to filter by
contamdf.prev05_Decontam1_retained <- RetainCommonASVs(contamdf.prev05, keep_list_Decontam1)
contamdf.prev05_Decontam1_retained$tested_contaminant <- as.logical(contamdf.prev05_Decontam1_retained$tested_contaminant) #must be logical for filters downstream
#note that the "contaminant" column will still show all as true, but the tested_contaminant column is the new column that includes info after keep_common step


ps.decontam1_prev <- prune_taxa(!contamdf.prev05_Decontam1_retained$tested_contaminant, ps.decontam1_rm) 
#4543 taxa and 608 samples 


tax_sums_Decontam1<- taxa_sums(ps.decontam1_rm) %>% as.matrix() %>% as.data.frame() %>%rownames_to_column()
colnames(tax_sums_Decontam1) <- c("ASV","reads")

Decontam1_removal <- ClassifyWhoIsRemoved(contamdf.prev05, keep_list_Decontam1) %>% left_join(tax_sums_Decontam1, by = "ASV")
table(Decontam1_removal$tested_contaminant)

#10% in 10 samples 
#FALSE SAVED  TRUE 
# 4542     1    37 



saved_ASVS_Decontam1 <- (Decontam1_removal %>% filter(tested_contaminant == "SAVED"))$ASV


```

# Plot it


```{r}
relabund_filtered <- filter(melt_relabund, OTU %in% saved_ASVS_Decontam1) 


bars <- relabund_filtered  %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = Genus)) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(interaction(uni,sample_type)), scales = "free_x", ncol = 6) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

#bars

#keeps unclassifed Enterobacteriaceae (red bars), vibrio (which is clearly real in LSTM donor pool, pedobacter (clealry present in one flask of LSTM larvae) sphingobacteirum (also seems to be water/larval assocateed))


removed_ASVS_decontam1 <- (Decontam1_removal %>% filter(tested_contaminant == "TRUE"))$ASV
relabund_filtered <- filter(melt_relabund, OTU %in% removed_ASVS_decontam1) 


bars_removed <- relabund_filtered  %>%
   ggplot(aes(x =sample_id, y = Abundance, fill = "deeppink1")) +
   geom_bar(stat = "identity") +
  labs(x = "",y = "Relative Abundance")+ 
   facet_wrap(~(interaction(uni,sample_type)), scales = "free_x", ncol = 6) +
  theme_cowplot() + 
  theme(text = element_text(size=20), axis.text.x=element_blank(),axis.ticks.x=element_blank()) +
   theme(strip.text.x = element_text(size = 8))+ 
  scale_y_continuous(expand = c(0,0.001))# move x-axis up so it's nearly touching bars but not obscuring 

bars_removed

#only staph at 10% in 10 samples , 10% in 5 samples had an aqatelia asv that was REALLy prevlaent in controls and not much elsewhere 

```


# Final cleaning

```{r}


ps.decontam_cutoff <- prune_samples(sample_sums(ps.decontam1_prev) >= 100, ps.decontam1_prev)
#604 samples
ps.decontam_cutoff <- subset_samples(ps.decontam_cutoff, sample_type != "PCR-ctrl")
#595 samples 
ps.decontam_cutoff <- prune_taxa(taxa_sums(ps.decontam_cutoff) > 0, ps.decontam_cutoff)   #remove taxa with 0 observations

saveRDS(ps.decontam_cutoff, "phyloseq/crPSdecontam_cutoff_12_11_2025")



#4493 taxa and 595 samples 
#final account of ASVs kept 
#tab_4493 <- taxa_sums(ps.decontam_cutoff) %>% as.matrix() %>% as.data.frame()
#write.csv(tab_4493, "investigate_4493.csv")

```









